The influence of nucleus accumbens shell D1 and D2 neurons on outcome-specific Pavlovian instrumental transfer

Version 1

Abstract

Abstract

The nucleus accumbens shell (NAc-S) and its projections to the ventral pallidum (VP) are thought to be critical for stimulus-based decisions. The NAc-S is predominantly composed of spiny projection neurons (SPNs) that express either the dopamine D1 (D1-SPNs) or the dopamine D2 receptor (D2-SPNs). Yet, the role of these two neuronal subpopulations and their inputs to the VP in stimulus-based decisions remains unknown. Here, we used optogenetics in female and male knock-in rats to selectively silence D1- or D2-SPNs and their projections to the VP at a time when the rats were required to use predictive stimuli to choose between two instrumental actions. Silencing either population of NAc-S SPNs disrupted choice. Silencing NAc-S D1-SPNs terminals in the VP also disrupted choice. However, choice was left intact by silencing NAc-S D2-SPNs terminals in the VP. Together, these findings provide novel insights into the cellular mechanisms and circuitry underlying stimulus-based decisions. We discuss how these insights are consistent with a recent model proposing that these decisions are controlled by an opioid-based memory system residing in the NAc-S.

Introduction

Our choices and decisions are often influenced by predictive signals available in the environment (Hollis, 1984). This influence is studied in the laboratory using the outcome-specific Pavlovian-instrumental transfer (PIT) task (Trapold & Overmier, 1972; Colwill & Rescorla, 1988; Holmes et al., 2010; Cartoni et al., 2016; Laurent & Balleine, 2021; Leung et al., 2024a), which comprises three stages. The first two stages are Pavlovian and instrumental conditioning, which can be administered in any order. In Pavlovian conditioning, two stimuli (S1 and S2; e.g., sounds or lights) are paired with two distinct and motivationally significant outcomes (O1 and O2; e.g., food outcomes). In instrumental conditioning, subjects are trained to perform two actions (A1 and A2), with each earning one of the two outcomes (i.e., A1◊O1 and A2◊O2). The final stage is the PIT test, which evaluates choice between the two actions in the presence (and absence) of the stimuli. Typically, each of the Pavlovian stimuli bias choice towards the action with which it shares an outcome: i.e., S1: A1>A2 and S2: A2>A1.

At a neural level, PIT has been found to rely on interactions between a number of structures, including cortical (Keistler et al., 2015; Bradfield et al., 2015; Bradfield et al., 2018; Tensaouti et al., 2025; Ostlund & Balleine, 2007; Lichtenberg et al., 2021; Sias et al., 2021; Lichtenberg et al., 2017), amygdala (Corbit & Balleine, 2005; Morse et al., 2020; Sias et al., 2024; Lichtenberg et al., 2017; Shiflett & Balleine, 2010; Derman et al., 2020; Prévost et al., 2012), basal ganglia (Corbit & Balleine, 2011; Corbit et al., 2001; Corbit & Janak, 2010; Morse et al., 2020; Laurent et al., 2012; Bertran-Gonzalez et al., 2013; Laurent et al., 2014; Corbit et al., 2016; Leung et al., 2024b; Leung & Balleine, 2015; Leung & Balleine, 2013; Laurent et al., 2015; Prévost et al., 2012), midbrain (Corbit et al., 2007; Sias et al., 2024; Seitz et al., 2022) and thalamic (Ostlund & Balleine, 2008; Leung et al., 2024b) territories. Among these, the nucleus accumbens shell (NAc-S) stands out because, unlike the others, it does not itself contribute to learning the stimulus-outcome (S-O) or the action-outcome (A-O) associations formed during the Pavlovian and instrumental stages of the PIT task and appears only to mediate their interaction (Corbit & Balleine, 2011; Corbit et al., 2001; Morse et al., 2020; Laurent et al., 2012; Laurent et al., 2014). Therefore, it has been proposed that the NAc-S acts as a central hub integrating S-O and A-O information at the time of PIT to guide choice between actions in an outcome-selective manner (Shiflett & Balleine, 2010; Bertran-Gonzalez & Laurent, 2018; Laurent & Balleine, 2021). Accordingly, PIT is disrupted by manipulations undermining NAc-S function during the PIT test (Morse et al., 2020; Corbit et al., 2016; Laurent et al., 2012; Laurent et al., 2014; Laurent et al., 2015).

The NAc-S is predominantly composed of spiny projection neurons (SPNs), which can be classified into two distinct subpopulations depending on the dopamine receptor they express (Gerfen & Surmeier, 2011). One population harbors the dopamine D1 receptors (D1Rs; D1-SPNs) whereas the other population express the dopamine D2 receptors (D2Rs; D2-SPNs). To date, only one study has examined whether activity in both populations is recruited during PIT (Laurent et al., 2014), finding that PIT is associated with an increase in ERK1/2 phosphorylation in D1-SPNs but not D2-SPNs. Further, blockade of D1Rs in NAc-S during choice between actions disrupted PIT whereas D2Rs blockade had no effect. Although these findings indicate a major involvement of NAc-S D1-SPNs in PIT, caution must be exerted when drawing conclusions on the role of D2-SPNs. While D1Rs are exclusively expressed on NAc-S D1-SPNs, D2Rs are present on cholinergic interneurons (CINs) and local presynaptic dopamine terminals in addition to D2-SPNs (Gerfen & Surmeier, 2011). Additionally, research shows that ERK1/2 phosphorylation may not capture all transcriptional events that occur in D2-SPNs (Bertran-Gonzalez et al., 2008; Matamales et al., 2020). Lastly, pharmacological D2R blockade has been found to be quite ineffective at influencing D2-SPNs function (Tozzi et al., 2007). Thus, the relative contribution of NAc-S D1- and D2-SPNs in driving choice between actions during PIT remains unclear.

The present experiments therefore aimed to provide an unambiguous assessment of the roles played by the two main populations of NAc-S SPNs during choice between actions in an outcome-specific PIT task. This assessment was achieved through optogenetic silencing in two knock-in rat lines that express Cre recombinase in either D1- or D2-SPNs (Pettibone et al., 2019). We first used tract-tracing and ex-vivo electrophysiology to confirm our capacity to selectively silence activity in NAc-S D1- or D2-SPNs. We then conducted two experiments that assessed outcome-specific task while D1- or D2-SPNs in the NAc-S were silenced during presentations of the predictive stimuli in the choice test. Finally, we assessed the degree to which the influence of the two SPNs populations on outcome-specific PIT depends on their projections to the ventral pallidum (VP). We focused on the VP for two main reasons: First, both NAc-S D1- and D2-SPNs send dense projections to this region (Lu et al., 1997; Kupchik et al., 2015); and, second, communication between the NAc-S and VP has previously been found to be critical for outcome-specific PIT (Leung & Balleine, 2013).

Results

Anterograde tracing and ex-vivo cell recordings in D1-Cre and A2a-Cre rats

We first examined the connectivity of striatal D1- and D2-SPNs in transgenic rats expressing Cre recombinase in neurons harboring the dopamine D1 receptor (D1- Cre rats) or the adenosine A2a receptor (A2a-Cre rats) (Pettibone et al., 2019). A2a receptors are predominantly expressed on striatal D2-SPNs (Schiffmann et al., 2007), thereby enabling selective targeting of these SPNs in the A2a-Cre line. D1-Cre (n = 1 female and 1 male) and A2a-Cre rats (n = 1 female and 1 male) were unilaterally infused with a Cre-dependent eYFP virus in the NAc-S (Figure 1 A). To validate expression in these rat lines, other D1-Cre (n = 4 males) and A2a-Cre rats (n = 3 males) were given this unilaterally infusion in the dorsal striatum (DS) with the same virus. In both transgenic strains, viral expression was largely restricted to local SPNs in the two targeted regions: ∼95% of eYFP-positive neurons expressed DARPP-32 (D32 | eYFP; Figures 1 C-F). Viral expression was consistent with D1- and D2-SPNs representing two neuronal populations similar in size: in both transgenic strains and targeted regions ∼40% of DARPP-32-postive neurons expressed eYFP (eYFP | D32; Figures 1 C-F). DS infusion in D1-Cre rats resulted in eYFP-positive terminals (Figures 1 B-C) in the substantia nigra pars reticulata (SNr) and the globus pallidus externus (GPe). By contrast, the same infusion in A2a-Cre rats only produced eYFP-positive terminals in the GPe (Figures 1 B and D). NAc-S infusion in both D1-Cre and A2a-Cre rats resulted in eYFP-positive terminals in the VP (Figures 1 E-F). By contrast, eYFP-positive terminals in the lateral hypothalamus (Figures 1 E-F) were only observed in D1-Cre rats. Overall, these results are in line with current knowledge about the cellular composition of the striatum (Gerfen & Surmeier, 2011; Bertran-Gonzalez et al., 2010) and well-established projection patterns of striatal D1- and D2-SPNs (Gerfen & Surmeier, 2011; Lu et al., 1997; Kupchik et al., 2015; O’Connor et al., 2015). We were therefore confident in the capacity of the two rat transgenic lines to reveal the function of NAc-S D1- and D2- SPNs.

Anterograde tracing in D1-Cre and A2a-Cre rats

A. D1-Cre or A2a-Cre rats were unilaterally infused in the dorsal striatum (DS) or nucleus accumbens shell (NAc-S) with DIO-eYFP. B. Sagittal micrographs obtained in D1-Cre (top) and A2A-Cre (bottom) rats following viral infusion in the DS. C. DIO-eYFP infusion in the DS of D1-Cre rats. Micrographs show eYFP expression in the DS, D32 staining and co-labelling (eYFP + D32) in the DS. They also show that DS D1-SPNs project to the substantia nigra pars reticulata (SNr) and the globus pallidus externus (GPe). Viral expression was restricted to putative SPNs (D32 | eYFP), with ∼40% of SPNs expressing eYFP (eYFP | D32). D. DIO-eYFP infusion in the DS of A2a-Cre rats. Micrographs show eYFP expression in the DS, D32 staining and co-labelling (eYFP + D32) in the DS. They also show that DS D2-SPNs project to the SNr but not the GPe. Viral expression was restricted to putative SPNs (D32 | eYFP), with ∼40% of SPNs expressing eYFP (eYFP | D32). E. DIO-eYFP infusion in the NAc-S of D1-Cre rats. Micrographs show eYFP expression in the NAc-S, D32 staining and co-labelling (eYFP + D32) in the DS. They also show that NAc-S D1-SPNs project to the ventral pallidum (VP) and the lateral hypothalamus (LH). Viral expression was restricted to putative SPNs (D32 | eYFP), with ∼41% of SPNs expressing eYFP (eYFP | D32). F. DIO-eYFP infusion in the NAc-S of A2a-Cre rats. Micrographs show eYFP expression in the NAc-S, D32 staining and co-labelling (eYFP + D32) in the DS. They also show that NAc-S D2-SPNs project to the VP but not the LH. Viral expression was restricted to putative SPNs (D32 | eYFP), with ∼40% of SPNs expressing eYFP (eYFP | D32).

Next, we used ex-vivo electrophysiological recordings to assess our capacity to silence activity of NAc-S D1- and D2-SPNs. D1-Cre and A2a-Cre rats were bilaterally infused in the NAc-S with either a null Cre-dependent eYFP virus (D1-Cre: 2 females; A2a-Cre: 2 females) or a Cre-dependent inhibitory halorhodospin (eNpHR3.0; D1-Cre: 2 females; A2a-Cre: 2 females) virus (Figures 2 A-B). LED light illumination (625 nm) of either NAc-S D1-SPNs (Figure 2 A) or NAc-S D2-SPNs (Figure 2 B) transfected with the eYFP virus had no effect on action potential firing. By contrast, the same illumination onto NAc-S D1- (Figure 2 A) or D2-SPNs (Figure 2 B) transfected with eNpHR3.0 suppressed action potential firing. Importantly, there was no difference in firing prior to or following LED light illumination, indicating that the optical manipulation did not alter D1- or D2-SPNs physiology. Together, these results validate the efficacy of eNpHR3.0 to specifically silence D1- and D2-SPNs activity at the time of LED light delivery.

Ex-vivo cell recordings in D1-Cre and A2a-Cre rats

A. D1-Cre were bilaterally infused in the NAc-S with DIO-eYFP (black) or DIO-eNpHR3.0 (blue). The representative raw traces of cell attached recordings are those of transfected neurons that were depolarized to elicit action potentials by injecting a brief positive current step (+150pA, 200ms duration, 0.5Hz). 625nm LED illumination (orange bar, continuous wave, 2mW) had no effect in eYFP transfected neurons (black) but it inhibited action potential in eNpHR3.0 transfected neurons (blue). The grouped data for recordings include 5 animals (overlapping data points). B. A2a-Cre were bilaterally infused in the NAc-S with DIO-eYFP (black) or DIO-eNpHR3.0 (red). The representative raw traces of cell attached recordings are those of transfected neurons that were depolarized to elicit action potentials by injecting a brief positive current step (+150pA, 200ms duration, 0.5Hz). 625nm LED illumination (orange bar, continuous wave, 2mW) had no effect in eYFP transfected neurons (black) but it inhibited action potential in eNpHR3.0 transfected neurons (red). The grouped data for recordings include 7 animals (overlapping data points).

NAc-S D1-SPNs mediate outcome-specific PIT

This experiment examined the effects of silencing NAc-S D1-SPNs on outcome-specific PIT. D1-Cre rats were bilaterally infused in the NAc-S with either the null Cre-dependent eYFP virus (eYFP: 5 females and 5 males) or the Cre-dependent halorhodopsin (eNpHR3.0: 5 females and 5 males) virus and were implanted with fiber-optic cannulas above the NAc-S (Figure 3 A and Supplementary Figures S1 A-B). Using the outcome-specific PIT protocol depicted in Figure 3 B, the rats first received Pavlovian conditioning during which two stimuli (S1 and S2; white noise or clicker) were paired with two food outcomes (O1 and O2; grain pellets or sucrose solution). Next, the rats underwent instrumental conditioning and learned that pressing one lever earned one of the food outcomes whereas pressing a second lever delivered the other outcome. Finally, an outcome-specific PIT test was administered and assessed choice between the two lever press actions in the presence or absence of each stimulus. Critically, half of the trials for each stimulus took place under LED light stimulation (625 nm, ON trials) whereas the other half of the trials was conducted without the stimulation (OFF trials). Thus, NAc-S D1-SPNs were silenced on half of the trials in rats infused with eNpHR3.0.

NAc-S D1-SPNs mediate outcome-specific PIT

A. D1-Cre rats were bilaterally infused in the NAc-S with DIO-eYFP (black) or DIO-eNpHR3.0 (blue). Fiber-optic cannulas were implanted above the NAc-S to provide 625nm LED illumination B. Schematic representation of the behavioral design; S1 and S2: noise and clicker stimuli (counterbalanced); O1 and O2: grain pellets and sucrose solution (counterbalanced); A1 and A2: left and right lever press (counterbalanced). At test, S1 and S2 were presented 4 times each, in a pseudorandom order. Half of the trials for each stimulus was conducted under 625nm LED illumination (ON; continuous wave; ∼10mW) whereas the LED remained inactivated during the other half of the trials (OFF). ON/OFF trials were counterbalanced. C. Outcome-specific PIT test: net lever presses when the stimuli predicted the same outcome as the action (Same) or when the stimuli predicted the different outcome (Different). Lever presses are shown for each group in trials conducted under 625nm LED illumination (ON) and in trials without illumination (OFF). Data are shown as mean ± SEM. Panel C includes individual data points for female (filled circle) and male (open circle) rats. Asterisks denote significant effect (* p < 0.05; ** p < 0.01; *** p < 0.001; n.s., nonsignificant).

Pavlovian and instrumental conditioning proceeded as expected (Supplementary Figures S1 C-D). The data of most interest are those from the outcome-specific PIT test and are presented in Figure 3 C (see also Supplementary Figure S1 E). They are plotted as the mean number of lever presses per minute when the stimulus predicted the same outcome as the action (Same) and when the stimulus predicted a different outcome to the action (Different). Thus, A1 was identified as “Same” and A2 as “Different” in the presence of S1. Conversely, A2 was identified as “Same” and A1 as “Different” in the presence of S2. Further, baseline responding (lever presses per minute during the two minutes preceding each stimulus presentation) was subtracted from the rates of responding during the stimuli since it did not differ between groups (Group – eYFP vs. eNpHR3.0: p = 0.42). This approach allowed us to focus on the net effect of the stimuli on choice performance.

Silencing NAc-S D1 SPNs eliminated outcome-specific PIT. Overall lever press rates were similar between groups (Group: p = 0.48), LED light condition (LED – ON vs OFF: p = 0.14), and the two factors did not interact (Group x Light: p = 0.07). However, lever press rates were higher on the action earning the same outcome as the stimuli compared to the action earning the different outcome (Lever – Same vs. Different: F(1,18) = 50.06; η2 = 0.74; p < 0.001), regardless of group (Group x Lever: p = 0.77). There was no Lever by LED light condition interaction (Lever x LED: p = 0.72) but critically, there was an interaction between Group, LED light condition, and Lever during the presentation of the predictive stimuli (Group x LED x Lever: F(1,18) = 7.19; η2 = 0.29; p = 0.02). Follow-up analyses revealed that rats in the eYFP group displayed intact outcome-specific PIT whether the LED light was OFF (F(1,9) = 6.14; η2 = 0.67; p < 0.05) or ON (F(1,9) = 39.46; η2 = 0.81; p < 0.001). By contrast, rats in the eNpHR3.0 group displayed outcome-specific PIT when the light was OFF (F(1,9) = 12.38; η2 = 0.58; p < 0.01) but not when it was ON (p = 0.67). Together, these results indicate that activity in NAc-S D1-SPNs is necessary for the outcome-specific influence exerted by predictive stimuli on choice between actions. This is consistent with previous pharmacological work showing that NAc-S D1Rs blockade removes specific PIT (Laurent et al., 2014). Importantly, the impairment produced by silencing NAc-S D1-SPNs was restricted to the influence of predictive stimuli on choice between actions. In the same rats, we found that this silencing had no effect on valued-based choice (Figure S1 F) using an outcome devaluation procedure: all rats performed the action that previously earned a valuable outcome more than the action that earned a devalued outcome. This agrees with studies reporting that NAc-S lesion or inactivation spares value-based choice (Corbit & Balleine, 2011; Corbit et al., 2001; Morse et al., 2020; Laurent et al., 2012; Laurent et al., 2014).

NAc-S D2-SPNs mediate outcome-specific PIT

The removal of outcome-specific PIT under silencing of NAc-S D1-SPNs reproduced the impairment previously observed with blockade of NAc-S D1Rs (Laurent et al., 2014). Interestingly, the latter pharmacological study did not find any effect following NAc-S infusion of a D2R antagonist. To evaluate this finding, the present experiment examined the effects of silencing NAc-S D2-SPNs on outcome-specific PIT. A2a-Cre rats were bilaterally infused in the NAc-S with either the null Cre-dependent eYFP virus (eYFP: 4 females and 4 males) or the Cre-dependent halorhodospin (eNpHR3.0: 4 females and 2 males) virus and were implanted with fiber-optic cannulas above the NAc-S (Figure 4 A and Supplementary Figures S2 A-B). The rats then received the outcome-specific PIT protocol previously described (Figure 3 B).

NAc-S D2-SPNs mediate for outcome-specific PIT

A. A2a-Cre rats were bilaterally infused in the NAc-S with DIO-eYFP (black) or DIO-eNpHR3.0 (red). Fiber-optic cannulas were implanted above the NAc-S to provide 625nm LED illumination. B. Outcome-specific PIT test: net lever presses when the stimuli predicted the same outcome as the action (Same) or when the stimuli predicted the different outcome (Different). Lever presses are shown for each group in trials conducted under 625nm LED illumination (ON) and in trials without illumination (OFF). Data are shown as mean ± SEM. Panel B includes individual data points for female (filled circle) and male (open circle) rats. Asterisks denote significant effect (* p < 0.05; ** p < 0.01; n.s., nonsignificant).

Pavlovian and instrumental conditioning went as expected (Supplementary Figures S2 C-D). The data from the outcome-specific PIT test are presented in Figure 4B, again with baseline subtracted since it did not differ between groups (Group: p = 0.89; see also Supplemental Figure 2 E). Silencing NAc-S D2-SPNs activity eliminated outcome-specific PIT. Overall lever press rates during the test were similar between groups (Group: p = 0.86), LED light condition (LED: p = 0.21), and the two factors did not interact (Group x LED: p = 0.70). Lever press rates were higher on the action earning the same outcome as the stimuli compared to the action earning the different outcome (Lever: F(1,12)= 22.77; η2 = 0.66; p < 0.001), regardless of group (Group x Lever: p = 0.46). There was no Lever by LED light condition interaction (Lever x LED: p = 0.15) but critically, there was an interaction between group, LED light condition, and Lever during the presentation of the predictive stimuli (Group x LED x Lever: F(1,12) = 8.73; η2 = 0.42; p < 0.01). Follow-up analyses revealed that control eYFP rats expressed outcome-specific PIT whether the LED light was OFF (F(1,7) = 16.25; η2 = 0.70; p < 0.01) or ON (F(1,7) = 20.58; η2 = 0.75; p < 0.01). By contrast, rats in the eNpHR3.0 group displayed outcome-specific PIT when the light was OFF (F(1,5) = 6.94; η2 = 0.58; p < 0.05) but not ON (p = 0.34). Thus, these results show for the first time that activity in NAc-S D2-SPNs is necessary for the outcome-specific influence exerted by predictive stimuli on choice between actions.

The present finding contrasts with our previous observation that outcome-specific PIT remains unaffected by NAc-S infusion of a D2R antagonist (Laurent et al., 2014). However, the capacity of such pharmacological manipulation to reveal the function of D2-SPNs is unclear. For instance, a D2R antagonist could be expected to enhance D2-SPN activity, since D2Rs are Gi coupled and so their activation should reduce adenylyl cyclase activity. Further, D2R are not exclusively expressed on this cell population (Gerfen & Surmeier, 2011), as they can be found on local CINs and presynaptic dopamine terminals. Finally, previous work found that pharmacological treatment targeting D2R can leave D2-SPN activity unaffected (Tozzi et al., 2007). These issues are overcome by the present optical manipulation in A2a-Cre rats, and we are therefore confident that NAc-S D2-SPNs activity is in fact critical for outcome-specific PIT. Importantly, we also found that the impairment produced by silencing NAc-S D2-SPNs was restricted to the influence of predictive stimuli on choice between actions. In the same rats, this silencing had no effect on valued-based choice (Figure S2 F).

NAc-S D1-SPNs projections to the VP mediate outcome-specific PIT

Previous work indicates that communication between the NAc-S and VP is critical for outcome-specific PIT (Leung & Balleine, 2013). Consistent with the literature (Lu et al., 1997; Kupchik et al., 2015), we observed that both NAc-S D1- and D2-SPNs densely innervate the VP (Figure 1 E-F). Since we found that activity in both neuron populations mediates the outcome-specific PIT effect, we investigated whether projections from each population contribute to the effect. We first focused on those originating from NAc-S D1-SPNs. D1-Cre rats were bilaterally infused in the NAc-S with either the null Cre-dependent eYFP virus (eYFP: 3 females and 5 males) or the Cre-dependent halorhodopsin (eNpHR3.0: 6 females and 6 males) virus and were implanted with fiber-optic cannulas above the VP (Figure 5 A and Supplementary Figures S3 A-B). The rats then received the behavioral protocol previously described (Figure 3 B).

NAc-S D1-SPNs projections to the VP mediate outcome-specific PIT

A. D1-Cre rats were bilaterally infused in the NAc-S with DIO-eYFP (black) or DIO-eNpHR3.0 (blue). Fiber-optic cannulas were implanted above the VP to provide 625nm LED illumination B. Outcome-specific PIT test: net lever presses when the stimuli predicted the same outcome as the action (Same) or when the stimuli predicted the different outcome (Different). Lever presses are shown for each group in trials conducted under 625nm LED illumination (ON) and in trials without illumination (OFF). Data are shown as mean ± SEM. Panel B includes individual data points for female (filled circle) and male (open circle) rats. Asterisks denote significant effect (** p < 0.01; *** p < 0.01; n.s., nonsignificant).

Pavlovian and instrumental conditioning went as expected (Supplementary Figures S3 C-D). The data from the outcome-specific PIT test are presented in Figure 5 B in the manner described previously since baseline responding did not differ between groups (Group: p = 0.34; see also Supplemental Figure 3 E). Silencing NAc-S D1-SPNs projections to the VP eliminated outcome-specific PIT. Overall lever press rates during the test were similar between groups (Group: p = 0.11). LED activation reduced these rates (LED: F(1,18)= 7.58; η2 = 0.30; p < 0.05) but this reduction depended on the group considered (Group x LED: F(1,18)= 9.78; η2 = 0.35; p < 0.01). Lever press rates were higher on the action earning the same outcome as the stimuli compared to the action earning the different outcome (Lever: F(1,18)= 35.73; η2 = 0.67; p < 0.001), regardless of group (Group x Lever: p = 0.19). There was a Lever by LED light condition interaction (Lever x LED: F(1,18)= 7.56; η2 = 0.29; p < 0.05) and critically, there was an interaction between group, LED light condition, and Lever during the presentation of the predictive stimuli (Group x LED x Lever: F(1,18) = 5.01; η2 = 0.22; p < 0.05). Follow-up analyses revealed that control eYFP rats expressed outcome-specific PIT whether the LED light was OFF (F(1,7) = 26.63; η2 = 0.79; p < 0.001) or ON (F(1,7) = 12.31; η2 = 0.64; p < 0.01). By contrast, rats in the eNpHR3.0 group displayed outcome-specific PIT when the light was OFF (F(1,11) = 26.53; η2 = 0.71; p < 0.001) but not ON (p = 0.48). Thus, these results show for the first time that NAc-S D1-SPNs mediate outcome-specific PIT via their projections to the VP. Importantly, we also found that the impairment produced by silencing NAc-S D1-SPNs terminals in the VP was restricted to the influence of predictive stimuli on choice between actions. In the same rats, this silencing had no effect on valued-based choice (Figure S3 F).

NAc-S D2-SPN projections to the VP do not mediate outcome-specific PIT

Since activity in NAc-S D2-SPNs is required for outcome-specific PIT (Figure 3) and these neurons innervate the VP (Figure 1 F), we assessed whether D2-SPNs projections to the VP are involved in outcome-specific PIT. A2-Cre rats were bilaterally infused in the NAc-S with either the null Cre-dependent eYFP virus (eYFP: 4 females and 5 males) or the Cre-dependent halorhodopsin (eNpHR3.0: 5 females and 6 males) virus and were implanted with fiber-optic cannulas above the VP (Figure 6 A and Supplementary Figures S4 A-B). The rats then received the behavioral protocol previously described (Figure 3 B).

NAc-S D2-SPNs projections to the VP do not mediate outcome-specific PIT

A. A2a-Cre rats were bilaterally infused in the NAc-S with DIO-eYFP (black) or DIO-eNpHR3.0 (red). Fiber-optic cannulas were implanted above the VP to provide 625nm LED illumination. B. Outcome-specific PIT test: net lever presses when the stimuli predicted the same outcome as the action (Same) or when the stimuli predicted the different outcome (Different). Lever presses are shown for each group in trials conducted under 625nm LED illumination (ON) and in trials without illumination (OFF). Data are shown as mean ± SEM. Panel B includes individual data points for female (filled circle) and male (open circle) rats.

Pavlovian and instrumental conditioning went as expected (Supplementary Figures S4 C-D). The data from the outcome-specific PIT test are presented in Figure 6 B as previously since baseline responding did not differ between groups (Group: p = 0.90; see also Supplemental Figure 4 E). Silencing NAc-S D2-SPNs projections to the VP had no effect on outcome-specific PIT. Overall lever press rates were similar between groups (Group: p = 0.56), LED light condition (LED: p = 0.75), and the two factors did not interact (Group x LED: p = 0.73). The rates were higher on the action earning the same outcome as the stimuli relative to the action earning the different outcome (Lever: F(1,13) = 22.71; η2 = 0.64; p < 0.001), irrespective of group (Group x Lever: p = 0.55). There was no interaction between groups, LED light condition, and Lever (Group x Light x Lever: p = 0.59). Thus, NAc-S D2-SPNs do not appear to mediate outcome-specific PIT via their projections to the VP. Likewise, we found no evidence that these projections influence valued-based choice (Figure S4 F).

Discussion

The present experiments investigated the role of NAc-S D1- and D2-SPNs in choice between actions using an outcome-specific PIT task. First, they combined anatomical tract-tracing and ex-vivo electrophysiology to demonstrate that two recently developed knock-in rat lines (Pettibone et al., 2019) enable selective silencing of activity in either population of NAc-S SPNs. Consistent with previous findings (Laurent et al., 2014), they found that NAc-S D1-SPNs are necessary for PIT since their silencing eliminated outcome-specific choice. Additionally, they showed that this choice is also abolished by silencing NAc-S D2-SPNs, providing the first evidence that both SPNs populations contribute to PIT expression. Finally, the last two experiments revealed for the first time that outcome-specific choice is likely to involve downstream regulation of VP function by NAc-S D1-SPNs but not NAc-S D2-SPNs. Together, these findings offer novel insights into the cellular mechanisms that govern the outcome-specific influence of predictive stimuli on choice between actions.

Convincing evidence had been provided for a significant role of NAc-S D1-SPNs in PIT (Laurent et al., 2014). The evidence was based on the observations that PIT is associated with an increase in ERK1/2 phosphorylation within these SPNs and that pharmacological blockade of NAc-S D1Rs eliminates outcome-specific choice between actions. These findings align with the present study, showing that PIT is eliminated when NAc-S D1-SPNs are silenced during presentations of the predictive stimuli at the time of choice. Importantly, the effect of NAc-S D1-SPNs silencing was specific to the assessment of choice between actions in the presence of predictive stimuli. For instance, the ability of the stimuli to elicit approaches towards the magazine where food used to be delivered remained intact despite NAc-S D1-SPNs silencing. Furthermore, the same silencing preserved the capacity to select between the two actions based on the value attributed to their respective outcomes. This selectivity in the impairment produced by NAc-S D1-SPNs is consistent with previous findings demonstrating that the NAc-S does not contribute to learning and retrieving the S-O and A-O associations produced by Pavlovian and instrumental conditioning (Corbit & Balleine, 2011; Corbit et al., 2001; Morse et al., 2020; Laurent et al., 2012; Laurent et al., 2014).

Previous research failed to provide any evidence supporting the involvement of NAc-S D2-SPNs in outcome-specific choice (Laurent et al., 2014). Specifically, PIT was found to produce no change in ERK1/2 phosphorylation in these neurons and was left intact by NAc-S D2Rs blockade. However, these assessments have significant limitations, including the widespread distribution of striatal D2Rs (Gerfen & Surmeier, 2011) and the inability of ERK1/2 phosphorylation or D2Rs antagonism to capture or distort activity in NAc-S D2-SPNs (Bertran-Gonzalez et al., 2008; Tozzi et al., 2007). The present study addressed these limitations by implementing optogenetic silencing in a knock-in rat line giving access to D2-SPNs through the targeting of the A2a receptor that is predominantly expressed in these neurons in the NAc-S (Schiffmann et al., 2007). We found that PIT is eliminated when NAc-S D2-SPNs are silenced during presentations of the predictive stimuli at the time of choice. Moreover, the same silencing left intact the ability of the predictive stimuli to elicit magazine approaches and preserved the capacity to choose between actions based on the value of their outcomes. Thus, we conclude that both NAc-S D1- and D2-SPNs are indispensable to outcome-selective PIT.

Consistent with the literature (Lu et al., 1997; Kupchik et al., 2015), we found that both NAc-S D1- and D2-SPNs send dense projections to the VP, which is often described as the major efferent of the nucleus accumbens (Heimer et al., 1991; Zahm & Heimer, 1990). Importantly, the VP has been shown to play a crucial role in outcome-specific choice (Leung & Balleine, 2013; Leung & Balleine, 2015; Leung et al., 2024b). VP neurons exhibit enhanced cFos activation following an outcome-specific PIT test, and their levels of activation correlate with PIT performance. Moreover, pharmacological inactivation of the VP eliminates outcome-specific choice, and the same elimination is observed when the NAc-S is disconnected from the VP. We therefore tested for a role of NAc-S D1- and D2-SPNs projections to the VP during choice between actions in our outcome-specific PIT task. We found that PIT was abolished by silencing NAc-S D1-SPNs terminals but not by silencing NAc-S D2-SPNs terminals. Neither silencing affected magazine approaches elicited by the predictive stimuli or choice between actions based on the value of their outcomes. It is important to note that our study does not provide any evidence about the efficacy of NAc-S D2-SPNs terminals silencing in the VP, and future experiments should aim to provide such evidence. Yet, the same silencing for NAc-S D1-SPNs terminals resulted in PIT elimination. Therefore, it seems reasonable to conclude that NAc-S D1-SPNs, but not NAc-S D2 SPNs, projections to the VP are indispensable to observe outcome-specific choice in a PIT task. Since NAc-S D2-SPNs appear to exclusively project to the VP (Humphries & Prescott, 2010; Kupchik et al., 2015), our findings suggest that these SPNs mediate outcome-specific PIT by locally regulating NAc-S function. By contrast, it remains to be determined whether NAc-S D1-SPNs also coordinate outcome-specific PIT via their projections to the lateral hypothalamus (O’Connor et al., 2015) and/or the ventral tegmental area (Humphries & Prescott, 2010), in addition to the VP.

The present findings are consistent with a recent model proposing that outcome-specific choice in PIT relies on an opioid-based memory residing in the NAc-S (Leung et al., 2024a; Laurent & Balleine, 2021). In this model, as the basolateral amygdala encodes and stores outcome-specific S-O associations across Pavlovian conditioning, it also drives the formation of the NAc-S memory, which involves the durable accumulation of delta-opioid receptors (DOPRs) on the somatic membrane of local cholinergic interneurons (CINs) (Bertran-Gonzalez et al., 2013; Morse et al., 2020). Although this memory is not necessary for Pavlovian conditioning per se, its expression is later required to enable the predictive stimuli to guide outcome-specific choice during the PIT test (Morse et al., 2020). The model specifically proposes that memory expression is controlled by fluctuations in glutamatergic and dopamine release from cortical and midbrain inputs to the NAc-S. One consequence is to activate NAc-S D2-SPNs and enkephalin discharge by these neurons. Enkephalin then binds onto DOPRs that had accumulated on local CINs and dampens acetylcholine secretion from the interneurons. The sudden drop in NAc-S acetylcholine frees D1-SPNs from the inhibitory tone imposed by acetylcholine occupancy of muscarinic M4 receptors that are exclusively found on these SPNs (Tayebati et al., 2004; Lobo et al., 2006; Guo et al., 2010; Jeon et al., 2010). The ultimate consequence is to promote NAc-S D1-SPNs function, including the coordination of outcome-specific choice in PIT by regulating activity in downstream brain regions such as the VP. The model predicts all the findings presented here. NAc-S D2-SPNs silencing eliminates PIT at the time of choice by preventing enkephalin release and thereby, expression of the DOPR-based memory. The main consequence of this expression is precluded by NAc-S D1-SPNs silencing at the time of choice, while silencing the terminals of these SPNs in the VP squash their ability to regulate the activity of this brain region to coordinate outcome-specific choice in PIT. Thus, one fundamental implication of the present findings is to strengthen the proposal that an opioid-based memory system in the NAc-S enables outcome-specific choice between actions.

In summary, the present experiments found that the two main populations of SPNs in the NAc-S are indispensable for outcome-specific choice between actions in a PIT task. They also revealed that NAc-S D1-SPNs coordinate this choice by downstream regulation of VP activity. By contrast, NAc-S D2-SPNs function in PIT appears to be restricted to modulating local activity. These findings provide novel insights in the cellular mechanisms and circuitry underlying the outcome-specific influence of predictive stimuli on choice between actions, and are consistent with a recent model proposing that this influence is mediated by an opioid-based memory system in the NAc-S. Beyond these mechanistic considerations, the findings offer an opportunity to gain novel insights about various disorders in which the outcome-specific influence of predictive stimuli over our choices and decisions is dysfunctional. These disorders include depression (Geurts et al., 2013; Nord et al., 2018), anxiety disorders (Quail et al., 2017; Krypotos & Engelhard, 2020), substance-use disorders (Heinz et al., 2019; Hogarth et al., 2018; Hogarth et al., 2018; Steins-Loeber et al., 2020; Garbusow et al., 2016; Garbusow et al., 2019; Hogarth et al., 2018), gambling disorders (Genauck et al., 2019), anorexia nervosa (Vogel et al., 2019) and obesity (Watson et al., 2014; Lehner et al., 2017; Meemken & Horstmann, 2019).

Materials and methods

Subjects

Two genetically modified knock-in rat lines were used and obtained from the breeding facility at the University of New South Wales (Sydney, Australia). D1-Cre rats (Rat Research & Resource Center, Columbia, MO, USA; LE-Drd1em1(iCre)Berke, RRRC#: 00856) expressed Cre recombinase in neurons expressing the dopamine D1 receptors (D1R). A2a-Cre rats (Rat Research & Resource Center, Columbia, MO, USA; LE-Adora2aem1(iCre)Berke, RRRC#: 00857) expressed Cre recombinase in neurons expressing the adenosine A2A receptors (A2R). The rats were at least 8 weeks old at the beginning of each experiment. Efforts were made to allocate an equal number of female and male rats in each group. However, these efforts were hampered by exclusions following post-mortem assessments of viral spread and fiber-optic cannulas placement. The final number of female and male rats did not provide sufficient power to analyse an effect of sex. Therefore, all test data present individual performance for female and male rats. Rats were housed in transparent plastic boxes with their littermates (up to four rats per box) in a climate-controlled colony room maintained on a 12 h light-dark cycle (lights on at 7:00am). Behavioural procedures were conducted during the light phase (8:00am to 6:00pm). Water and standard lab chow were available ad libitum prior to the start of each experiment. Rats were food restricted to maintain them at ∼90% of their ad libitum body weight during behavioural protocols. The Animal Care and Ethics Committee at the University of New South Wales approved all procedures.

Viruses

The following viruses were used: AAV5-EF1a-DIO-eYFP (eYFP; AddGene; Watertown, MA, USA) and AAV5-EF1a-DIO-eNpHR3.0-eYFP (eNpHR3.0; UNC Vector Core; Chapel Hill, NC, USA).

Surgeries

Stereotaxic surgery was conducted under isoflurane gas anaesthesia (0.8L/min; induction at 4-5%, maintenance 2-2.5%). Animals were placed in a stereotaxic frame (Kopf Instruments, Tujunga, CA, USA), the surgical area was shaved, and betadine iodine antiseptic solution was applied. At the incision site, bupivacaine hydrochloride (0.5%; 0.1-0.2mL) was injected subcutaneously. In the lower flank, Metacam (1mg/kg) was injected subcutaneously. An incision was made along the midline to expose the skull, the membrane on top of the skull was cleared, and the skull was adjusted to align bregma and lambda on the same horizontal plane.

For viral infusion, small holes were drilled in the skull above the NAc-S using the following coordinates (indicated in mm from bregma): +1.8 or +1.9 anteroposterior (A/P); ±0.85 or ±0.95 mediolateral (M/L); -7.7 or 7.85 dorsoventral (D/V). An Infuse/Withdraw Pump (Standard Infuse/Withdraw Pump 11 Elite; Holliston, Harvard Apparatus, MA, USA) in combination with glass 1µL syringes (86200; Hamilton Company, Reno, NV, USA) were used to perform infusions. 0.5µL of the virus was infused in each hemisphere at a rate of 0.1µL/min. Following each infusion, the needle was left in position for 5-10 min to allow for diffusion. The incision site was closed and secured with surgical sutures (684G; Ethicon, North Ryde, NSW, Australia) or surgical staples (59027; Stoelting, Wood Dale, IL, USA), and betadine antiseptic ointment was applied to the area. Animals then received subcutaneous injections of the antibiotic Duplocillin (0.15ml/kg) and sodium chloride (0.9%, 2mL) and placed on a heat mat for recovery. Once alert and responsive, animals were returned to their home cages and monitored daily as they recovered for a minimum of one week following surgery with ad libitum access to food and water.

Rats received a second surgery to bilaterally implant fibre-optic cannulas (10mm length, 0.48mm diameter; Doric Lenses, Québec, QC, Canada) after completion of the initial 16 days of training (Pavlovian and Instrumental). The surgical area was prepared using the same method as described for viral surgeries. Three jewellers’ screws were inserted into the skull, distributed anterior and posterior to the implantation site and secured in place with dental cement. Fibre-optic cannulas were implanted into the targeted brain region. The following coordinates (indicated in mm from bregma) were used for the NAc-S: +1.8 or +1.9 A/P; -0.85 or -0.95 M/L for left hemisphere; +2.2 M/L with -10° angle for right hemisphere; -6.75 D/V for left hemisphere; -6.8 or -6.9 D/V for right hemisphere. The following coordinates (indicated in mm from bregma) were used for the VP: +0.24 A/P; ±1.8 M/L; -7.5 or - 8.0 D/V. Post-implantation, rats received the same care as described for viral surgeries.

Behavioral apparatus

Training and testing were conducted in 12 identical operant chambers (ENV-007-VP; L 29.53 x W 23.5 x H 27.31 cm; MED Associates, Fairfax, VT, USA) enclosed within light- and sound-resistant cabinets (ENV-018MD; L 59.69 x W 40.64 x H 55.88 cm; MED Associates, Fairfax, VT, USA). The chambers consisted of two stainless-steel side walls, three clear polycarbonate walls and ceiling, and a stainless-steel grid floor. Each chamber was equipped with a recessed magazine connected to a 45mg grain pellet (F0165; Bio-Serv Biotechnologies, Flemington, NJ, USA) dispenser and a pump fitted with a 60mL syringe that delivered a 0.2mL sucrose solution (20% w/v; white sugar; Winc, Erskine Park, NSW, Australia) into the left side of the magazine. Head entries into the magazine were detected via an infrared beam that crossed the magazine opening. One retractable lever was located on either side of the magazine. A house light (3W, 24V) situated at the top of the wall opposite to the food magazine was used to illuminate the chamber during the training and test sessions. Each chamber contained a Sonalert white noise generator (80dB), and a 28V DC mechanical relay that delivered a 2Hz clicker stimulus. Training and testing sessions were programmed and controlled by computers external to the testing rooms using MED-PC V software (MED Associates, Fairfax, VT, USA) which also recorded experimental data from each session. All cabinets were fitted with cameras (IPC-K35AP; Dahua, Artarmon, NSW, Australia) to record and view real-time chamber activity using D-ViewCam DCS-100 software (D-Link, North Ryde, NSW, Australia). The ceiling of these chambers contained a hole which enabled the connection of the patch cable between the LED and each freely moving rat.

Outcome devaluation was conducted in a separate room equipped with 16 individual, open-top plastic boxes with stainless steel wire mesh lids. During devaluation, the lights were dimmed, and each individual chamber was fitted with either a ceramic dish for grain pellet devaluation, or a plastic sipper bottle for sucrose solution devaluation.

Optogenetic equipment

Fibre-optic cannulas (10mm; Doric Lenses, Québec, QC, Canada) consisted of a zircona ferrule contained within a stainless-steel receptacle from which a silica/polymer flat-tip fibre extended from. Fibre-optic patch cords (custom; Doric Lenses, Québec, QC, Canada) were enclosed in a flexible cladding that consisted of one transparent fibre that split into two smaller fibres for attachment to the implanted fibre-optic cannulas. Delivery of the LED light was controlled through a TTL adapter which converted 28V DC output to a TTL transition (MED Associates, Fairfax, VT, USA). The adaptor was connected to an LED driver (Doric Lenses, Québec, QC, Canada) and Connectorised LED light source (Doric Lenses, Québec, QC, Canada) with a fibre-optic rotary joint for attachment to the fibre-optic patch cords. LED light was delivered as orange light (625nm) and measured at least 10mW at the cannula tip using a photometer (PM200; ThorLabs, Newton, NJ, USA) when the LED was switched on.

Behavioral procedures

Pavlovian conditioning

Pavlovian conditioning involved 8 daily 1hr training sessions. Two auditory conditioned stimuli (S1 and S2; white noise or clicker) were paired with two food outcomes (O1 and O2; grain pellets or sucrose solution). Each stimulus was presented for 2min in duration, 4 times each in a pseudorandom order within each training session. Stimulus presentations were separated by pseudorandom intertrial intervals (ITIs) ranging between 3 to 5min in length (4min average). The stimulus order and ITIs were changed daily. Stimulus-outcome (S-O) pairings were counterbalanced within- and between-groups. Food outcomes were delivered on a random time 30s schedule throughout the stimulus presentation. Magazine entries were recorded throughout the session, and conditioned responding was analysed by comparing magazine entries in the 2min prior to stimulus presentation and during the 2min stimulus presentation period.

Instrumental conditioning

During instrumental conditioning, two actions (A1 and A2; left or right lever presses) were trained with the two different outcomes (O1 and O2) in 2 separate daily sessions for 8 days. The order of the sessions was counterbalanced, as were the A-O relationships, which were further counterbalanced with the S-O relationships. Each session ended when 20 outcomes were earned or when 30min had elapsed. For the first 2 days, actions were continuously reinforced (i.e. one action earned one outcome delivery). Across the next three sessions, actions were rewarded on a random ratio 5 (RR5) schedule (i.e. each action delivered an outcome with a probability of 0.2). For the final 3 days, animals were rewarded on an RR10 schedule (i.e. each response delivered an outcome with a probability of 0.1). In the last experiment (NAc-S D1-SPN◊VP projections), rats only received 2 days on RR10. Lever press actions were measured throughout the duration of the session, and performance during instrumental training was assessed using the mean number of lever presses per min.

Pavlovian Instrumental transfer test

Across the three days preceding this test, rats were given one session of Pavlovian retraining and two separate sessions of instrumental retraining. Immediately before the test, patch cables were attached to the fibre-optic cannulas and the LED. During the test, both levers were inserted into the chamber simultaneously and remained in the chamber for the duration of the test. The test was conducted in extinction (i.e. no outcomes were delivered). To reduce baseline (pre S) lever presses, responding on both levers was extinguished for 8min prior to any stimuli being presented. Each stimulus was presented for 2min, four times each during the test in the following order: noise-clicker-clicker-noise-clicker-noise-noise-clicker. Each stimulus presentation was separated by a fixed 3min interval. Critically, the LED was activated during half of the trials (ON) but left inactive during the other half (OFF). The order for ON and OFF trials was counterbalanced. Responses were recorded prior to and during the S presentation. These were then collated into ‘pre’ (lever press rates in the two min prior to S presentation), ‘same’ (when S shares the same outcome with the action) and ‘different’ (when S has a different outcome with the action) following data collection.

Choice tests following outcome devaluation

Across the two consecutive days preceding these tests, rats were given 2 daily instrumental conditioning sessions (RR10). In the last experiment (NAc-S D1-SPNs◊VP projections), rats first received one day of Pavlovian conditioning followed by 3 days of instrumental conditioning (RR10). On the last two days of instrumental conditioning, following the second session, rats were habituated to the devaluation chambers for 30min. During these habituation sessions, animals had access to one outcome (counterbalanced). On the first test day, rats were placed in the devaluation chambers for 1hr where they had ad libitum access to one of the previously trained outcomes (O1 or O2). Once removed from the devaluation boxes, patch cables were attached to the fibre-optic cannulas and the LED and the first choice test was conducted. Both levers were made available for 2min each, four times during the session with an intertrial interval of 2min where the levers were retracted. The LED was activated during 2 of the 2min trials where the levers were available (ON) and was left inactive during the other 2 trials (OFF). The order of the ON and OFF trials was counterbalanced. A similar procedure was implemented the following day except that the other outcome was devalued. Lever presses and magazine entries were recorded while the levers were available.

Tissue preparation and histology

Transcardial fixation and brain sectioning

Rats were euthanised with an intraperitoneal injection of sodium pentobarbital (1mL) and transcardially perfused with 4% paraformaldehyde (PFA) in 0.1M sodium phosphate buffer (PBS, pH 7.5; 400mL for rats). Brains were immediately removed and placed in individual specimen jars containing PFA and stored at 4°C overnight for post-fixation. Brains were sliced into 40μm coronal or sagittal sections in 0.1M Tris-buffered saline (TBS, 0.25M Tris, 0.5M NaCl, pH 7.5) using a vibratome (VT1200; Leica Microsystems, North Ryde, NSW, Australia) and stored at -20°C in cryoprotective solution containing 30% ethylene glycol, 30% glycerol and 0.1M PBS.

Immunofluorescence

Free-floating sections were selected in TBS to contain the region of interest and washed three times for 10 min each in TBS solution. Sections were then incubated in 0.5% Triton-X, 10% NHS in TBS blocking solution for two hours. Sections were washed again in TBS solution three times for 10 min each. Sections were then incubated between 12-48h (dependent on antibody) in a solution containing the primary antibodies in 0.2% Triton-X, 2% NHS in TBS at 4°C. The primary antibodies were the following: purified mouse anti DARPP-32 (1:300; BD Biosciences, Macquarie Park, Australia; #AB398980) and polyclonal rabbit anti-GFP (1:1000; Invitrogen, Mulgrave, Australia; #AB221569). After the specified incubation period, the sections were washed for three times in TBS for 10min each and then transferred into a solution containing the secondary antibodies in TBS solution at room temperature for 1h. The secondary antibodies were the following: donkey anti-mouse Cy3 (1:400; Jackson ImmunoResearch Laboratories, West Grove, USA; #AB2340813) and donkey anti-rabbit Alexa-488 (1:500; Jackson ImmunoResearch Laboratories, West Grove, USA; #AB2313584). Next, sections were washed three times in TBS for 10min and then mounted onto glass slides using Vectashield hardset antifade mounting medium (H-1400; Vector Laboratories, Newark, CA, USA). Sections that did not require immunofluorescence staining were washed three times for 10min each in TBS and mounted using Vectashield hardset antifade mounting medium.

Microscopy and imaging

Imaging of the mounted brain sections was performed on a spinning disk confocal microscope (Zyla 4.2 sCMOS; Andor Technology, Belfast, UK) using the Diskovery multi-modal imaging platform (Oxford Instruments, Belfast, UK). The Diskovery platform was mounted on an inverted microscope body with a motorised stage and Perfect Focus System (Eclipse Ti; Nikon, Melville, NY, USA). Image capture was controlled using the NIS-Elements Software (Nikon, Melville, NY, USA). Single high-magnification field-of-view images were generated by automatic stitching of multiple adjacent frames from within a defined area using the motorised stage (optical magnification: 20X; pixel depth: 16-bit). All images were processed using Fiji ImageJ open-source software (Rueden et al., 2017). Additional images were taken with a confocal microscope (BX61W1; Olympus, Shinjuku, Tokyo, Japan) using Fluoview software (FV1000; Olympus, Shinjuku, Tokyo, Japan) for further placement analysis. The images were used to assess neuronal pathways, as well as the spread of viral infusions and placement of fibre-optic cannulas. Based on the latter, the following numbers of rats were excluded from the final statistical analyses due to misplaced viral or fibre-optics misplacements: NAc-S D1-SPNs silencing (n=18), NAc-S D2-SPNs silencing (n=9), NAc-S D1-SPNs◊VP silencing (n=X), NAc-S D2-SPNs◊VP silencing (n=5).

Electrophysiological recordings

Brain slice preparation

Rats were euthanized under isoflurane gas anesthesia (4% in air). Brains were immediately removed and sliced using a vibratome (VT1200; Leica Microsystems, North Ryde, NSW, Australia) in ice-cold oxygenated sucrose buffer solution (consisting of [in mM]: 241 sucrose, 28 NaHCO3, 11 glucose, 1.4 NaH2PO4, 3.3 KCl, 0.2 CaCl2, 7 MgCl2). Coronal brain slices (300μm thickness) containing the NAc were sampled and maintained at 33°C in a submerged chamber containing physiological saline (consisting of [in mM]: 124 NaCl, 2.5 KCl, 1.25 NaH2PO4, 1 MgCl2, 1 CaCl2, 10 glucose and 26 NaHCO3) and equilibrated with 95% O2 and 5% CO2.

Recordings

Following 1h equilibration, slices were transferred to a recording chamber. An upright microscope (BX50WI; Olympus, Shinjuku, Tokyo, Japan) was used to visualise neurons using differential interference contrast (DIC) Dodt tube optics and eYFP fluorescence, and superfused continuously (1.5mL/min) with oxygenated physiological saline at 33°C. Whole-cell patch-clamp recordings were performed using glass pipettes (tip size 2–5 MΩ) containing internal solution (consisting of [in mM]: 115 K gluconate, 20 NaCl, 1 MgCl2, 10 HEPES, 11 EGTA, 5 Mg-ATP, 0.33 Na-GTP, and 5 phosphocreatine di(tris) salt, pH 7.3, osmolarity 285-290 mOsm/L). Neurons sampled during recording were marked with Biocytin (0.1%) added to the internal solution. A Multiclamp 700B amplifier (Molecular Devices, San Jose, CA, USA) connected to a Macintosh computer and interface ITC-18 (InstruTECH, Longmont, CO, USA) was used for data acquisition. Liquid junction potentials of - 10mV were not corrected. In current-clamp mode, membrane potentials were sampled at 5kHz (low pass filter 2 kHz; Axograph X, Axograph, CA, USA). Stock solution of drug was diluted to working concentration in the extracellular solution immediately before use and applied by continuous superfusion. Data from whole-cell recordings were included in analyses if the neurons appeared healthy under DIC on the monitor screen.

LED stimulation

Transfected neurons with eNpHR3.0 or eYFP in the NAc-S were illuminated by a continuous wave of LED light (625nm, 2 minutes, 2mW; ThorLabs, Newton, NJ, USA) onto the brain slice under the 40X water-immersion objective. Since SPNs in ex-vivo brain slices are not spontaneously active, brief injection of positive current step pulses were applied to make them fire action potentials.

Post-hoc histological analysis

Immediately after physiological recording, brain slices were fixed overnight using 4% PFA in 0.16M phosphate buffer (PB) solution, washed and then transferred into 0.5% Triton-X in PB solution for 3 days to permeabilise cells. Slices were washed, mounted onto glass slides, and then coverslipped with Fluoromount-G mounting medium (Southern Biotech, Birmingham, AL, USA). A 2D projection image was then obtained from a collated image stack using the confocal microscope (X61W1; Olympus, Shinjuku, Tokyo, Japan).

Statistical analyses

The data presented met the assumptions of the statistical test used. The main measures were the number of magazine entries and lever presses per minute. The difference between groups were analyzed using a planned orthogonal contrast procedure controlling the per contrast error rate (Hays, 1963). A first contrast (Group) compared rats that had been infused with the control eYFP virus and rats that had been infused with the inhibitory eNpHR3.0 virus. Changes across days (Day) during Pavlovian and instrumental conditioning were assessed by planned linear trend analyses. Pavlovian conditioning and the PIT tests included a within-subject factor (Period) comparing magazine entry rates during the 2min stimuli against rates during a 2min baseline period immediately preceding the stimuli. During the PIT tests, a within-subject factor (Lever) examined the net (i.e., minus baseline) rates of lever presses on the response (Same) earning the same outcome as predicted by the stimulus against those on the response (Different) earning the different outcome. A second within-subject factor (LED) examined these rats when the LED was activated (ON) or not activated (OFF). The choice tests following outcome devaluation used a within-subject factor of response identity (Lever) comparing rates of lever presses on the response (Valued) earning the valued outcome against those on the response (Devalued) earning the devalued outcome. It also used the LED within-subject factor described above. All analyses were carried out using the PSY statistical software (School of Psychology, the University of New South Wales, Australia). The Type I error rate was controlled at alpha = 0.05 for each contrast tested. If interactions were detected, follow-up simple effects analyses were calculated to determine the source of the interactions. For each significant statistical comparison, we report measures of effect size partial eta-squared (η2; η2 = 0.01 is a small effect, η2 = 0.06 a medium effect and η2 = 0.14 a large effect). Data will be made available upon request.

eLife Assessment

This study provides novel and convincing evidence that both dopamine D1 and D2 expressing neurons in the nucleus accumbens shell are crucial for the expression of cue-guided action selection, a fundamental component of decision-making. The research is systematic and rigorous in using optogenetic inhibition of either D1- or D2-expressing medium spiny neurons in the NAc shell to reveal attenuation of sensory-specific Pavlovian-Instrumental transfer, while largely sparing value-based decision on an instrumental task. Findings in this report build on prior research and resolve some conflicts in the literature regarding decision making.

Reviewer #1 (Public review):

In the current article, Octavia Soegyono and colleagues study "The influence of nucleus accumbens shell D1 and D2 neurons on outcome-specific Pavlovian instrumental transfer", building on extensive findings from the same lab. While there is a consensus about the specific involvement of the Shell part of the Nucleus Accumbens (NAc) in specific stimulus-based actions in choice settings (and not in General Pavlovian instrumental transfer - gPIT, as opposed to the Core part of the NAc), mechanisms at the cellular and circuitry levels remain to be explored. In the present work, using sophisticated methods (rat Cre-transgenic lines from both sexes, optogenetics, and the well-established behavioral paradigm outcome-specific PIT-sPIT), Octavia Soegyono and colleagues decipher the differential contribution of dopamine receptors D1 and D2 expressing spiny projection neurons (SPNs).

After validating the viral strategy and the specificity of the targeting (immunochemistry and electrophysiology), the authors demonstrate that while both NAc Shell D1- and D2-SPNs participate in mediating sPIT, NAc Shell D1-SPNs projections to the Ventral Pallidum (VP, previously demonstrated as crucial for sPIT), but not D2-SPNs, mediates sPIT. They also show that these effects were specific to stimulus-based actions, as value-based choices were left intact in all manipulations.

This is a well-designed study, and the results are well supported by the experimental evidence. The paper is extremely pleasant to read and adds to the current literature.

Reviewer #2 (Public review):

Summary:

This manuscript by Soegyono et al. describes a series of experiments designed to probe the involvement of dopamine D1 and D2 neurons within the nucleus accumbens shell in outcome-specific Pavlovian-instrumental transfer (osPIT), a well-controlled assay of cue-guided action selection based on congruent outcome associations. They used an optogenetic approach to phasically silence NAc shell D1 (D1-Cre mice) or D2 (A2a-Cre mice) neurons during a subset of osPIT trials. Both manipulations disrupted cue-guided action selection but had no effects on negative control measures/tasks (concomitant approach behavior, separate valued guided choice task), nor were any osPIT impairments found in reporter-only control groups. Separate experiments revealed that selective inhibition of NAc shell D1 but not D2 inputs to ventral pallidum was required for osPIT expression, thereby advancing understanding of the basal ganglia circuitry underpinning this important aspect of decision making.

Strengths:

The combinatorial viral and optogenetic approaches used here were convincingly validated through anatomical tract-tracing and ex vivo electrophysiology. The behavioral assays are sophisticated and well-controlled to parse cue and value-guided action selection. The inclusion of reporter-only control groups is rigorous and rules out nonspecific effects of the light manipulation. The findings are novel and address a critical question in the literature. Prior work using less decisive methods had implicated NAc shell D1 neurons in osPIT but suggested that D2 neurons may not be involved. The optogenetic manipulations used in the current study provide a more direct test of their involvement and convincingly demonstrate that both populations play an important role. Prior work had also implicated NAc shell connections to ventral pallidum in osPIT, but the current study reveals the selective involvement of D1 but not D2 neurons in this circuit. The authors do a good job of discussing their findings, including their nuanced interpretation that NAc shell D2 neurons may contribute to osPIT through their local regulation of NAc shell microcircuitry.

Weaknesses:

The current study exclusively used an optogenetic approach to probe the function of D1 and D2 NAc shell neurons. Providing a complementary assessment with chemogenetics or other appropriate methods would strengthen conclusions, particularly the novel demonstration of D2 NAc shell involvement. Likewise, the null result of optically inhibiting D2 inputs to the ventral pallidum leaves open the possibility that a more complete or sustained disruption of this pathway may have impaired osPIT.

Reviewer #3 (Public review):

Summary:

The authors present data demonstrating that optogenetic inhibition of either D1- or D2-MSNs in the NAc Shell attenuates expression of sensory-specific PIT while largely sparing value-based decision on an instrumental task. They also provide evidence that SS-PIT depends on D1-MSN projections from the NAc-Shell to the VP, whereas projections from D2-MSNs to the VP do not contribute to SS-PIT.

Strengths:

This is clearly written. The evidence largely supports the authors' interpretations, and these effects are somewhat novel, so they help advance our understanding of PIT and NAc-Shell function.

Weaknesses:

I think the interpretation of some of the effects (specifically the claim that D1-MSNs do not contribute to value-based decision making) is not fully supported by the data presented.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

In the current article, Octavia Soegyono and colleagues study "The influence of nucleus accumbens shell D1 and D2 neurons on outcome-specific Pavlovian instrumental transfer", building on extensive findings from the same lab. While there is a consensus about the specific involvement of the Shell part of the Nucleus Accumbens (NAc) in specific stimulus-based actions in choice settings (and not in General Pavlovian instrumental transfer - gPIT, as opposed to the Core part of the NAc), mechanisms at the cellular and circuitry levels remain to be explored. In the present work, using sophisticated methods (rat Cre-transgenic lines from both sexes, optogenetics, and the well-established behavioral paradigm outcome-specific PIT-sPIT), Octavia Soegyono and colleagues decipher the diNerential contribution of dopamine receptors D1 and D2 expressing spiny projection neurons (SPNs).

After validating the viral strategy and the specificity of the targeting (immunochemistry and electrophysiology), the authors demonstrate that while both NAc Shell D1- and D2SPNs participate in mediating sPIT, NAc Shell D1-SPNs projections to the Ventral Pallidum (VP, previously demonstrated as crucial for sPIT), but not D2-SPNs, mediates sPIT. They also show that these eNects were specific to stimulus-based actions, as valuebased choices were left intact in all manipulations.

This is a well-designed study, and the results are well supported by the experimental evidence. The paper is extremely pleasant to read and adds to the current literature.

We thank the Reviewer for their positive assessment.

Reviewer 2 (Public Review):

Summary:

This manuscript by Soegyono et al. describes a series of experiments designed to probe the involvement of dopamine D1 and D2 neurons within the nucleus accumbens shell in outcome-specific Pavlovian-instrumental transfer (osPIT), a well-controlled assay of cueguided action selection based on congruent outcome associations. They used an optogenetic approach to phasically silence NAc shell D1 (D1-Cre mice) or D2 (A2a-Cre mice) neurons during a subset of osPIT trials. Both manipulations disrupted cue-guided action selection but had no eNects on negative control measures/tasks (concomitant approach behavior, separate valued guided choice task), nor were any osPIT impairments found in reporter-only control groups. Separate experiments revealed that selective inhibition of NAc shell D1 but not D2 inputs to ventral pallidum was required for osPIT expression, thereby advancing understanding of the basal ganglia circuitry underpinning this important aspect of decision making.

Strengths:

The combinatorial viral and optogenetic approaches used here were convincingly validated through anatomical tract-tracing and ex vivo electrophysiology. The behavioral assays are sophisticated and well-controlled to parse cue and value-guided action selection. The inclusion of reporter-only control groups is rigorous and rules out nonspecific eNects of the light manipulation. The findings are novel and address a critical question in the literature. Prior work using less decisive methods had implicated NAc shell D1 neurons in osPIT but suggested that D2 neurons may not be involved. The optogenetic manipulations used in the current study provide a more direct test of their involvement and convincingly demonstrate that both populations play an important role. Prior work had also implicated NAc shell connections to ventral pallidum in osPIT, but the current study reveals the selective involvement of D1 but not D2 neurons in this circuit. The authors do a good job of discussing their findings, including their nuanced interpretation that NAc shell D2 neurons may contribute to osPIT through their local regulation of NAc shell microcircuitry.

We thank the Reviewer for their positive assessment.

Weaknesses:

The current study exclusively used an optogenetic approach to probe the function of D1 and D2 NAc shell neurons. Providing a complementary assessment with chemogenetics or other appropriate methods would strengthen conclusions, particularly the novel demonstration of D2 NAc shell involvement. Likewise, the null result of optically inhibiting D2 inputs to the ventral pallidum leaves open the possibility that a more complete or sustained disruption of this pathway may have impaired osPIT.

We acknowledge the reviewer's valuable suggestion that demonstrating NAc-S D1- and D2-SPNs engagement in outcome-specific PIT through another technique would strengthen our optogenetic findings. Several approaches could provide this validation. Chemogenetic manipulation, as the reviewer suggested, represents one compelling option. Alternatively, immunohistochemical assessment of phosphorylated histone H3 at serine 10 (P-H3) oMers another promising avenue, given its established utility in reporting striatal SPNs plasticity in the dorsal striatum (Matamales et al., 2020). We hope to complete such an assessment in future work since it would address the limitations of previous work that relied solely on ERK1/2 phosphorylation measures in NAc-S SPNs (Laurent et al., 2014). The manuscript was modified to report these future avenues of research (page 12).

Regarding the null result from optical silencing of D2 terminals in the ventral pallidum, we agree with the reviewer's assessment. While we acknowledge this limitation in the current manuscript (page 13), we aim to address this gap in future studies to provide a more complete mechanistic understanding of the circuit.

Reviewer 3 (Public Review):

Summary:

The authors present data demonstrating that optogenetic inhibition of either D1- or D2MSNs in the NAc Shell attenuates expression of sensory-specific PIT while largely sparing value-based decision on an instrumental task. They also provide evidence that SS-PIT depends on D1-MSN projections from the NAc-Shell to the VP, whereas projections from D2-MSNs to the VP do not contribute to SS-PIT.

Strengths:

This is clearly written. The evidence largely supports the authors' interpretations, and these eNects are somewhat novel, so they help advance our understanding of PIT and NAc-Shell function.

We thank the Reviewer for their positive assessment.

Weaknesses:

I think the interpretation of some of the eNects (specifically the claim that D1-MSNs do not contribute to value-based decision making) is not fully supported by the data presented.

We appreciate the reviewer's comment regarding the marginal attenuation of valuebased choice observed following NAc-S D1-SPN silencing. While this manipulation did produce a slight reduction in choice performance, the behavior remained largely intact. We are hesitant to interpret this marginal eMect as evidence for a direct role of NAc-S D1SPNs in value-based decision-making, particularly given the substantial literature demonstrating that NAc-S manipulations typically preserve such choice behavior (Corbit et al., 2001; Corbit & Balleine, 2011; Laurent et al., 2012). Furthermore, previous work has shown that NAc-S D1 receptor blockade impairs outcome-specific PIT while leaving value-based choice unaMected (Laurent et al., 2014). We favor an alternative explanation for our observed marginal reduction. As documented in Supplemental Figure 1, viral transduction extended slightly into the nucleus accumbens core (NAc-C), a region established as critical for value-based decision-making (Corbit et al., 2001; Corbit & Balleine, 2011; Laurent et al., 2012; Parkes et al., 2015). The marginal impairment may therefore reflect inadvertent silencing of a small number of NAc-C D1-SPNs rather than a functional contribution from NAc-S D1-SPNs. Future studies specifically targeting larger NAc-C D1-SPN populations would help clarify this possibility and provide definitive resolution of this question.

Reviewer 1 (Recommendations for the Author):

My main concerns and comments are listed below.

(1) Could the authors provide the "raw" data of the PIT tests, such as PreSame vs Same vs PreDiNerent vs DiNerent? Could the authors clarify how the Net responding was calculated? Was it Same minus PreSame & DiNerent minus PreDiNerent, or was the average of PreSame and PreDiNerent used in this calculation?

The raw data for PIT testing across all experiments are now included in the Supplemental Figures (Supplemental Figures S1E, S2E, S3E, and S4E). Baseline responding was quantified as the average number of lever presses per minute for both actions during the two-minute period (i.e., average of PreSame and PreDiMerent) preceding each stimulus presentation. This methodology has been clarified in the revised manuscript (page 7).

(2) While both sexes are utilized in the current study, no statistical analysis is provided. Can the authors please comment on this point and provide these analyses (for both training and tests)?

As noted in the original manuscript, the final sample sizes for female and male rats were insuMicient to provide adequate statistical power for sex-based analyses (page 15). To address this limitation, we have now cited a previous study from our laboratory (Burton et al., 2014) that conducted such analyses with suMicient power in identical behavioural tasks. That study identified only marginal sex diMerences in performance, with female rats exhibiting slightly higher magazine entry rates during Pavlovian conditioning. Importantly, no diMerences were observed in outcome-specific PIT or value-based choice performance between sexes.

(3) Regarding Figure 1 - Anterograde tracing in D1-Cre and A2a-Cre rats (from line 976), I have one major and one minor question:

(3.1) I do not understand the rationale of showing anterograde tracing from the Dorsal Striatum (DS) as this region is not studied in the current work. Moreover, sagittal micrographs of D1-Cre and A2a-Cre would be relevant here. Could the authors please provide these micrographs and explain the rationale for doing tracing in DS?

We included dorsal striatum (DS) tracing data as a reference because the projection patterns of D1 and D2 SPNs in this region are well-established and extensively characterized, in contrast to the more limited literature on these cell types in the NAc-S. Regarding the comment about sagittal micrographs, we are uncertain of the specific concern as these images are presented in Figure 1B.

If the reviewer is requesting sagittal micrographs for NAc-S anterograde tracing, we did not employ this approach because: (1) the NAc-S and ventral pallidum are anatomically adjacent regions and (2) the medial-lateral coordinates of the ventral pallidum and lateral hypothalamus do not align optimally with those of the NAc-S, limiting the utility of sagittal analysis for these projections.

(3.2) There is no description about how the quantifications were done: manually? Automatically? What script or plugin was used? If automated, what were the thresholding conditions? How many brain sections along the anteroposterior axis? What was the density of these subpopulations? Can the authors include a methodological section to address this point?

We apologize for the omission of quantification methods used to assess viral transduction specificity. This methodological description has now been added to the revised manuscript (page 22). Briefly, we employed a manual procedure in two sections per rat, and cell counts were completed in a defined region of interest located around the viral infusion site.

(4) Lex A & Hauber (2008) Dopamine D1 and D2 receptors in the nucleus accumbens core and shell mediate Pavlovian-instrumental transfer. Learning & memory 15:483- 491, should be cited and discussed. It also seems that the contribution of the main dopaminergic source of the brain, the ventral tegmental area, is not cited, while it has been investigated in PIT in at least 3 studies regarding sPIT only, notably the VP-VTA pathway (Leung & Balleine 2015, accurately cited already).

We did not include the Lex & Hauber (2008) study because its experimental design (single lever and single outcome) prevents diMerentiation between the eMects of Pavlovian stimuli on action performance (general PIT) versus action selection (outcome-specific PIT, as examined in the present study). Drawing connections between their findings and our results would require speculative interpretations regarding whether their observed eMects reflect general or outcome-specific PIT mechanisms, which could distract from the core findings reported in the article.

Several studies examining the role of the VTA in outcome-specific PIT were referenced in the manuscript's introduction. Following the reviewer's recommendation, these references have also been incorporated into the discussion section (page 13).

(5) While not directly the focus of this study, it would be interesting to highlight the accumbens dissociation between General vs Specific PIT, and how the dopaminergic system (diNerentially?) influences both forms of PIT.

We agree with the reviewer that the double dissociation between nucleus accumbens core/shell function and general/specific PIT is an interesting topic. However, the present manuscript does not examine this dissociation, the nucleus accumbens core, or general PIT. Similarly, our study does not directly investigate the dopaminergic system per se. We believe that discussing these topics would distract from our core findings and substantially increase manuscript length without contributing novel data directly relevant to these areas.

(6) While authors indicate that conditioned response to auditory stimuli (magazine visits) are persevered in all groups, suggesting intact sensitivity to the general motivational properties of reward-predictive stimuli (lines 344, 360), authors can't conclude about the specificity of this behavior i.e. does the subject use a mental representation of O1 when experiencing S1, leading to a magazine visits to retrieve O1 (and same for S2-O2), or not? Two food ports would be needed to address this question; also, authors should comment on the fact that competition between instrumental & pavlovian responses does not explain the deficits observed.

We agree with the Reviewer that magazine entry data cannot be used to draw conclusions about specificity, and we do not make such claims in our manuscript. We are therefore unclear about the specific concern being raised. Following the Reviewer’s recommendation, we have commented on the fact that response competition could not explain the results obtained (page 11, see also supplemental discussion).

The minor comments are listed below.

(7) A high number of rats were excluded (> 32 total), and the number of rats excluded for NAc-S D1-SPNs-VP is not indicated.

We apologize for omitting the number of rats excluded from the experiment examining NAc-S D1-SPN projections to the ventral pallidum. This information has been added to the revised manuscript (page 22).

(7.1) Can authors please comment on the elevated number of exclusions?

A total of 133 rats were used across the reported experiments, with 40 rats excluded based on post-mortem analyses. This represents an attrition rate of approximately 30%, which we consider reasonable given that most animals received two separate viral infusions and two separate fiber-optic cannula implantations, and that the inclusion of both female and male rats contributed to some variability in coordinates and so targeting.

(7.2) Can authors please present the performance of these animals during the tasks (OFF conditions, and for control ones, both ON & OFF conditions)?

Rats were excluded after assessing the spread of viral infusions, placement of fibre-optic cannulas and potential damage due to the surgical procedures (page 21). The requested data are presented below and plotted in the same manner as in Figures 3-6. The pattern of performance in excluded animals was highly variable.

(8) For tracing, only males were used, and for electrophysiology, only females were used.

(8.1) Can authors please comment on not using both sexes in these experiments?

We agree that equal allocation of female and male rats in the experiments presented in Figures 1-2 would have been preferable. Animal availability was the sole factor determining these allocations. Importantly, both female and male D1-Cre and A2A-Cre rats were used for the NAc-S tracing studies, and no sex diMerences were observed in the projection patterns. The article describing the two transgenic lines of rats did not report any sex diMerence (Pettibone et al., 2019).

(8.2) Is there evidence in the literature that the electrophysiological properties of female versus male SPNs could diNer?

The literature indicates that there is no sex diMerence in the electrophysiological properties of NAc-S SPNs (Cao et al., 2018; Willett et al., 2016).

(8.3) It seems like there is a discrepancy between the number of animals used as presented in the Figure 2 legend versus what is described in the main text. In the Figure legend, I understand that 5 animals were used for D1-Cre/DIO-eNpHR3.0 validation, and 7 animals for A2a-Cre/DIO-eNpHR3.0; however, the main text indicates the use of a total of 8 animals instead of the 12 presented in the Figure legend. Can authors please address this mismatch or clarify?

The number of rats reported in the main text and Figure 2 legend was correct. However, recordings sometimes involved multiple cells from the same animal, and this aspect of the data was incorrectly reported and generated confusion. We have clarified the numbers in both the main text and Figure 2 legend to distinguish between animal counts and cell counts.

(9) Overall, in the study, have the authors checked for outliers?

Performance across all training and testing stages was inspected to identify potential behavioral outliers in each experiment. Abnormal performance during a single session within a multi-session stage was not considered suMicient grounds for outlier designation. Based on these criteria, no subjects remaining after post-mortem analyses exhibited performance patterns warranting exclusion through statistical outlier analysis. However, we have conducted the specific analyses requested by the Reviewer, as described below.

(9.1) In Figure 3, it seems that one female in the eYFP group, in the OFF situation, for the diNerent condition, has a higher level of responding than the others. Can authors please confirm or refute this visual observation with the appropriate statistical analysis?

Statistical analysis (z-score) confirmed the reviewer's observation regarding responding of the diMerent action in the OFF condition for this subject (|z| = 2.58). Similar extreme responding was observed in the ON condition (|z| = 2.03). Analyzing responding on the diMerent action in isolation is not informative in the context of outcome-specific PIT. Additional analyses revealed |z| < 2 when examining the magnitude of choice discrimination in outcome-specific PIT (i.e., net same versus net diMerent responding) in both ON and OFF conditions. Furthermore, this subject showed |z| < 2 across all other experimental stages. Based on these analyses, we conclude that the subject should be kept in all analyses.

(9.2) In Figure 5, it seems that one male, in the ON situation, in the diNerent condition, has a quite higher level of responding - is this subject an outlier? If so, how does it aNect the statistical analysis after being removed? And who is this subject in the OFF condition?

The reviewer has identified two diMerent male rats infused with the eNpHR3.0 virus and has asked closer examination of their performance.

The first rat showed outlier-level responding on the diMerent action in the ON condition (|z| = 2.89) but normal responding for all other measures across LED conditions (|z| < 2). Additional analyses revealed |z| = 2.55 when examining choice discrimination magnitude in outcome-specific PIT during the ON condition but not during the OFF condition (|z| = 0.62). This subject exhibited |z| < 2 across all other experimental stages.

The second rat showed outlier-level responding on the same action in the OFF condition (|z| = 2.02) but normal responding for all other measures across LED conditions (|z| < 2). Additional analyses revealed |z| = 2.12 when examining choice discrimination magnitude in outcome-specific PIT during the OFF condition but not during the ON condition (|z| = 0.67). This subject also exhibited |z| < 2 across all other experimental stages.

We excluded these two subjects and conducted the same analyses as described in the original manuscript. Baseline responding did not diMer between groups (p = 0.14), allowing to look at the net eMect of the stimuli. Overall lever presses were greater in the eYFP rats (Group: F(1,16) = 6.08, p < 0.05; η2 = 0.28) and were reduced by LED activation (LED: F(1,16) = 9.52, p < 0.01; η2 = 0.44) and this reduction depended on the group considered (Group x LED: F(1,16) = 12.125, p < 0.001; η2 = 0.43). Lever press rates were higher on the action earning the same outcome as the stimuli compared to the action earning the diMerent outcome (Lever: F(1,16)= 49.32; η2 = 0.76; p < 0.001), regardless of group (Group x Lever: p = 0.14). There was a Lever by LED light condition interaction (Lever x LED: F(1,16)= 5.25; η2 = 0.24; p < 0.05) but no an interaction between group, LED light condition, and Lever during the presentation of the predictive stimuli (p = 0.10). Given the significant Group x LED and Lever x LED interactions, additional analyses were conducted to determine the source of these interactions. In eYFP rats, LED activation had no eMect (LED: p = 0.70) and lever presses were greater on the same action (Lever: (F(1,9) = 23.94, p < 0.001; η2 = 0.79) regardless of LED condition (LED x Lever: p = 0.72). By contrast, in eNpHR3.0 rats, lever presses were reduced by LED activation (LED: F(1,9) = 23.97, p < 0.001; η2 = 0.73), were greater on the same action (Lever: F(1,9) = 16.920, p < 0.001; η2 = 0.65) and the two factors interacted (LED x Lever: F(1,9) = 9.12, p < 0.01; η2 = 0.50). These rats demonstrated outcome-specific PIT in the OFF condition (F(1,9) = 27.26, p < 0.001; η2 = 0.75) but not in the ON condition (p = 0.08).

Overall, excluding these two rats altered the statistical analyses, but both the original and revised analyses yielded the same outcome: silencing the NAc-S D1-SPN to VP pathway disrupted PIT. More importantly, we do not believe there are suMicient grounds to exclude the two rats identified by the reviewer. These animals did not display outlier-level responding across training stages or during the choice test. Their potential classification as outliers would be based on responding during only one LED condition and not the other, with notably opposite patterns between the two rats despite belonging to the same experimental group.

(10) I think it would be appreciable if in the cartoons from Figure 5.A and 6.A, the SPNs neurons were color-coded as in the results (test plots) and the supplementary figures (histological color-coding), such as D1- in blue & D2-SPNs in red.

Our current color-coding system uses blue for D1-SPNs transduced with eNpHR3.0 and red for D2-SPNs transduced with eNpHR3.0. The D1-SPNs and D2-SPNs shown in Figures 5A and 6A represent cells transduced with either eYFP (control) or eNpHR3.0 virus and therefore cannot be assigned the blue or red color, which is reserved for eNpHR3.0transduced cells specifically. The micrographs in the Supplemental Figures maintain consistency with the color-coding established in the main figures.

(11) As there are (relatively small) variations in the control performance in term of Net responding (from ~3 to ~7 responses per min), I wonder what would be the result of pooling eYFP groups from the two first experiments (Figures 3 & 4) and from the two last ones (Figures 5 & 6) - would the same statically results stand or vary (as eYFP vs D1-Cre vs A2a-Cre rats)? In particular for Figures 3 & 4, with and without the potential outlier, if it's indeed an outlier.

We considered the Reviewer’s recommendation but do not believe the requested analysis is appropriate. The Reviewer is requesting the pooling of data from subjects of distinct transgenic strains (D1-Cre and A2A-Cre rats) that underwent surgical and behavioral procedures at diMerent time points, sometimes months apart. Each experiment was designed with necessary controls to enable adequate statistical analyses for testing our specific hypotheses.

(12) Presence of cameras in operant cages is mentioned in methods, but no data is presented regarding recordings, though authors mention that they allow for real-time observations of behavior. I suggest removing "to record" or adding a statement about the fact that no videos were recorded or used in the present study.

We have removed “to record” from the manuscript (page 18).

(13) In all supplementary Figures, "F" is wrongly indicated as "E".

We thank the Reviewer for reporting these errors, which have been corrected.

(14) While the authors acknowledge that the eNicacy of optogenetic inhibition of terminals is questionable, I think that more details are required to address this point in the discussion (existing literature?). Maybe, the combination of an anterograde tracer from SPNs to VP, to label VP neurons (to facilitate patching these neurons), and the Credependent inhibitory opsin in the NAc Shell, with optogenetic illumination at the level of the VP, along with electrophysiological recordings of VP neurons, could help address this question but may, reasonably, seem challenging technically.

Our manuscript does not state that optogenetic inhibition of terminals is questionable. It acknowledges that we do not provide any evidence about the eMicacy of the approach. Regardless, we have provided additional details and suggestions to address this lack of evidence (page 13).

(15) A nice addition could be an illustration of the proposed model (from line 374), but it may be unnecessary.

We have carefully considered the reviewer's recommendation. The proposed model is detailed in three published articles, including one that is freely accessible, which we have cited when presenting the model in our manuscript (page 14). This reference should provide interested readers with easy access to a comprehensive illustration of the model.

Reviewer 2 (Recommendations for the Author):

As noted in my public comments, this is a truly excellent and compelling study. I have only a few minor comments.

(1) I could not find the coordinates/parameters for the dorsal striatal AAV injections for that component of the tract tracing experiment.

We apologize for this omission, which has now been corrected (page 16).

(2) Please add the final group sizes to the figure captions.

We followed the Reviewer’s recommendation and added group sizes in the main figure captions.

(3) The discussion of group exclusions (p 21 line 637) seems to accidentally omit (n = X) the number of NAc-S D1-SPNs-VP mice excluded.

We apologize for this omission, which has now been corrected (page 22).

(4) There were some labeling issues in the supplementary figures (perhaps elsewhere, too). Specifically, panel E was listed twice (once for F) in captions.

We apologize for this error, which has now been corrected.

(5) Inspection of the magazine entry data from PIT tests suggests that the optogenetic manipulations may have had some eNects on this behavior and would encourage the authors to probe further. There was a significant group diNerence for D1-SPN inhibition and a marginal group eNect for D2-SPNs. The fact that these eNects were in opposite directions is intriguing, although not easily interpreted based on the canonical D1/D2 model. Of course, the eNects are not specific to the light-on trials, but this could be due to carryover into light-oN trials. An analysis of trial-order eNects seems crucial for interpreting these eNects. One might also consider normalizing for pre-test baseline performance. Response rates during Pavlovian conditioning seem to suggest that D2eNpHR mice showed slightly higher conditioned responding during training, which contrasts with their low entry rates at test. I don't see any of this as problematic -- but more should be done to interpret these findings.

We thank the reviewer for raising this interesting point regarding magazine entry rates. Since these data are presented in the Supplemental Figures, we have added a section in the Supplemental Material file that elaborates on these findings. This section does not address trial order eMects, as trial order was fully counterbalanced in our experiments and the relevant statistical analyses would lack adequate power. Baseline normalization was not conducted because the reviewer's suggestion was based on their assumption that eNpHR3.0 rats in the D2-SPNs experiment showed slightly higher magazine entries during Pavlovian training. However, this was not the case. In fact, like the eNpHR3.0 rats in the D1-SPNs experiment, they tended to display lower magazine entries during training. The added section therefore focuses on the potential role of response competition during outcome-specific PIT tests. Although we concluded that response competition cannot explain our findings, we believe it may complicate interpretation of magazine entry behavior. Thus, we recommend that future studies examine the role of NAc-S SPNs using purely Pavlovian tasks. It is worth nothing that we have recently completed experiments (unpublished) examining NAc-S D1- and D2-SPN silencing during stimulus presentation in a Pavlovian task identical to the one used here. Silencing of either SPN population had no eMect on magazine entry behavior.

Reviewer 3 (Recommendations for the Author):

Broad comments:

Throughout the manuscript, the authors draw parallels between the eNect established via pharmacological manipulations and those shown here with optogenetic manipulation. I understand using the pharmacological data to launch this investigation, but these two procedures address very diNerent physiological questions. In the case of a pharmacological manipulation, the targets are receptors, wherever they are expressed, and in the case of D2 receptors, this means altering function in both pre-synaptically expressed autoreceptors and post-synaptically expressed D2 MSN receptors. In the case of an optogenetic approach, the target is a specific cell population with a high degree of temporal control. So I would just caution against comparing results from these types of studies too closely.

Related to this point is the consideration of the physiological relevance of the manipulation. Under normal conditions, dopamine acts at D1-like receptors to increase the probability of cell firing via Ga signaling. In contrast, dopamine binding of D2-like receptors decreases the cell's firing probability (signaling via Gi/o). Thus, shunting D1MSN activation provides a clear impression of the role of these cells and, putatively, the role of dopamine acting on these cells. However, inhibiting D2-MSNs more closely mimics these cells' response to dopamine (though optogenetic manipulations are likely far more impactful than Gi signaling). All this is to say that when we consider the results presented here in Experiment 2, it might suggest that during PIT testing, normal performance may require a halting of DA release onto D2-MSNs. This is highly speculative, of course, just a thought worth considering.

We agree with the comments made by the Reviewer, and the original manuscript included statements acknowledging that pharmacological approaches are limited in the capacity to inform about the function of NAc-S SPNs (pages 4 and 9). As noted by the Reviewer, these limitations are especially salient when considering NAc-S D2-SPNs. Based on the Reviewer’s comment, we have modified our discussion to further underscore these limitations (page 12). Finally, we agree with the suggestion that PIT may require a halting of DA release onto D2-SPNs. This is consistent with the model presented, whereby D2-SPNs function is required to trigger enkephalin release (page 13).

Section-Specific Comments and Questions:

Results:

Anterograde tracing and ex vivo cell recordings in D1 Cre and A2a Cre rats: Why are there no statistics reported for the e-phys data in this section? Was this merely a qualitative demonstration? I realize that the A2a-Cre condition only shows 3 recordings, so I appreciate the limitations in analyzing the data presented.

The reviewer is correct that we initially intended to provide a qualitative demonstration. However, we have now included statistical analyses for the ex vivo recordings. It is important to note that there were at least 5 recordings per condition, though overlapping data points may give the impression of fewer recordings in certain conditions. We have provided the exact number of recordings in both the main text (page 5) and figure legend.

What does trial by trial analysis look like, because in addition to the eNects of extinction, do you know if the responsiveness of the opsin to light stimulation is altered after repeated exposures, or whether the cells themselves become compromised in any way with repeated light-inhibition, particularly given the relatively long 2m duration of the trial.

The Reviewer raises an interesting point, and we provide complete trial-by-trial data for each experiment below. As identified by the Reviewer, there is some evidence for extinction, although it remained modest. Importantly, the data suggest that light stimulation did not aMect the physiology of the targeted cells. In eNpHR3.0 rats, performance across OFF trials remained stable (both for Same and DiMerent) even though they were preceded by ON trials, indicating no carryover eMects from optical stimulation.

The statistics for the choice test are not reported for eNpHR-D1-Cre rats, but do show a weakening of the instrumental devaluation eNect "Group x Lever x LED: F1,18 = 10.04, p < 0.01, = 0.36". The post hoc comparisons showed that all groups showed devaluation, but it is evident that there is a weakening of this eNect when the LED was on (η2 = 0.41) vs oN (η2 = 0.78), so I think the authors should soften the claim that NAcS-D1s are not involved in value-based decision-making. (Also, there is a typo in the legend in Figure S1, where the caption for panel "F" is listed as "E".) I also think that this could be potentially interesting in light of the fact that with circuit manipulation, this same weakening of the instrumental devaluation eNect was not observed. To me, this suggests that D1-NAcS that project to a diNerent region (not VP) contribute to value-based decision making.

This comment overlaps with one made in the Public Review, for which we have already provided a response. Given its importance, we have added a section addressing this point in the supplemental discussion of the Supplementary Material file, which aligns with the location of the relevant data. The caption labelling error has been corrected.

Materials and Methods:

Subjects:

Were these heterozygous or homozygous rats? If hetero, what rats were used for crossbreeding (sex, strain, and vendor)? Was genotyping done by the lab or outsourced to commercial services? If genotyping was done within the lab, please provide a brief description of the protocol used. How was food restriction established and maintained (i.e., how many days to bring weights down, and was maintenance achieved by rationing or by limiting ad lib access to food for some period in the day)?

The information requested by the Reviewer have been added to the subjects section (pages 15-16).

Were rats pair/group housed after implantation of optic fibers?

We have clarified that rats were group houses throughout (see subjects section; pages 15-16).

Behavioral Procedures:

How long did each 0.2ml sucrose infusion take? For pellets, for each US delivery, was it a single pellet or two in quick succession?

We have modified the method section to indicate that the sucrose was delivered across 2 seconds and that a single pellet was provided (page 17).

The CS to ITI duration ratio is quite low. Is there a reason such a short ratio was used in training?

These parameters are those used in all our previous experiments on outcome-specific PIT. There is no specific reason for using such a ratio, except that it shortens the length of the training session.

Relative to the end of training, when were the optical implantation surgeries conducted, and how much recovery time was given before initiating reminder training and testing?

Fibre-optic implantation was conducted 3-4 days after training and another 3-4 days were given for recovery. This has been clarified in the Materials and methods section (pages 15-16).

I think a diagram or schematic showing the timeline for surgeries, training, and testing would be helpful to the audience.

We opted for a text-based experimental timeline rather than a diagram due to slight temporal variations across experiments (page 15).

On trials, when the LED was on, was light delivered continuously or pulsed? Do these opto-receptors 'bleach' within such a long window?

We apologize for the lack of clarity; the light was delivered continuously. We have modified the manuscript (pages 6 and 19) and figure legend accordingly. The postmortem analysis did not provide evidence for photobleaching (Supplemental Figures) and as noted above, the behavioural results do not indicate any negative physiological impact on cell function.

Immunofluorescence: The blocking solution used during IHC is described as "NHS"; is this normal horse serum?

The Reviewer is correct; NHS stands for normal horse serum. This has been added (page 21).

Microscopy and imaging:

For the description of rats excluded due to placement or viral spread problems, an n=X is listed for the NAc S D1 SPNs --> VP silencing group. Is this a typo, or was that meant to read as n=0? Also, was there a major sex diNerence in the attrition rate? If so, I think reporting the sex of the lost subjects might be beneficial to the scientific community, as it might reflect a need for better guidance on sex-specific coordinates for targeting small nuclei.

We apologize for the error regarding the number of excluded animals. This error has been corrected (page 23). There were no major sex diMerences in the attrition rate. The manuscript has been updated to provide information about the sex of excluded animals (page 23).

References

Cao, J., Willett, J. A., Dorris, D. M., & Meitzen, J. (2018). Sex DiMerences in Medium Spiny Neuron Excitability and Glutamatergic Synaptic Input: Heterogeneity Across Striatal Regions and Evidence for Estradiol-Dependent Sexual DiMerentiation. Front Endocrinol (Lausanne), 9, 173. https://doi.org/10.3389/fendo.2018.00173

Corbit, L. H., Muir, J. L., Balleine, B. W., & Balleine, B. W. (2001). The role of the nucleus accumbens in instrumental conditioning: Evidence of a functional dissociation between accumbens core and shell. J Neurosci, 21(9), 3251-3260. http://eutils.ncbi.nlm.nih.gov/entrez/eutils/elink.fcgi?dbfrom=pubmed&id=11312 310&retmode=ref&cmd=prlinks

Corbit, L. H., & Balleine, B. W. (2011). The general and outcome-specific forms of Pavlovian-instrumental transfer are diMerentially mediated by the nucleus accumbens core and shell. J Neurosci, 31(33), 11786-11794. https://doi.org/10.1523/JNEUROSCI.2711-11.2011

Laurent, V., Bertran-Gonzalez, J., Chieng, B. C., & Balleine, B. W. (2014). δ-Opioid and Dopaminergic Processes in Accumbens Shell Modulate the Cholinergic Control of Predictive Learning and Choice. J Neurosci, 34(4), 1358-1369. https://doi.org/10.1523/JNEUROSCI.4592-13.2014

Laurent, V., Leung, B., Maidment, N., & Balleine, B. W. (2012). μ- and δ-opioid-related processes in the accumbens core and shell diMerentially mediate the influence of reward-guided and stimulus-guided decisions on choice. J Neurosci, 32(5), 1875-1883. https://doi.org/10.1523/JNEUROSCI.4688-11.2012

Matamales, M., McGovern, A. E., Mi, J. D., Mazzone, S. B., Balleine, B. W., & BertranGonzalez, J. (2020). Local D2- to D1-neuron transmodulation updates goal-directed learning in the striatum. Science, 367(6477), 549-555. https://doi.org/10.1126/science.aaz5751

Parkes, S. L., Bradfield, L. A., & Balleine, B. W. (2015). Interaction of insular cortex and ventral striatum mediates the eMect of incentive memory on choice between goaldirected actions. J Neurosci, 35(16), 6464-6471. https://doi.org/10.1523/JNEUROSCI.4153-14.2015

Pettibone, J. R., Yu, J. Y., Derman, R. C., Faust, T. W., Hughes, E. D., Filipiak, W. E., Saunders, T. L., Ferrario, C. R., & Berke, J. D. (2019). Knock-In Rat Lines with Cre Recombinase at the Dopamine D1 and Adenosine 2a Receptor Loci. eNeuro, 6(5). https://doi.org/10.1523/ENEURO.0163-19.2019

Willett, J. A., Will, T., Hauser, C. A., Dorris, D. M., Cao, J., & Meitzen, J. (2016). No Evidence for Sex DiMerences in the Electrophysiological Properties and Excitatory Synaptic Input onto Nucleus Accumbens Shell Medium Spiny Neurons. eNeuro, 3(1), ENEURO.0147-15.2016. https://doi.org/10.1523/ENEURO.0147-15.2016

Version 2

Abstract

The nucleus accumbens shell (NAc-S) and its projections to the ventral pallidum (VP) are thought to be critical for stimulus-based decisions. The NAc-S is predominantly composed of spiny projection neurons (SPNs) that express either the dopamine D1 (D1-SPNs) or the dopamine D2 receptor (D2-SPNs). Yet, the role of these two neuronal subpopulations and their inputs to the VP in stimulus-based decisions remains unknown. Here, we used optogenetics in female and male knock-in rats to selectively silence D1-or D2-SPNs and their projections to the VP at a time when the rats were required to use predictive stimuli to choose between two instrumental actions. Silencing either population of NAc-S SPNs disrupted choice. Silencing NAc-S D1-SPNs terminals in the VP also disrupted choice. However, choice was left intact by silencing NAc-S D2-SPNs terminals in the VP. Together, these findings provide novel insights into the cellular mechanisms and circuitry underlying stimulus-based decisions. We discuss how these insights are consistent with a recent model proposing that these decisions are controlled by an opioid-based memory system residing in the NAc-S.

Introduction

Our choices and decisions are often influenced by predictive signals available in the environment (Hollis, 1984). This influence is studied in the laboratory using the outcome-specific Pavlovian-instrumental transfer (PIT) task (Trapold & Overmier, 1972; Colwill & Rescorla, 1988; Holmes et al., 2010; Cartoni et al., 2016; Laurent & Balleine, 2021; Leung et al., 2024a), which comprises three stages. The first two stages are Pavlovian and instrumental conditioning, which can be administered in any order. In Pavlovian conditioning, two stimuli (S1 and S2; e.g., sounds or lights) are paired with two distinct and motivationally significant outcomes (O1 and O2; e.g., food outcomes). In instrumental conditioning, subjects are trained to perform two actions (A1 and A2), with each earning one of the two outcomes (i.e., A1→O1 and A2→O2). The final stage is the PIT test, which evaluates choice between the two actions in the presence (and absence) of the stimuli. Typically, each of the Pavlovian stimuli bias choice towards the action with which it shares an outcome: i.e., S1: A1>A2 and S2: A2>A1.

At a neural level, PIT has been found to rely on interactions between a number of structures, including cortical (Keistler et al., 2015; Bradfield et al., 2015; Bradfield et al., 2018; Tensaouti et al., 2025; Ostlund & Balleine, 2007; Lichtenberg et al., 2021; Sias et al., 2021; Lichtenberg et al., 2017), amygdala (Corbit & Balleine, 2005; Morse et al., 2020; Sias et al., 2024; Lichtenberg et al., 2017; Shiflett & Balleine, 2010; Derman et al., 2020; Prévost et al., 2012), basal ganglia (Corbit & Balleine, 2011; Corbit et al., 2001; Corbit & Janak, 2010; Morse et al., 2020; Laurent et al., 2012; Bertran-Gonzalez et al., 2013; Laurent et al., 2014; Corbit et al., 2016; Leung et al., 2024b; Leung & Balleine, 2015; Leung & Balleine, 2013; Laurent et al., 2015; Prévost et al., 2012), midbrain (Corbit et al., 2007; Sias et al., 2024; Seitz et al., 2022) and thalamic (Ostlund & Balleine, 2008; Leung et al., 2024b) territories. Among these, the nucleus accumbens shell (NAc-S) stands out because, unlike the others, it does not itself contribute to learning the stimulus-outcome (S-O) or the action-outcome (A-O) associations formed during the Pavlovian and instrumental stages of the PIT task and appears only to mediate their interaction (Corbit & Balleine, 2011; Corbit et al., 2001; Morse et al., 2020; Laurent et al., 2012; Laurent et al., 2014). Therefore, it has been proposed that the NAc-S acts as a central hub integrating S-O and A-O information at the time of PIT to guide choice between actions in an outcome-selective manner (Shiflett & Balleine, 2010; Bertran-Gonzalez & Laurent, 2018; Laurent & Balleine, 2021). Accordingly, PIT is disrupted by manipulations undermining NAc-S function during the PIT test (Morse et al., 2020; Corbit et al., 2016; Laurent et al., 2012; Laurent et al., 2014; Laurent et al., 2015).

The NAc-S is predominantly composed of spiny projection neurons (SPNs), which can be classified into two distinct subpopulations depending on the dopamine receptor they express (Gerfen & Surmeier, 2011). One population harbors the dopamine D1 receptors (D1Rs; D1-SPNs) whereas the other population express the dopamine D2 receptors (D2Rs; D2-SPNs). To date, only one study has examined whether activity in both populations is recruited during PIT (Laurent et al., 2014), finding that PIT is associated with an increase in ERK1/2 phosphorylation in D1-SPNs but not D2-SPNs. Further, blockade of D1Rs in NAc-S during choice between actions disrupted PIT whereas D2Rs blockade had no ekect. Although these findings indicate a major involvement of NAc-S D1-SPNs in PIT, caution must be exerted when drawing conclusions on the role of D2-SPNs. While D1Rs are exclusively expressed on NAc-S D1-SPNs, D2Rs are present on cholinergic interneurons (CINs) and local presynaptic dopamine terminals in addition to D2-SPNs (Gerfen & Surmeier, 2011). Additionally, research shows that ERK1/2 phosphorylation may not capture all transcriptional events that occur in D2-SPNs (Bertran-Gonzalez et al., 2008; Matamales et al., 2020). Lastly, pharmacological D2Rs blockade has been found to be quite inekective at influencing D2-SPNs function (Tozzi et al., 2007). Thus, the relative contribution of NAc-S D1-and D2-SPNs in driving choice between actions during PIT remains unclear.

The present experiments therefore aimed to provide an unambiguous assessment of the roles played by the two main populations of NAc-S SPNs during choice between actions in an outcome-specific PIT task. This assessment was achieved through optogenetic silencing in two knock-in rat lines that express Cre recombinase in either D1-or D2-SPNs (Pettibone et al., 2019). We first used tract-tracing and ex-vivo electrophysiology to confirm our capacity to selectively silence activity in NAc-S D1-or D2-SPNs. We then conducted two experiments that assessed outcome-specific PIT while D1-or D2-SPNs in the NAc-S were silenced during presentations of the predictive stimuli in the choice test. Finally, we assessed the degree to which the influence of the two SPNs populations on outcome-specific PIT depends on their projections to the ventral pallidum (VP). We focused on the VP for two main reasons: First, both NAc-S D1-and D2-SPNs send dense projections to this region (Lu et al., 1997; Kupchik et al., 2015); and, second, communication between the NAc-S and VP has previously been found to be critical for outcome-specific PIT (Leung & Balleine, 2013).

Results

Anterograde tracing and ex-vivo cell recordings in D1-Cre and A2a-Cre rats

We first examined the connectivity of striatal D1-and D2-SPNs in transgenic rats expressing Cre recombinase in neurons harboring the dopamine D1 receptor (D1-Cre rats) or the adenosine A2a receptor (A2a-Cre rats) (Pettibone et al., 2019). A2a receptors are predominantly expressed on striatal D2-SPNs (Schikmann et al., 2007), thereby enabling selective targeting of these SPNs in the A2a-Cre line. D1-Cre (n = 1 female and 1 male) and A2a-Cre rats (n = 1 female and 1 male) were unilaterally infused with a Cre-dependent eYFP virus in the NAc-S (Figure 1 A). To validate expression in these rat lines, other D1-Cre (n = 4 males) and A2a-Cre rats (n = 3 males) were given this unilaterally infusion in the dorsal striatum (DS) with the same virus. In both transgenic strains, viral expression was largely restricted to local SPNs in the two targeted regions: ∼95% of eYFP-positive neurons expressed DARPP-32 (D32 | eYFP; Figures 1 C-F). Viral expression was consistent with D1-and D2-SPNs representing two neuronal populations similar in size: in both transgenic strains and targeted regions ∼40% of DARPP-32-postive neurons expressed eYFP (eYFP | D32; Figures 1 C-F). DS infusion in D1-Cre rats resulted in eYFP-positive terminals (Figures 1 B-C) in the substantia nigra pars reticulata (SNr) and the globus pallidus externus (GPe). By contrast, the same infusion in A2a-Cre rats only produced eYFP-positive terminals in the GPe (Figures 1 B and D). NAc-S infusion in both D1-Cre and A2a-Cre rats resulted in eYFP-positive terminals in the VP (Figures 1 E-F). By contrast, eYFP-positive terminals in the lateral hypothalamus (Figures 1 E-F) were only observed in D1-Cre rats. Overall, these results are in line with current knowledge about the cellular composition of the striatum (Gerfen & Surmeier, 2011; Bertran-Gonzalez et al., 2010) and well-established projection patterns of striatal D1-and D2-SPNs (Gerfen & Surmeier, 2011; Lu et al., 1997; Kupchik et al., 2015; O’Connor et al., 2015). We were therefore confident in the capacity of the two rat transgenic lines to reveal the function of NAc-S D1-and D2-SPNs.

Anterograde tracing in D1-Cre and A2a-Cre rats.

A. D1-Cre or A2a-Cre rats were unilaterally infused in the dorsal striatum (DS; D1-Cre: n = 4 males; A2A-Cre: n = 3 males) or nucleus accumbens shell (NAc-S; D1-Cre: n = 1 female and 1 male; A2A-Cre: n = 3 males) with DIO-eYFP. B. Sagittal micrographs obtained in D1-Cre (top) and A2A-Cre (bottom) rats following viral infusion in the DS. C. DIO-eYFP infusion in the DS of D1-Cre rats. Micrographs show eYFP expression in the DS, D32 staining and co-labelling (eYFP + D32) in the DS. They also show that DS D1-SPNs project to the substantia nigra pars reticulata (SNr) and the globus pallidus externus (GPe). Viral expression was restricted to putative SPNs (D32 | eYFP), with ∼40% of SPNs expressing eYFP (eYFP | D32). D. DIO-eYFP infusion in the DS of A2a-Cre rats. Micrographs show eYFP expression in the DS, D32 staining and co-labelling (eYFP + D32) in the DS. They also show that DS D2-SPNs project to the SNr but not the GPe. Viral expression was restricted to putative SPNs (D32 | eYFP), with ∼40% of SPNs expressing eYFP (eYFP | D32). E. DIO-eYFP infusion in the NAc-S of D1-Cre rats. Micrographs show eYFP expression in the NAc-S, D32 staining and co-labelling (eYFP + D32) in the DS. They also show that NAc-S D1-SPNs project to the ventral pallidum (VP) and the lateral hypothalamus (LH). Viral expression was restricted to putative SPNs (D32 | eYFP), with ∼41% of SPNs expressing eYFP (eYFP | D32). F. DIO-eYFP infusion in the NAc-S of A2a-Cre rats. Micrographs show eYFP expression in the NAc-S, D32 staining and co-labelling (eYFP + D32) in the DS. They also show that NAc-S D2-SPNs project to the VP but not the LH. Viral expression was restricted to putative SPNs (D32 | eYFP), with ∼40% of SPNs expressing eYFP (eYFP | D32).

Next, we used ex-vivo electrophysiological recordings to assess our capacity to silence activity of NAc-S D1-and D2-SPNs. D1-Cre and A2a-Cre rats were bilaterally infused in the NAc-S with either a null Cre-dependent eYFP virus (D1-Cre: 2 females; A2a-Cre: 2 females) or a Cre-dependent inhibitory halorhodospin (eNpHR3.0; D1-Cre: 2 females; A2a-Cre: 2 females) virus (Figures 2 A-B). LED light illumination (625 nm, continuous) of either NAc-S D1-SPNs (Figure 2 A) or NAc-S D2-SPNs (Figure 2 B) transfected with the eYFP virus had no ekect on action potential firing (Baseline vs. LED ON; D1-Cre: p = 1, 5 cells; A2A-Cre: p = 0.35, 8 cells). By contrast, the same illumination onto NAc-S D1-(Figure 2 A) or D2-SPNs (Figure 2 B) transfected with eNpHR3.0 suppressed action potential firing (D1-Cre: F(1,6) = 16.82; η2 = 0.74, p < 0.001, 7 cells; A2A-Cre: F(1,5) = 135.00; η2 = 0.96, p < 0.001, 6 cells). Importantly, there was no dikerence in firing prior to or following LED light illumination, indicating that the optical manipulation did not alter D1-or D2-SPNs physiology. Together, these results validate the ekicacy of eNpHR3.0 to specifically silence D1-and D2-SPNs activity at the time of LED light delivery.

Ex-vivo cell recordings in D1-Cre and A2a-Cre rats.

A. D1-Cre were bilaterally infused in the NAc-S with DIO-eYFP (black; 2 females) or DIO-eNpHR3.0 (blue; eNpHR3.0; 2 females). The representative raw traces of cell attached recordings are those of transfected neurons that were depolarized to elicit action potentials by injecting a brief positive current step (+150pA, 200ms duration, 0.5Hz). 625nm LED illumination (orange bar, continuous wave, 2mW) had no ekect in eYFP transfected neurons (black; 5 cells) but it inhibited action potential in eNpHR3.0 transfected neurons (blue; 7 cells). The grouped data for recordings include overlapping data points. B. A2a-Cre were bilaterally infused in the NAc-S with DIO-eYFP (black; 2 females) or DIO-eNpHR3.0 (red; 2 females). The representative raw traces of cell attached recordings are those of transfected neurons that were depolarized to elicit action potentials by injecting a brief positive current step (+150pA, 200ms duration, 0.5Hz). 625nm LED illumination (orange bar, continuous wave, 2mW) had no ekect in eYFP transfected neurons (black; 8 cells) but it inhibited action potential in eNpHR3.0 transfected neurons (red; 6 cells). The grouped data for recordings include overlapping data points.

NAc-S D1-SPNs mediate outcome-specific PIT

This experiment examined the ekects of silencing NAc-S D1-SPNs on outcome-specific PIT. D1-Cre rats were bilaterally infused in the NAc-S with either the null Cre-dependent eYFP virus (eYFP: 5 females and 5 males) or the Cre-dependent halorhodopsin (eNpHR3.0: 5 females and 5 males) virus and were implanted with fiber-optic cannulas above the NAc-S (Figure 3 A and Supplementary Figures S1 A-B). Using the outcome-specific PIT protocol depicted in Figure 3 B, the rats first received Pavlovian conditioning during which two stimuli (S1 and S2; white noise or clicker) were paired with two food outcomes (O1 and O2; grain pellets or sucrose solution). Next, the rats underwent instrumental conditioning and learned that pressing one lever earned one of the food outcomes whereas pressing a second lever delivered the other outcome. Finally, an outcome-specific PIT test was administered and assessed choice between the two lever press actions in the presence or absence of each stimulus. Critically, half of the trials for each stimulus took place under LED light stimulation (625 nm, continuous; ON trials) whereas the other half of the trials was conducted without the stimulation (OFF trials). Thus, NAc-S D1-SPNs were silenced on half of the trials in rats infused with eNpHR3.0.

NAc-S D1-SPNs mediate outcome-specific PIT.

A. D1-Cre rats were bilaterally infused in the NAc-S with DIO-eYFP (black; 5 females and 5 males) or DIO-eNpHR3.0 (blue; 5 females and 5 males). Fiber-optic cannulas were implanted above the NAc-S to provide 625nm LED illumination (continuous). B. Schematic representation of the behavioral design; S1 and S2: noise and clicker stimuli (counterbalanced); O1 and O2: grain pellets and sucrose solution (counterbalanced); A1 and A2: left and right lever press (counterbalanced). At test, S1 and S2 were presented 4 times each, in a pseudorandom order. Half of the trials for each stimulus was conducted under 625nm LED illumination (ON; continuous wave; ∼10mW) whereas the LED remained inactivated during the other half of the trials (OFF). ON/OFF trials were counterbalanced. C. Outcome-specific PIT test: net lever presses when the stimuli predicted the same outcome as the action (Same) or when the stimuli predicted the dikerent outcome (Dikerent). Lever presses are shown for each group in trials conducted under 625nm LED illumination (ON) and in trials without illumination (OFF). Data are shown as mean ± SEM. Panel C includes individual data points for female (filled circle) and male (open circle) rats. Asterisks denote significant ekect (* p < 0.05; ** p < 0.01; *** p < 0.001; n.s., nonsignificant).

Pavlovian and instrumental conditioning proceeded as expected (Supplementary Figures S1 C-D). The data of most interest are those from the outcome-specific PIT test and are presented in Figure 3 C (see also Supplementary Figure S1 E-F). They are plotted as the mean number of lever presses per minute when the stimulus predicted the same outcome as the action (Same) and when the stimulus predicted a dikerent outcome to the action (Dikerent). Thus, A1 was identified as “Same” and A2 as “Dikerent” in the presence of S1. Conversely, A2 was identified as “Same” and A1 as “Dikerent” in the presence of S2. Further, baseline responding (number of lever presses per minute on the two actions during the two minutes preceding each stimulus presentation) was subtracted from the rates of responding during the stimuli since it did not diker between groups (Group – eYFP vs. eNpHR3.0: p = 0.42). This approach allowed us to focus on the net ekect of the stimuli on choice performance.

Silencing NAc-S D1 SPNs eliminated outcome-specific PIT. Overall lever press rates were similar between groups (Group: p = 0.48), LED light condition (LED – ON vs OFF: p = 0.14), and the two factors did not interact (Group x Light: p = 0.07). However, lever press rates were higher on the action earning the same outcome as the stimuli compared to the action earning the dikerent outcome (Lever – Same vs. Dikerent: F(1,18) = 50.06; η2 = 0.74; p < 0.001), regardless of group (Group x Lever: p = 0.77). There was no Lever by LED light condition interaction (Lever x LED: p = 0.72) but critically, there was an interaction between Group, LED light condition, and Lever during the presentation of the predictive stimuli (Group x LED x Lever: F(1,18) = 7.19; η2 = 0.29; p = 0.02). Follow-up analyses revealed that rats in the eYFP group displayed intact outcome-specific PIT whether the LED light was OFF (F(1,9) = 6.14; η2 = 0.67; p < 0.05) or ON (F(1,9) = 39.46; η2 = 0.81; p < 0.001). By contrast, rats in the eNpHR3.0 group displayed outcome-specific PIT when the light was OFF (F(1,9) = 12.38; η2 = 0.58; p < 0.01) but not when it was ON (p = 0.67). Together, these results indicate that activity in NAc-S D1-SPNs is necessary for the outcome-specific influence exerted by predictive stimuli on choice between actions. This is consistent with previous pharmacological work showing that NAc-S D1Rs blockade removes specific PIT (Laurent et al., 2014). Importantly, the impairment produced by silencing NAc-S D1-SPNs was restricted to the influence of predictive stimuli on choice between actions. In the same rats, we found that this silencing had no ekect on valued-based choice (Supplemental Figure S1 G) using an outcome devaluation procedure: all rats performed the action that previously earned a valuable outcome more than the action that earned a devalued outcome. This agrees with studies reporting that NAc-S lesion or inactivation spares value-based choice (Corbit & Balleine, 2011; Corbit et al., 2001; Morse et al., 2020; Laurent et al., 2012; Laurent et al., 2014).

NAc-S D2-SPNs mediate outcome-specific PIT

The removal of outcome-specific PIT under silencing of NAc-S D1-SPNs reproduced the impairment previously observed with blockade of NAc-S D1Rs (Laurent et al., 2014). Interestingly, the latter pharmacological study did not find any ekect following NAc-S infusion of a D2Rs antagonist. To evaluate this finding, the present experiment examined the ekects of silencing NAc-S D2-SPNs on outcome-specific PIT. A2a-Cre rats were bilaterally infused in the NAc-S with either the null Cre-dependent eYFP virus (eYFP: 4 females and 4 males) or the Cre-dependent halorhodospin (eNpHR3.0: 4 females and 2 males) virus and were implanted with fiber-optic cannulas above the NAc-S (Figure 4 A and Supplementary Figures S2 A-B). The rats then received the outcome-specific PIT protocol previously described (Figure 3 B).

NAc-S D2-SPNs mediate for outcome-specific PIT.

A. A2a-Cre rats were bilaterally infused in the NAc-S with DIO-eYFP (black; 4 females and 4 males) or DIO-eNpHR3.0 (red; 4 females and 2 males). Fiber-optic cannulas were implanted above the NAc-S to provide 625nm LED illumination (continuous). B. Outcome-specific PIT test: net lever presses when the stimuli predicted the same outcome as the action (Same) or when the stimuli predicted the dikerent outcome (Dikerent). Lever presses are shown for each group in trials conducted under 625nm LED illumination (ON) and in trials without illumination (OFF). Data are shown as mean ± SEM. Panel B includes individual data points for female (filled circle) and male (open circle) rats. Asterisks denote significant ekect (* p < 0.05; ** p < 0.01; n.s., nonsignificant).

Pavlovian and instrumental conditioning went as expected (Supplementary Figures S2 C-D). The data from the outcome-specific PIT test are presented in Figure 4 B, again with baseline subtracted since it did not diker between groups (Group: p = 0.89; see also Supplemental Figure 2 E-F). Silencing NAc-S D2-SPNs activity eliminated outcome-specific PIT. Overall lever press rates during the test were similar between groups (Group: p = 0.86), LED light condition (LED: p = 0.21), and the two factors did not interact (Group x LED: p = 0.70). Lever press rates were higher on the action earning the same outcome as the stimuli compared to the action earning the dikerent outcome (Lever: F(1,12)= 22.77; η2 = 0.66; p < 0.001), regardless of group (Group x Lever: p = 0.46). There was no Lever by LED light condition interaction (Lever x LED: p = 0.15) but critically, there was an interaction between group, LED light condition, and Lever during the presentation of the predictive stimuli (Group x LED x Lever: F(1,12) = 8.73; η2 = 0.42; p < 0.01). Follow-up analyses revealed that control eYFP rats expressed outcome-specific PIT whether the LED light was OFF (F(1,7) = 16.25; η2 = 0.70; p < 0.01) or ON (F(1,7) = 20.58; η2 = 0.75; p < 0.01).

By contrast, rats in the eNpHR3.0 group displayed outcome-specific PIT when the light was OFF (F(1,5) = 6.94; η2 = 0.58; p < 0.05) but not ON (p = 0.34). Thus, these results show for the first time that activity in NAc-S D2-SPNs is necessary for the outcome-specific influence exerted by predictive stimuli on choice between actions.

The present finding contrasts with our previous observation that outcome-specific PIT remains unakected by NAc-S infusion of a D2Rs antagonist (Laurent et al., 2014). However, the capacity of such pharmacological manipulation to reveal the function of D2-SPNs is unclear. For instance, a D2Rs antagonist could be expected to enhance D2-SPN activity, since D2Rs are Gi coupled and so their activation should reduce adenylyl cyclase activity. Further, D2Rs are not exclusively expressed on this cell population (Gerfen & Surmeier, 2011), as they can be found on local CINs and presynaptic dopamine terminals. Finally, previous work found that pharmacological treatment targeting D2Rs can leave D2-SPNs activity unakected (Tozzi et al., 2007). These issues are overcome by the present optical manipulation in A2a-Cre rats, and we are therefore confident that NAc-S D2-SPNs activity is in fact critical for outcome-specific PIT. Importantly, we also found that the impairment produced by silencing NAc-S D2-SPNs was restricted to the influence of predictive stimuli on choice between actions. In the same rats, this silencing had no ekect on valued-based choice (Figure S2 G).

NAc-S D1-SPNs projections to the VP mediate outcome-specific PIT

Previous work indicates that communication between the NAc-S and VP is critical for outcome-specific PIT (Leung & Balleine, 2013). Consistent with the literature (Lu et al., 1997; Kupchik et al., 2015), we observed that both NAc-S D1-and D2-SPNs densely innervate the VP (Figure 1 E-F). Since we found that activity in both neuron populations mediates the outcome-specific PIT ekect, we investigated whether projections from each population contribute to the ekect. We first focused on those originating from NAc-S D1-SPNs. D1-Cre rats were bilaterally infused in the NAc-S with either the null Cre-dependent eYFP virus (eYFP: 3 females and 5 males) or the Cre-dependent halorhodopsin (eNpHR3.0: 6 females and 6 males) virus and were implanted with fiber-optic cannulas above the VP (Figure 5 A and Supplementary Figures S3 A-B). The rats then received the behavioral protocol previously described (Figure 3 B).

NAc-S D1-SPNs projections to the VP mediate outcome-specific PIT.

A. D1-Cre rats were bilaterally infused in the NAc-S with DIO-eYFP (black; 3 females and 5 males) or DIO-eNpHR3.0 (blue; 6 females and 6 males). Fiber-optic cannulas were implanted above the VP to provide 625nm LED illumination (continuous). B. Outcome-specific PIT test: net lever presses when the stimuli predicted the same outcome as the action (Same) or when the stimuli predicted the dikerent outcome (Dikerent). Lever presses are shown for each group in trials conducted under 625nm LED illumination (ON) and in trials without illumination (OFF). Data are shown as mean ± SEM. Panel B includes individual data points for female (filled circle) and male (open circle) rats. Asterisks denote significant ekect (** p < 0.01; *** p < 0.01; n.s., nonsignificant).

Pavlovian and instrumental conditioning went as expected (Supplementary Figures S3 C-D). The data from the outcome-specific PIT test are presented in Figure 5 B in the manner described previously since baseline responding did not diker between groups (Group: p = 0.34; see also Supplemental Figures 3 E-F). Silencing NAc-S D1-SPNs projections to the VP eliminated outcome-specific PIT. Overall lever press rates during the test were similar between groups (Group: p = 0.11). LED activation reduced these rates (LED: F(1,18)= 7.58; η2 = 0.30; p < 0.05) but this reduction depended on the group considered (Group x LED: F(1,18)= 9.78; η2 = 0.35; p < 0.01). Lever press rates were higher on the action earning the same outcome as the stimuli compared to the action earning the dikerent outcome (Lever: F(1,18)= 35.73; η2 = 0.67; p < 0.001), regardless of group (Group x Lever: p = 0.19). There was a Lever by LED light condition interaction (Lever x LED: F(1,18)= 7.56; η2 = 0.29; p < 0.05) and critically, there was an interaction between group, LED light condition, and Lever during the presentation of the predictive stimuli (Group x LED x Lever: F(1,18) = 5.01; η2 = 0.22; p < 0.05). Follow-up analyses revealed that control eYFP rats expressed outcome-specific PIT whether the LED light was OFF (F(1,7) = 26.63; η2 = 0.79; p < 0.001) or ON (F(1,7) = 12.31; η2 = 0.64; p < 0.01). By contrast, rats in the eNpHR3.0 group displayed outcome-specific PIT when the light was OFF (F(1,11) = 26.53; η2 = 0.71; p < 0.001) but not ON (p = 0.48). Thus, these results show for the first time that NAc-S D1-SPNs mediate outcome-specific PIT via their projections to the VP. Importantly, we also found that the impairment produced by silencing NAc-S D1-SPNs terminals in the VP was restricted to the influence of predictive stimuli on choice between actions. In the same rats, this silencing had no ekect on valued-based choice (Figure S3 G).

NAc-S D2-SPNs projections to the VP do not mediate outcome-specific PIT

Since activity in NAc-S D2-SPNs is required for outcome-specific PIT (Figure 3) and these neurons innervate the VP (Figure 1 F), we assessed whether D2-SPNs projections to the VP are involved in outcome-specific PIT. A2a-Cre rats were bilaterally infused in the NAc-S with either the null Cre-dependent eYFP virus (eYFP: 4 females and 5 males) or the Cre-dependent halorhodopsin (eNpHR3.0: 5 females and 6 males) virus and were implanted with fiber-optic cannulas above the VP (Figure 6 A and Supplementary Figures S4 A-B). The rats then received the behavioral protocol previously described (Figure 3 B).

NAc-S D2-SPNs projections to the VP do not mediate outcome-specific PIT.

A. A2a-Cre rats were bilaterally infused in the NAc-S with DIO-eYFP (black; 4 females and 5 males) or DIO-eNpHR3.0 (red; 5 females and 6 males). Fiber-optic cannulas were implanted above the VP to provide 625nm LED illumination (continuous). B. Outcome-specific PIT test: net lever presses when the stimuli predicted the same outcome as the action (Same) or when the stimuli predicted the dikerent outcome (Dikerent). Lever presses are shown for each group in trials conducted under 625nm LED illumination (ON) and in trials without illumination (OFF). Data are shown as mean ± SEM. Panel B includes individual data points for female (filled circle) and male (open circle) rats.

Pavlovian and instrumental conditioning went as expected (Supplementary Figures S4 C-D). The data from the outcome-specific PIT test are presented in Figure 6 B as previously since baseline responding did not diker between groups (Group: p = 0.90; see also Supplemental Figures 4 E-F). Silencing NAc-S D2-SPNs projections to the VP had no ekect on outcome-specific PIT. Overall lever press rates were similar between groups (Group: p = 0.56), LED light condition (LED: p = 0.75), and the two factors did not interact (Group x LED: p = 0.73). The rates were higher on the action earning the same outcome as the stimuli relative to the action earning the dikerent outcome (Lever: F(1,13) = 22.71; η2 = 0.64; p < 0.001), irrespective of group (Group x Lever: p = 0.55). There was no interaction between groups, LED light condition, and Lever (Group x Light x Lever: p = 0.59). Thus, NAc-S D2-SPNs do not appear to mediate outcome-specific PIT via their projections to the VP. Likewise, we found no evidence that these projections influence valued-based choice (Figure S4 G).

Discussion

The present experiments investigated the role of NAc-S D1-and D2-SPNs in choice between actions using an outcome-specific PIT task. First, they combined anatomical tract-tracing and ex-vivo electrophysiology to demonstrate that two recently developed knock-in rat lines (Pettibone et al., 2019) enable selective silencing of activity in either population of NAc-S SPNs. Consistent with previous findings (Laurent et al., 2014), they found that NAc-S D1-SPNs are necessary for PIT since their silencing eliminated outcome-specific choice. Additionally, they showed that this choice is also abolished by silencing NAc-S D2-SPNs, providing the first evidence that both SPNs populations contribute to PIT expression. Finally, the last two experiments revealed for the first time that outcome-specific choice is likely to involve downstream regulation of VP function by NAc-S D1-SPNs but not NAc-S D2-SPNs. Together, these findings oker novel insights into the cellular mechanisms that govern the outcome-specific influence of predictive stimuli on choice between actions.

Convincing evidence had been provided for a significant role of NAc-S D1-SPNs in PIT (Laurent et al., 2014). The evidence was based on the observations that PIT is associated with an increase in ERK1/2 phosphorylation within these SPNs and that pharmacological blockade of NAc-S D1Rs eliminates outcome-specific choice between actions. These findings align with the present study, showing that PIT is eliminated when NAc-S D1-SPNs are silenced during presentations of the predictive stimuli at the time of choice. Importantly, the ekect of NAc-S D1-SPNs silencing was specific to the assessment of choice between actions in the presence of predictive stimuli. For instance, the ability of stimuli to elicit approach behaviors toward the magazine remained intact despite NAc-S D1-SPNs silencing, indicating that the silencing ekect was not mediated by modulating potential competition between Pavlovian and instrumental responses (Lovibond, 1981; Holmes et al., 2010). Furthermore, NAc-S D1-SPNs silencing preserved the capacity to select between actions based on the value of their respective outcomes. This selectivity in the impairment produced by NAc-S D1-SPNs is consistent with previous findings demonstrating that the NAc-S does not contribute to learning and retrieving the S-O and A-O associations produced by Pavlovian and instrumental conditioning (Corbit & Balleine, 2011; Corbit et al., 2001; Morse et al., 2020; Laurent et al., 2012; Laurent et al., 2014).

Previous research failed to provide any evidence supporting the involvement of NAc-S D2-SPNs in outcome-specific choice (Laurent et al., 2014). Specifically, PIT was found to produce no change in ERK1/2 phosphorylation in these neurons and was left intact by NAc-S D2Rs blockade. However, these assessments have significant limitations, including the widespread distribution of striatal D2Rs (Gerfen & Surmeier, 2011) and the inability of ERK1/2 phosphorylation or D2Rs antagonism to capture or distort activity in NAc-S D2-SPNs (Bertran-Gonzalez et al., 2008; Tozzi et al., 2007). Further, pharmacological blockade of D2Rs would be expected to enhance D2-SPNs activity, since D2Rs are Gi coupled and so their activation reduces adenylyl cyclase activity. The present study addressed these limitations by implementing optogenetic silencing in a knock-in rat line giving access to D2-SPNs through the targeting of the A2a receptor that is predominantly expressed in these neurons in the NAc-S (Schikmann et al., 2007). We found that PIT is eliminated when NAc-S D2-SPNs are silenced during presentations of the predictive stimuli at the time of choice. Moreover, the same silencing left intact the ability of the predictive stimuli to elicit magazine approaches and preserved the capacity to choose between actions based on the value of their outcomes. Thus, we conclude that both NAc-S D1-and D2-SPNs are indispensable to outcome-specific PIT. Nevertheless, it will be important for future studies to confirm NAc-S D2-SPNs involvement in PIT using alternative approaches, such as chemogenetic or immunohistochemical assessments employing a marker capable of capturing the function of this neuronal population (Matamales et al., 2020).

Consistent with the literature (Lu et al., 1997; Kupchik et al., 2015), we found that both NAc-S D1-and D2-SPNs send dense projections to the VP, which is often described as the major ekerent of the nucleus accumbens (Heimer et al., 1991; Zahm & Heimer, 1990). Importantly, the VP has been shown to play a crucial role in outcome-specific choice (Leung & Balleine, 2013; Leung & Balleine, 2015; Leung et al., 2024b). VP neurons exhibit enhanced cFos activation following an outcome-specific PIT test, and their levels of activation correlate with PIT performance. Moreover, pharmacological inactivation of the VP eliminates outcome-specific choice, and the same elimination is observed when the NAc-S is disconnected from the VP. We therefore tested for a role of NAc-S D1-and D2-SPNs projections to the VP during choice between actions in our outcome-specific PIT task. We found that PIT was abolished by silencing NAc-S D1-SPNs terminals but not by silencing NAc-S D2-SPNs terminals. Neither silencing akected magazine approaches elicited by the predictive stimuli or choice between actions based on the value of their outcomes. It is important to note that our study does not provide any evidence about the ekicacy of NAc-S D2-SPNs terminals silencing in the VP, and future experiments should aim to provide such evidence or adopt other methods to study this pathway. This could involve using opsins with enhanced axonal silencing ekicacy (Mahn et al., 2021; Copits et al., 2021), or employing alternative methods known to disrupt neurotransmitter release such as chemogenetics (Rost et al., 2022). Yet, the same silencing for NAc-S D1-SPNs terminals resulted in PIT elimination. Therefore, it seems reasonable to conclude that NAc-S D1-SPNs, but not NAc-S D2 SPNs, projections to the VP are indispensable to observe outcome-specific choice in a PIT task. Since NAc-S D2-SPNs appear to exclusively project to the VP (Humphries & Prescott, 2010; Kupchik et al., 2015), our findings suggest that these SPNs mediate outcome-specific PIT by locally regulating NAc-S function. By contrast, it remains to be determined whether NAc-S D1-SPNs also coordinate outcome-specific PIT via their projections to the lateral hypothalamus (O’Connor et al., 2015) and/or the ventral tegmental area (VTA) (Humphries & Prescott, 2010), in addition to the VP.

The present findings are consistent with a recent model proposing that outcome-specific choice in PIT relies on an opioid-based memory residing in the NAc-S (for full description and illustration of the model see Leung et al., 2024a; Laurent & Balleine, 2021; Morse et al., 2020). In this model, as the basolateral amygdala encodes and stores outcome-specific S-O associations across Pavlovian conditioning, it also drives the formation of the NAc-S memory, which involves the durable accumulation of delta-opioid receptors (DOPRs) on the somatic membrane of local cholinergic interneurons (CINs) (Bertran-Gonzalez et al., 2013; Morse et al., 2020). Although this memory is not necessary for Pavlovian conditioning per se, its expression is later required to enable the predictive stimuli to guide outcome-specific choice during the PIT test (Morse et al., 2020). The model specifically proposes that memory expression is controlled by fluctuations in glutamatergic release from cortical inputs and dopamine release from projections originating from the VTA, as this brain region has been found to be important for outcome-specific PIT (Corbit et al., 2007; Sias et al., 2024; Seitz et al., 2022; Leung & Balleine, 2015). One consequence is to activate NAc-S D2-SPNs and enkephalin discharge by these neurons. Enkephalin then binds onto DOPRs that had accumulated on local CINs and dampens acetylcholine secretion from the interneurons. The sudden drop in NAc-S acetylcholine frees D1-SPNs from the inhibitory tone imposed by acetylcholine occupancy of muscarinic M4 receptors that are exclusively found on these SPNs (Tayebati et al., 2004; Lobo et al., 2006; Guo et al., 2010; Jeon et al., 2010). The ultimate consequence is to promote NAc-S D1-SPNs function, including the coordination of outcome-specific choice in PIT by regulating activity in downstream brain regions such as the VP. The model predicts all the findings presented here. NAc-S D2-SPNs silencing eliminates PIT at the time of choice by preventing enkephalin release and thereby, expression of the DOPR-based memory. The main consequence of this expression is precluded by NAc-S D1-SPNs silencing at the time of choice, while silencing the terminals of these SPNs in the VP squash their ability to regulate the activity of this brain region to coordinate outcome-specific choice in PIT. Thus, one fundamental implication of the present findings is to strengthen the proposal that an opioid-based memory system in the NAc-S enables outcome-specific choice between actions.

In summary, the present experiments found that the two main populations of SPNs in the NAc-S are indispensable for outcome-specific choice between actions in a PIT task. They also revealed that NAc-S D1-SPNs coordinate this choice by downstream regulation of VP activity. By contrast, NAc-S D2-SPNs function in PIT appears to be restricted to modulating local activity. These findings provide novel insights in the cellular mechanisms and circuitry underlying the outcome-specific influence of predictive stimuli on choice between actions, and are consistent with a recent model proposing that this influence is mediated by an opioid-based memory system in the NAc-S. Beyond these mechanistic considerations, the findings oker an opportunity to gain novel insights about various disorders in which the outcome-specific influence of predictive stimuli over our choices and decisions is dysfunctional. These disorders include depression (Geurts et al., 2013; Nord et al., 2018), anxiety disorders (Quail et al., 2017; Krypotos & Engelhard, 2020), substance-use disorders (Heinz et al., 2019; Hogarth et al., 2018; Hogarth et al., 2018; Steins-Loeber et al., 2020; Garbusow et al., 2016; Garbusow et al., 2019; Hogarth et al., 2018), gambling disorders (Genauck et al., 2019), anorexia nervosa (Vogel et al., 2019) and obesity (Watson et al., 2014; Lehner et al., 2017; Meemken & Horstmann, 2019).

Materials and methods

Subjects

Two genetically modified knock-in rat lines were used and obtained from the breeding facility at the University of New South Wales (Sydney, Australia). D1-Cre rats (Rat Research & Resource Center, Columbia, MO, USA; LE-Drd1em1(iCre)Berke, RRRC#: 00856) expressed Cre recombinase in neurons expressing the dopamine D1 receptors (D1R). A2a-Cre rats (Rat Research & Resource Center, Columbia, MO, USA; LE-Adora2aem1(iCre)Berke, RRRC#: 00857) expressed Cre recombinase in neurons expressing the adenosine A2A receptors (A2R). All rats were heterozygous and generated by crossing a heterozygous male with a Long-Evans wild-type female obtained from a colony maintained by the University of New South Wales (founding animals were sourced from Envigo/Inotiv, Blue Spruce outbred; HsdBlu:LE). Genotyping was completed by sending ear clippings to Transnetyx (Cordova, TN, USA). The rats were at least 8 weeks old at the beginning of each experiment. Ekorts were made to allocate an equal number of female and male rats in each group. However, these ekorts were hampered by exclusions following post-mortem assessments of viral spread and fiber-optic cannulas placement. The final number of female and male rats did not provide sukicient power to analyze an ekect of sex (our previous work found no influence of sex on PIT or value-based choice; Burton et al., 2024). Therefore, all test data present individual performance for female and male rats. Rats were housed in transparent plastic boxes with their littermates (up to four rats per box) throughout, and in a climate-controlled colony room maintained on a 12 h light-dark cycle (lights on at 7:00am). Behavioral procedures were conducted during the light phase (8:00am to 6:00pm). Water and standard lab chow were available ad libitum prior to the start of each experiment. Rats were food restricted to maintain them at ∼90% of their ad libitum body weight during the behavioral protocols. Food restriction was initiated 3-5 days prior to the start of the protocols and was maintained throughout all training and testing phases. Rats received a daily food amount, which was adjusted based on body weight measurements recorded every two days. The Animal Care and Ethics Committee at the University of New South Wales approved all procedures.

Experimental timeline

Rats underwent initial stereotaxic surgery for viral infusion, followed by 7-11 days recovery. After recovery, rats received 8 days of Pavlovian conditioning and 8 days of instrumental conditioning, except in the VP experiments (NAc-S D1-and D2-SPNs→VP projection experiment) where only 7 days of instrumental conditioning were given. Following a 3-4 day interval, rats received a second surgery to implant fiber-optic cannulas and 3-4 days additional recovery. Rats then completed reminder sessions: one Pavlovian and two instrumental conditioning sessions, except in the last experiment (NAc-S D2-SPNs→VP projection experiment) where three instrumental sessions were given. Testing began the following day with the PIT test, followed by two instrumental conditioning reminder sessions over the next two days. Finally, rats underwent outcome devaluation using sensory-specific satiety and choice tests over two consecutive days.

Viruses

The following viruses were used: AAV5-EF1a-DIO-eYFP (eYFP; AddGene; Watertown, MA, USA) and AAV5-EF1a-DIO-eNpHR3.0-eYFP (eNpHR3.0; UNC Vector Core; Chapel Hill, NC, USA).

Surgeries

Stereotaxic surgery was conducted under isoflurane gas anesthesia (0.8L/min; induction at 4-5%, maintenance 2-2.5%). Animals were placed in a stereotaxic frame (Kopf Instruments, Tujunga, CA, USA), the surgical area was shaved, and betadine iodine antiseptic solution was applied. At the incision site, bupivacaine hydrochloride (0.5%; 0.1-0.2mL) was injected subcutaneously. In the lower flank, Metacam (1mg/kg) was injected subcutaneously. An incision was made along the midline to expose the skull, the membrane on top of the skull was cleared, and the skull was adjusted to align bregma and lambda on the same horizontal plane.

For viral infusion, small holes were drilled in the skull above the DS or NAc-S. The following coordinates (indicated in mm from bregma) were used for the DS:-0.45 anteroposterior (A/P); ±2.6 (M/L);-4.85 dorsoventral (D/V). For the NAc-S, the coordinates were: +1.8 or +1.9 A/P; ±0.85 or ±0.95 mediolateral M/L;-7.7 or-7.85 D/V. An Infuse/Withdraw Pump (Standard Infuse/Withdraw Pump 11 Elite; Holliston, Harvard Apparatus, MA, USA) in combination with glass 1µL syringes (86200; Hamilton Company, Reno, NV, USA) were used to perform infusions. 0.5µL of the virus was infused in each hemisphere at a rate of 0.1µL/min. Following each infusion, the needle was left in position for 5-10 min to allow for dikusion. The incision site was closed and secured with surgical sutures (684G; Ethicon, North Ryde, NSW, Australia) or surgical staples (59027; Stoelting, Wood Dale, IL, USA), and betadine antiseptic ointment was applied to the area. Animals then received subcutaneous injections of the antibiotic Duplocillin (0.15ml/kg) and sodium chloride (0.9%, 2mL) and placed on a heat mat for recovery. Once alert and responsive, animals were returned to their home cages and monitored daily as they recovered for a minimum of one week following surgery with ad libitum access to food and water.

Rats received a second surgery to bilaterally implant fiber-optic cannulas (10mm length, 0.48mm diameter; Doric Lenses, Québec, QC, Canada) after completion of the initial 15-16 days of training (Pavlovian and Instrumental). The surgical area was prepared using the same method as described for viral surgeries. Three jewellers’ screws were inserted into the skull, distributed anterior and posterior to the implantation site and secured in place with dental cement. Fiber-optic cannulas were implanted into the targeted brain region. The following coordinates (indicated in mm from bregma) were used for the NAc-S: +1.8 or +1.9 A/P;-0.85 or-0.95 M/L for left hemisphere; +2.2 M/L with-10° angle for right hemisphere;-6.75 D/V for left hemisphere;-6.8 or-6.9 D/V for right hemisphere. The following coordinates (indicated in mm from bregma) were used for the VP: +0.24 A/P; ±1.8 M/L;-7.5 or-8.0 D/V. Post-implantation, rats received the same care as described for viral surgeries and were allowed to recover for 3-4 days before resuming the behavioral procedures.

Behavioral apparatus

Training and testing were conducted in 12 identical operant chambers (ENV-007-VP; L 29.53 x W 23.5 x H 27.31 cm; MED Associates, Fairfax, VT, USA) enclosed within light-and sound-resistant cabinets (ENV-018MD; L 59.69 x W 40.64 x H 55.88 cm; MED Associates, Fairfax, VT, USA). The chambers consisted of two stainless-steel side walls, three clear polycarbonate walls and ceiling, and a stainless-steel grid floor. Each chamber was equipped with a recessed magazine connected to a 45mg grain pellet (F0165; Bio-Serv Biotechnologies, Flemington, NJ, USA; one pellet per delivery) dispenser and a pump fitted with a 60mL syringe that delivered a 0.2mL sucrose solution (20% w/v; delivered across 2 seconds; white sugar; Winc, Erskine Park, NSW, Australia) into the left side of the magazine. Head entries into the magazine were detected via an infrared beam that crossed the magazine opening. One retractable lever was located on either side of the magazine. A house light (3W, 24V) situated at the top of the wall opposite to the food magazine was used to illuminate the chamber during the training and test sessions. Each chamber contained a Sonalert white noise generator (80dB), and a 28V DC mechanical relay that delivered a 2Hz clicker stimulus. Training and testing sessions were programmed and controlled by computers external to the testing rooms using MED-PC V software (MED Associates, Fairfax, VT, USA) which also recorded experimental data from each session. All cabinets were fitted with cameras (IPC-K35AP; Dahua, Artarmon, NSW, Australia) to view real-time chamber activity using D-ViewCam DCS-100 software (D-Link, North Ryde, NSW, Australia). The ceiling of these chambers contained a hole which enabled the connection of the patch cable between the LED and each freely moving rat.

Outcome devaluation was conducted in a separate room equipped with 16 individual, open-top plastic boxes with stainless steel wire mesh lids. During devaluation, the lights were dimmed, and each individual chamber was fitted with either a ceramic dish for grain pellet devaluation, or a plastic sipper bottle for sucrose solution devaluation.

Optogenetic equipment

Fiber-optic cannulas (10mm; Doric Lenses, Québec, QC, Canada) consisted of a zircona ferrule contained within a stainless-steel receptacle from which a silica/polymer flat-tip fiber extended from. Fiber-optic patch cords (custom; Doric Lenses, Québec, QC, Canada) were enclosed in a flexible cladding that consisted of one transparent fiber that split into two smaller fibers for attachment to the implanted fiber-optic cannulas. Delivery of the LED light was controlled through a TTL adapter which converted 28V DC output to a TTL transition (MED Associates, Fairfax, VT, USA). The adaptor was connected to an LED driver (Doric Lenses, Québec, QC, Canada) and Connectorised LED light source (Doric Lenses, Québec, QC, Canada) with a fiber-optic rotary joint for attachment to the fiber-optic patch cords. LED light was delivered as orange light (625nm; continuous) and measured at least 10mW at the cannula tip using a photometer (PM200; ThorLabs, Newton, NJ, USA) when the LED was switched on.

Behavioral procedures

Pavlovian conditioning

Pavlovian conditioning involved 8 daily 1hr training sessions. Two auditory conditioned stimuli (S1 and S2; white noise or clicker) were paired with two food outcomes (O1 and O2; grain pellets or sucrose solution). Each stimulus was presented for 2min in duration, 4 times each in a pseudorandom order within each training session. Stimulus presentations were separated by pseudorandom intertrial intervals (ITIs) ranging between 3 to 5min in length (4min average). The stimulus order and ITIs were changed daily. Stimulus-outcome (S-O) pairings were counterbalanced within-and between-groups. Food outcomes were delivered on a random time 30s schedule throughout the stimulus presentation. Magazine entries were recorded throughout the session, and conditioned responding was analyzed by comparing magazine entries in the 2min prior to stimulus presentation and during the 2min stimulus presentation period.

Instrumental conditioning

During instrumental conditioning, two actions (A1 and A2; left or right lever presses) were trained with the two dikerent outcomes (O1 and O2) in 2 separate daily sessions for 8 days. The order of the sessions was counterbalanced, as were the A-O relationships, which were further counterbalanced with the S-O relationships. Each session ended when 20 outcomes were earned or when 30min had elapsed. For the first 2 days, actions were continuously reinforced (i.e. one action earned one outcome delivery). Across the next three sessions, actions were rewarded on a random ratio 5 (RR5) schedule (i.e. each action delivered an outcome with a probability of 0.2). For the final 3 days, animals were rewarded on an RR10 schedule (i.e. each response delivered an outcome with a probability of 0.1). In the VP experiments (NAc-S D1-and D2-SPNs to VP projections), rats only received 2 days on RR10. Lever press actions were measured throughout the duration of the session, and performance during instrumental training was assessed using the mean number of lever presses per min.

Pavlovian Instrumental transfer test

Across the three days preceding this test, rats were given one session of Pavlovian retraining and two separate sessions of instrumental retraining. In the VP experiments (NAc-S D1-and D2-SPNs→VP projections), rats first received one day of Pavlovian conditioning followed by 3 days of instrumental conditioning (RR10). Immediately before the test, patch cables were attached to the fiber-optic cannulas and the LED. During the test, both levers were inserted into the chamber simultaneously and remained in the chamber for the duration of the test. The test was conducted in extinction (i.e. no outcomes were delivered). To reduce baseline (pre S) lever presses, responding on both levers was extinguished for 8min prior to any stimuli being presented. Each stimulus was presented for 2min, four times each during the test in the following order: noise-clicker-clicker-noise-clicker-noise-noise-clicker. Each stimulus presentation was separated by a fixed 3min interval. Critically, the LED was activated during half of the trials (ON) but left inactive during the other half (OFF). The order for ON and OFF trials was counterbalanced. Responses were recorded prior to and during the S presentation. These were then collated into ‘pre’ (lever press rates in the two min prior to S presentation), ‘same’ (when S shares the same outcome with the action) and ‘dikerent’ (when S has a di9erent outcome with the action) following data collection.

Choice tests following outcome devaluation

Across the two consecutive days preceding these tests, rats were given 2 daily instrumental conditioning sessions (RR10). Following the second session, rats were habituated to the devaluation chambers for 30min. During these habituation sessions, animals had access to one outcome (counterbalanced). On the first test day, rats were placed in the devaluation chambers for 1hr where they had ad libitum access to one of the previously trained outcomes (O1 or O2). Once removed from the devaluation boxes, patch cables were attached to the fiber-optic cannulas and the LED, and the first choice test was conducted. Both levers were made available for 2min each, four times during the session with an intertrial interval of 2min where the levers were retracted. The LED was activated during 2 of the 2min trials where the levers were available (ON) and was left inactive during the other 2 trials (OFF). The order of the ON and OFF trials was counterbalanced. A similar procedure was implemented the following day except that the other outcome was devalued. Lever presses and magazine entries were recorded while the levers were available.

Tissue preparation and histology

Transcardial fixation and brain sectioning

Rats were euthanised with an intraperitoneal injection of sodium pentobarbital (1mL) and transcardially perfused with 4% paraformaldehyde (PFA) in 0.1M sodium phosphate buker (PBS, pH 7.5; 400mL for rats). Brains were immediately removed and placed in individual specimen jars containing PFA and stored at 4°C overnight for post-fixation. Brains were sliced into 40μm coronal or sagittal sections in 0.1M Tris-bukered saline (TBS, 0.25M Tris, 0.5M NaCl, pH 7.5) using a vibratome (VT1200; Leica Microsystems, North Ryde, NSW, Australia) and stored at-20°C in cryoprotective solution containing 30% ethylene glycol, 30% glycerol and 0.1M PBS.

Immunofluorescence

Free-floating sections were selected in TBS to contain the region of interest and washed three times for 10min each in TBS solution. Sections were then incubated in 0.5% Triton-X, 10% Normal Horse Serum (NHS) in TBS blocking solution for two hours. Sections were washed again in TBS solution three times for 10 min each. Sections were then incubated between 12-48h (dependent on antibody) in a solution containing the primary antibodies in 0.2% Triton-X, 2% NHS in TBS at 4°C. The primary antibodies were the following: purified mouse anti DARPP-32 (1:300; BD Biosciences, Macquarie Park, Australia; #AB398980) and polyclonal rabbit anti-GFP (1:1000; Invitrogen, Mulgrave, Australia; #AB221569). After the specified incubation period, the sections were washed for three times in TBS for 10min each and then transferred into a solution containing the secondary antibodies in TBS solution at room temperature for 1h. The secondary antibodies were the following: donkey anti-mouse Cy3 (1:400; Jackson ImmunoResearch Laboratories, West Grove, USA; #AB2340813) and donkey anti-rabbit Alexa-488 (1:500; Jackson ImmunoResearch Laboratories, West Grove, USA; #AB2313584). Next, sections were washed three times in TBS for 10min and then mounted onto glass slides using Vectashield hardset antifade mounting medium (H-1400; Vector Laboratories, Newark, CA, USA). Sections that did not require immunofluorescence staining were washed three times for 10min each in TBS and mounted using Vectashield hardset antifade mounting medium.

Microscopy and imaging

Imaging of the mounted brain sections was performed on a spinning disk confocal microscope (Zyla 4.2 sCMOS; Andor Technology, Belfast, UK) using the Diskovery multi-modal imaging platform (Oxford Instruments, Belfast, UK). The Diskovery platform was mounted on an inverted microscope body with a motorized stage and Perfect Focus System (Eclipse Ti; Nikon, Melville, NY, USA). Image capture was controlled using the NIS-Elements Software (Nikon, Melville, NY, USA). Single high-magnification field-of-view images were generated by automatic stitching of multiple adjacent frames from within a defined area using the motorized stage (optical magnification: 20X; pixel depth: 16-bit). All images were processed using Fiji ImageJ open-source software (Rueden et al., 2017). Additional images were taken with a confocal microscope (BX61W1; Olympus, Shinjuku, Tokyo, Japan) using Fluoview software (FV1000; Olympus, Shinjuku, Tokyo, Japan) for further placement analysis. The images were used to assess neuronal pathways, as well as the spread of viral infusions, placement of fiber-optic cannulas and potential damages due to the surgical procedures. Based on the latter, the following numbers of rats were excluded from the final statistical analyses due to misplaced viral or fiber-optics misplacements: NAc-S D1-SPNs silencing (n = 9 females and 9 males), NAc-S D2-SPNs silencing (n = 4 females and 5 males), NAc-S D1-SPNs→VP silencing (n = 5 females and 10 males), NAc-S D2-SPNs→VP silencing (n = 4 females and 1 male).

Immunofluorescence quantifications

The specificity of viral transduction in D1-Cre and A2-Cre rats (Figures S1 C-F) was achieved through manual cell counting and involved calculating a percentage of the total number of fluorophore-labeled (eYFP) cells relative to the total number of DARPP-32-positive cells, and a percentage of the total number of DARPP-32-positive cells relative to the total number of fluorophore-labeled (eYFP) cells. Two slices per rat were selected around the following anteroposterior coordinates (in mm relative to bregma):-0.45 for DS and +1.8 for NAc-S. Before quantification, all images were randomly renumbered, and cell counts were performed in Fiji Image J in one region of interest (ROI) per image (ROI size: 0.25 mm2 for both DS and NAc-S). The ROI was defined in the eYFP channel around the viral injection site. The average number of DARPP-32-positive cells per ROI was: 194 ± 59 for DS and 132± 24 for NAc-S.

Electrophysiological recordings

Brain slice preparation

Rats were euthanized under isoflurane gas anesthesia (4% in air). Brains were immediately removed and sliced using a vibratome (VT1200; Leica Microsystems, North Ryde, NSW, Australia) in ice-cold oxygenated sucrose buker solution (consisting of [in mM]: 241 sucrose, 28 NaHCO3, 11 glucose, 1.4 NaH2PO4, 3.3 KCl, 0.2 CaCl2, 7 MgCl2). Coronal brain slices (300μm thickness) containing the NAc were sampled and maintained at 33°C in a submerged chamber containing physiological saline (consisting of [in mM]: 124 NaCl, 2.5 KCl, 1.25 NaH2PO4, 1 MgCl2, 1 CaCl2, 10 glucose and 26 NaHCO3) and equilibrated with 95% O2 and 5% CO2.

Recordings

Following 1h equilibration, slices were transferred to a recording chamber. An upright microscope (BX50WI; Olympus, Shinjuku, Tokyo, Japan) was used to visualise neurons using dikerential interference contrast (DIC) Dodt tube optics and eYFP fluorescence, and superfused continuously (1.5mL/min) with oxygenated physiological saline at 33°C. Whole-cell patch-clamp recordings were performed using glass pipettes (tip size 2–5 MΩ) containing internal solution (consisting of [in mM]: 115 K gluconate, 20 NaCl, 1 MgCl2, 10 HEPES, 11 EGTA, 5 Mg-ATP, 0.33 Na-GTP, and 5 phosphocreatine di(tris) salt, pH 7.3, osmolarity 285-290 mOsm/L). Neurons sampled during recording were marked with Biocytin (0.1%) added to the internal solution. A Multiclamp 700B amplifier (Molecular Devices, San Jose, CA, USA) connected to a Macintosh computer and interface ITC-18 (InstruTECH, Longmont, CO, USA) was used for data acquisition. Liquid junction potentials of-10mV were not corrected. In current-clamp mode, membrane potentials were sampled at 5kHz (low pass filter 2kHz; Axograph X, Axograph, CA, USA). Stock solution of drug was diluted to working concentration in the extracellular solution immediately before use and applied by continuous superfusion. Data from whole-cell recordings were included in analyses if the neurons appeared healthy under DIC on the monitor screen.

LED stimulation

Transfected neurons with eNpHR3.0 or eYFP in the NAc-S were illuminated by a continuous wave of LED light (625nm, 2 minutes, 2mW; ThorLabs, Newton, NJ, USA) onto the brain slice under the 40X water-immersion objective. Since SPNs in ex-vivo brain slices are not spontaneously active, brief injection of positive current step pulses were applied to make them fire action potentials.

Post-hoc histological analysis

Immediately after physiological recording, brain slices were fixed overnight using 4% PFA in 0.16M phosphate buker (PB) solution, washed and then transferred into 0.5% Triton-X in PB solution for 3 days to permeabilise cells. Slices were washed, mounted onto glass slides, and then coverslipped with Fluoromount-G mounting medium (Southern Biotech, Birmingham, AL, USA). A 2D projection image was then obtained from a collated image stack using the confocal microscope (X61W1; Olympus, Shinjuku, Tokyo, Japan).

Statistical analyses

The data presented met the assumptions of the statistical test used. The main measures were the number of magazine entries and lever presses per minute. The dikerence between groups were analyzed using a planned orthogonal contrast procedure controlling the per contrast error rate (Hays, 1963). A first contrast (Group) compared rats that had been infused with the control eYFP virus and rats that had been infused with the inhibitory eNpHR3.0 virus. Changes across days (Day) during Pavlovian and instrumental conditioning were assessed by planned linear trend analyses. Pavlovian conditioning and the PIT tests included a within-subject factor (Period) comparing magazine entry rates during the 2min stimuli against rates during a 2min baseline period immediately preceding the stimuli. During the PIT tests, a within-subject factor (Lever) examined the net (i.e., minus baseline) rates of lever presses on the response (Same) earning the same outcome as predicted by the stimulus against those on the response (Dikerent) earning the dikerent outcome. A second within-subject factor (LED) examined these rats when the LED was activated (ON) or not activated (OFF). The choice tests following outcome devaluation used a within-subject factor of response identity (Lever) comparing rates of lever presses on the response (Valued) earning the valued outcome against those on the response (Devalued) earning the devalued outcome. It also used the LED within-subject factor described above. All analyses were carried out using the PSY statistical software (School of Psychology, the University of New South Wales, Australia). The Type I error rate was controlled at alpha = 0.05 for each contrast tested. If interactions were detected, follow-up simple ekects analyses were calculated to determine the source of the interactions. For each significant statistical comparison, we report measures of ekect size partial eta-squared (η2; η2 = 0.01 is a small ekect, η2 = 0.06 a medium ekect and η2 = 0.14 a large ekect). Data will be made available upon request.

eLife Assessment

This study provides novel and convincing evidence that both dopamine D1 and D2 expressing neurons in the nucleus accumbens shell are crucial for the expression of cue-guided action selection, a core component of decision-making. The research is systematic and rigorous in using optogenetic inhibition of either D1- or D2-expressing medium spiny neurons in the NAc shell to reveal attenuation of sensory-specific Pavlovian-Instrumental transfer, while largely sparing value-based decision on an instrumental task. The important findings in this report build on prior research and resolve some conflicts in the literature regarding decision-making.

Reviewer #1 (Public review):

In the current article, Octavia Soegyono and colleagues study "The influence of nucleus accumbens shell D1 and D2 neurons on outcome-specific Pavlovian instrumental transfer", building on extensive findings from the same lab. While there is a consensus about the specific involvement of the Shell part of the Nucleus Accumbens (NAc) in specific stimulus-based actions in choice settings (and not in General Pavlovian instrumental transfer - gPIT, as opposed to the Core part of the NAc), mechanisms at the cellular and circuitry levels remain to be explored. In the present work, using sophisticated methods (rat Cre-transgenic lines from both sexes, optogenetics and the well-established behavioral paradigm outcome-specific PIT - sPIT), Octavia Soegyono and colleagues decipher the differential contribution of dopamine receptors D1 and D2 expressing-spiny projection neurons (SPNs).

After validating the viral strategy and the specificity of the targeting (immunochemistry and electrophysiology), the authors demonstrate that while both NAc Shell D1- and D2-SPNs participate in mediating sPIT, NAc Shell D1-SPNs projections to the Ventral Pallidum (VP, previously demonstrated as crucial for sPIT), but not D2-SPNs, mediates sPIT. They also show that these effects were specific to stimulus-based actions, as value-based choices were left intact in all manipulations.

This is a well-designed study and the results are well supported by the experimental evidence. The paper is extremely pleasant to read and add to the current literature.

Comments on revisions:

We thank the authors for their detailed responses and for addressing our comments and concerns.

To further improve consistency and transparency, we kindly request that the authors provide, for Supplemental Figures S1-S4, panels E (raw data for lever presses during the PIT test), the individual data points together with the corresponding statistical analyses in the figure legends.

In addition, regarding Supplemental Figure S3, panel E, we note the absence of a PIT effect in the eYFP group under the ON condition, which appears to differ from the net response reported in the main Figure 5, panel B. Could the authors clarify this apparent discrepancy?

We also note a discrepancy between the authors' statement in their response ("40 rats excluded based on post-mortem analyses") and the number of excluded animals reported in the Materials and Methods section, which adds up to 47. We kindly ask the authors to clarify this point for consistency.

Finally, as a minor point, we suggest indicating the total number of animals used in the study in the Materials and Methods section.

Reviewer #2 (Public review):

Summary:

This manuscript by Soegyono et a. describes a series of experiments designed to probe the involvement of dopamine D1 and D2 neurons within the nucleus accumbens shell in outcome-specific Pavlovian-instrumental transfer (osPIT), a well-controlled assay of cue-guided action selection based on congruent outcome associations. They used an optogenetic approach to phasically silence NAc shell D1 (D1-Cre mice) or D2 (A2a-Cre mice) neurons during a subset of osPIT trials. Both manipulations disrupted cue-guided action selection but had no effects on negative control measures/tasks (concomitant approach behavior, separate valued guided choice task), nor were any osPIT impairments found in reporter only control groups. Separate experiments revealed that selective inhibition of NAc shell D1 but not D2 inputs to ventral pallidum were required for osPIT expression, thereby advancing understanding of the basal ganglia circuitry underpinning this important aspect of decision making.

Strengths:

The combinatorial viral and optogenetic approaches used here were convincingly validated through anatomical tract-tracing and ex vivo electrophysiology. The behavioral assays are sophisticated and well-controlled to parse cue and value guided action selection. The inclusion of reporter only control groups is rigorous and rules out nonspecific effects of the light manipulation. The findings are novel and address a critical question in the literature. Prior work using less decisive methods had implicated NAc shell D1 neurons in osPIT but suggested that D2 neurons may not be involved. The optogenetic manipulations used in the current study provides a more direct test of their involvement and convincingly demonstrate that both populations play an important role. Prior work had also implicated NAc shell connections to ventral pallidum in osPIT, but the current study reveals the selective involvement of D1 but not D2 neurons in this circuit. The authors do a good job of discussing their findings, including their nuanced interpretation that NAc shell D2 neurons may contribute to osPIT through their local regulation of NAc shell microcircuitry.

Weaknesses:

The current study exclusively used an optogenetic approach to probe the function of D1 and D2 NAc shell neurons. Providing a complementary assessment with chemogenetics or other appropriate methods would strengthen conclusions, particularly the novel demonstration for D2 NAc shell involvement. Likewise, the null result of optically inhibiting D2 inputs to ventral pallidum leaves open the possibility that a more complete or sustained disruption of this pathway may have impaired osPIT.

Conclusions:

The research described here was successful in providing critical new insights into the contributions of NAc D1 and D2 neurons in cue-guided action selection. The authors' data interpretation and conclusions are well reasoned and appropriate. They also provide a thoughtful discussion of study limitations and implications for future research. This research is therefore likely to have a significant impact on the field.

Comments on revisions:

I have reviewed the rebuttal and revised manuscript and have no remaining concerns.

Author response:

The following is the authors’ response to the previous reviews

Reviewer#1 (Public Review):

In the current article, Octavia Soegyono and colleagues study "The influence of nucleus accumbens shell D1 and D2 neurons on outcome-specific Pavlovian instrumental transfer", building on extensive findings from the same lab. While there is a consensus about the specific involvement of the Shell part of the Nucleus Accumbens (NAc) in specific stimulus-based actions in choice settings (and not in General Pavlovian instrumental transfer - gPIT, as opposed to the Core part of the NAc), mechanisms at the cellular and circuitry levels remain to be explored. In the present work, using sophisticated methods (rat Cre-transgenic lines from both sexes, optogenetics and the well-established behavioral paradigm outcome-specific PIT - sPIT), Octavia Soegyono and colleagues decipher the diOerential contribution of dopamine receptors D1 and D2 expressing-spiny projection neurons (SPNs).

After validating the viral strategy and the specificity of the targeting (immunochemistry and electrophysiology), the authors demonstrate that while both NAc Shell D1- and D2SPNs participate in mediating sPIT, NAc Shell D1-SPNs projections to the Ventral Pallidum (VP, previously demonstrated as crucial for sPIT), but not D2-SPNs, mediates sPIT. They also show that these eOects were specific to stimulus-based actions, as valuebased choices were left intact in all manipulations.

This is a well-designed study and the results are well supported by the experimental evidence. The paper is extremely pleasant to read and add to the current literature.

We thank the Reviewer for their positive assessment.

Comments on revisions:

We thank the authors for their detailed responses and for addressing our comments and concerns.

To further improve consistency and transparency, we kindly request that the authors provide, for Supplemental Figures S1-S4, panels E (raw data for lever presses during the PIT test), the individual data points together with the corresponding statistical analyses in the figure legends.

Panel E of Figures S1-S4 now includes the individual data points. The outcome-specific data have already been analysed, and we report these analyses in the main manuscript. These analyses are more informative than those requested by the Reviewer since they report the net eFects of the stimuli on choice between actions while controlling for potential individual baseline instrumental performance. All data remain fully transparent and are publicly available on an online repository in accordance with eLife policies (see relevant section in Materials and Methods).

In addition, regarding Supplemental Figure S3, panel E, we note the absence of a PIT eOect in the eYFP group under the ON condition, which appears to diOer from the net response reported in the main Figure 5, panel B. Could the authors clarify this apparent discrepancy?

We apologize for the error, which has now been corrected.

We also note a discrepancy between the authors' statement in their response ("40 rats excluded based on post-mortem analyses") and the number of excluded animals reported in the Materials and Methods section, which adds up to 47. We kindly ask the authors to clarify this point for consistency.

We thank the Reviewer for identifying the error reported in our initial response. The total number of animals excluded was 47, as reported in the manuscript.

Finally, as a minor point, we suggest indicating the total number of animals used in the study in the Materials and Methods section.

The total number of animals has been included in the Materials and Methods section.

Reviewer #2 (Public Review):

Summary:

This manuscript by Soegyono et a. describes a series of experiments designed to probe the involvement of dopamine D1 and D2 neurons within the nucleus accumbens shell in outcome-specific Pavlovian-instrumental transfer (osPIT), a well-controlled assay of cueguided action selection based on congruent outcome associations. They used an optogenetic approach to phasically silence NAc shell D1 (D1-Cre mice) or D2 (A2a-Cre mice) neurons during a subset of osPIT trials. Both manipulations disrupted cue-guided action selection but had no eOects on negative control measures/tasks (concomitant approach behavior, separate valued guided choice task), nor were any osPIT impairments found in reporter only control groups. Separate experiments revealed that selective inhibition of NAc shell D1 but not D2 inputs to ventral pallidum were required for osPIT expression, thereby advancing understanding of the basal ganglia circuitry underpinning this important aspect of decision making.

Strengths:

The combinatorial viral and optogenetic approaches used here were convincingly validated through anatomical tract-tracing and ex vivo electrophysiology. The behavioral assays are sophisticated and well-controlled to parse cue and value guided action selection. The inclusion of reporter only control groups is rigorous and rules out nonspecific eOects of the light manipulation. The findings are novel and address a critical question in the literature. Prior work using less decisive methods had implicated NAc shell D1 neurons in osPIT but suggested that D2 neurons may not be involved. The optogenetic manipulations used in the current study provides a more direct test of their involvement and convincingly demonstrate that both populations play an important role. Prior work had also implicated NAc shell connections to ventral pallidum in osPIT, but the current study reveals the selective involvement of D1 but not D2 neurons in this circuit. The authors do a good job of discussing their findings, including their nuanced interpretation that NAc shell D2 neurons may contribute to osPIT through their local regulation of NAc shell microcircuitry.

We thank the Reviewer for their positive assessment.

Weaknesses:

The current study exclusively used an optogenetic approach to probe the function of D1 and D2 NAc shell neurons. Providing a complementary assessment with chemogenetics or other appropriate methods would strengthen conclusions, particularly the novel demonstration for D2 NAc shell involvement. Likewise, the null result of optically inhibiting D2 inputs to ventral pallidum leaves open the possibility that a more complete or sustained disruption of this pathway may have impaired osPIT.

We acknowledge the reviewer's valuable suggestion that demonstrating NAc-S D1- and D2-SPNs engagement in outcome-specific PIT through another technique would strengthen our optogenetic findings. Several approaches could provide this validation. Chemogenetic manipulation, as the reviewer suggested, represents one compelling option. Alternatively, immunohistochemical assessment of phosphorylated histone H3 at serine 10 (P-H3) oFers another promising avenue, given its established utility in reporting striatal SPNs plasticity in the dorsal striatum (Matamales et al., 2020). We hope to complete such an assessment in future work since it would address the limitations of previous work that relied solely on ERK1/2 phosphorylation measures in NAc-S SPNs (Laurent et al., 2014). The manuscript was modified to report these future avenues of research (page 12).

Regarding the null result from optical silencing of D2 terminals in the ventral pallidum, we agree with the reviewer's assessment. While we acknowledge this limitation in the current manuscript (page 13), we aim to address this gap in future studies to provide a more complete mechanistic understanding of the circuit.

Conclusions:

The research described here was successful in providing critical new insights into the contributions of NAc D1 and D2 neurons in cue-guided action selection. The authors' data interpretation and conclusions are well reasoned and appropriate. They also provide a thoughtful discussion of study limitations and implications for future research. This research is therefore likely to have a significant impact on the field.

We thank the Reviewer for their positive assessment.

Comments on revisions:

I have reviewed the rebuttal and revised manuscript and have no remaining concerns.

We are pleased to have addressed the Reviewer’s query.

References

Laurent, V., Bertran-Gonzalez, J., Chieng, B. C., & Balleine, B. W. (2014). δ-Opioid and Dopaminergic Processes in Accumbens Shell Modulate the Cholinergic Control of Predictive Learning and Choice. J Neurosci, 34(4), 1358-1369. https://doi.org/10.1523/JNEUROSCI.4592-13.2014

Matamales, M., McGovern, A. E., Mi, J. D., Mazzone, S. B., Balleine, B. W., & BertranGonzalez, J. (2020). Local D2- to D1-neuron transmodulation updates goal-directed learning in the striatum. Science, 367(6477), 549-555. https://doi.org/10.1126/science.aaz5751

Version 3

Abstract

The nucleus accumbens shell (NAc-S) and its projections to the ventral pallidum (VP) are thought to be critical for stimulus-based decisions. The NAc-S is predominantly composed of spiny projection neurons (SPNs) that express either the dopamine D1 (D1-SPNs) or the dopamine D2 receptor (D2-SPNs). Yet, the role of these two neuronal subpopulations and their inputs to the VP in stimulus-based decisions remains unknown. Here, we used optogenetics in female and male knock-in rats to selectively silence D1-or D2-SPNs and their projections to the VP at a time when the rats were required to use predictive stimuli to choose between two instrumental actions. Silencing either population of NAc-S SPNs disrupted choice. Silencing NAc-S D1-SPNs terminals in the VP also disrupted choice. However, choice was left intact by silencing NAc-S D2-SPNs terminals in the VP. Together, these findings provide novel insights into the cellular mechanisms and circuitry underlying stimulus-based decisions. We discuss how these insights are consistent with a recent model proposing that these decisions are controlled by an opioid-based memory system residing in the NAc-S.

Introduction

Our choices and decisions are often influenced by predictive signals available in the environment (Hollis, 1984). This influence is studied in the laboratory using the outcome-specific Pavlovian-instrumental transfer (PIT) task (Trapold & Overmier, 1972; Colwill & Rescorla, 1988; Holmes et al., 2010; Cartoni et al., 2016; Laurent & Balleine, 2021; Leung et al., 2024a), which comprises three stages. The first two stages are Pavlovian and instrumental conditioning, which can be administered in any order. In Pavlovian conditioning, two stimuli (S1 and S2; e.g., sounds or lights) are paired with two distinct and motivationally significant outcomes (O1 and O2; e.g., food outcomes). In instrumental conditioning, subjects are trained to perform two actions (A1 and A2), with each earning one of the two outcomes (i.e., A1→O1 and A2→O2). The final stage is the PIT test, which evaluates choice between the two actions in the presence (and absence) of the stimuli. Typically, each of the Pavlovian stimuli bias choice towards the action with which it shares an outcome: i.e., S1: A1>A2 and S2: A2>A1.

At a neural level, PIT has been found to rely on interactions between a number of structures, including cortical (Keistler et al., 2015; Bradfield et al., 2015; Bradfield et al., 2018; Tensaouti et al., 2025; Ostlund & Balleine, 2007; Lichtenberg et al., 2021; Sias et al., 2021; Lichtenberg et al., 2017), amygdala (Corbit & Balleine, 2005; Morse et al., 2020; Sias et al., 2024; Lichtenberg et al., 2017; Shiflett & Balleine, 2010; Derman et al., 2020; Prévost et al., 2012), basal ganglia (Corbit & Balleine, 2011; Corbit et al., 2001; Corbit & Janak, 2010; Morse et al., 2020; Laurent et al., 2012; Bertran-Gonzalez et al., 2013; Laurent et al., 2014; Corbit et al., 2016; Leung et al., 2024b; Leung & Balleine, 2015; Leung & Balleine, 2013; Laurent et al., 2015; Prévost et al., 2012), midbrain (Corbit et al., 2007; Sias et al., 2024; Seitz et al., 2022) and thalamic (Ostlund & Balleine, 2008; Leung et al., 2024b) territories. Among these, the nucleus accumbens shell (NAc-S) stands out because, unlike the others, it does not itself contribute to learning the stimulus-outcome (S-O) or the action-outcome (A-O) associations formed during the Pavlovian and instrumental stages of the PIT task and appears only to mediate their interaction (Corbit & Balleine, 2011; Corbit et al., 2001; Morse et al., 2020; Laurent et al., 2012; Laurent et al., 2014). Therefore, it has been proposed that the NAc-S acts as a central hub integrating S-O and A-O information at the time of PIT to guide choice between actions in an outcome-selective manner (Shiflett & Balleine, 2010; Bertran-Gonzalez & Laurent, 2018; Laurent & Balleine, 2021). Accordingly, PIT is disrupted by manipulations undermining NAc-S function during the PIT test (Morse et al., 2020; Corbit et al., 2016; Laurent et al., 2012; Laurent et al., 2014; Laurent et al., 2015).

The NAc-S is predominantly composed of spiny projection neurons (SPNs), which can be classified into two distinct subpopulations depending on the dopamine receptor they express (Gerfen & Surmeier, 2011). One population harbors the dopamine D1 receptors (D1Rs; D1-SPNs) whereas the other population express the dopamine D2 receptors (D2Rs; D2-SPNs). To date, only one study has examined whether activity in both populations is recruited during PIT (Laurent et al., 2014), finding that PIT is associated with an increase in ERK1/2 phosphorylation in D1-SPNs but not D2-SPNs. Further, blockade of D1Rs in NAc-S during choice between actions disrupted PIT whereas D2Rs blockade had no ekect. Although these findings indicate a major involvement of NAc-S D1-SPNs in PIT, caution must be exerted when drawing conclusions on the role of D2-SPNs. While D1Rs are exclusively expressed on NAc-S D1-SPNs, D2Rs are present on cholinergic interneurons (CINs) and local presynaptic dopamine terminals in addition to D2-SPNs (Gerfen & Surmeier, 2011). Additionally, research shows that ERK1/2 phosphorylation may not capture all transcriptional events that occur in D2-SPNs (Bertran-Gonzalez et al., 2008; Matamales et al., 2020). Lastly, pharmacological D2Rs blockade has been found to be quite inekective at influencing D2-SPNs function (Tozzi et al., 2007). Thus, the relative contribution of NAc-S D1-and D2-SPNs in driving choice between actions during PIT remains unclear.

The present experiments therefore aimed to provide an unambiguous assessment of the roles played by the two main populations of NAc-S SPNs during choice between actions in an outcome-specific PIT task. This assessment was achieved through optogenetic silencing in two knock-in rat lines that express Cre recombinase in either D1-or D2-SPNs (Pettibone et al., 2019). We first used tract-tracing and ex-vivo electrophysiology to confirm our capacity to selectively silence activity in NAc-S D1-or D2-SPNs. We then conducted two experiments that assessed outcome-specific PIT while D1-or D2-SPNs in the NAc-S were silenced during presentations of the predictive stimuli in the choice test. Finally, we assessed the degree to which the influence of the two SPNs populations on outcome-specific PIT depends on their projections to the ventral pallidum (VP). We focused on the VP for two main reasons: First, both NAc-S D1-and D2-SPNs send dense projections to this region (Lu et al., 1997; Kupchik et al., 2015); and, second, communication between the NAc-S and VP has previously been found to be critical for outcome-specific PIT (Leung & Balleine, 2013).

Results

Anterograde tracing and ex-vivo cell recordings in D1-Cre and A2a-Cre rats

We first examined the connectivity of striatal D1-and D2-SPNs in transgenic rats expressing Cre recombinase in neurons harboring the dopamine D1 receptor (D1-Cre rats) or the adenosine A2a receptor (A2a-Cre rats) (Pettibone et al., 2019). A2a receptors are predominantly expressed on striatal D2-SPNs (Schikmann et al., 2007), thereby enabling selective targeting of these SPNs in the A2a-Cre line. D1-Cre (n = 1 female and 1 male) and A2a-Cre rats (n = 1 female and 1 male) were unilaterally infused with a Cre-dependent eYFP virus in the NAc-S (Figure 1 A). To validate expression in these rat lines, other D1-Cre (n = 4 males) and A2a-Cre rats (n = 3 males) were given this unilaterally infusion in the dorsal striatum (DS) with the same virus. In both transgenic strains, viral expression was largely restricted to local SPNs in the two targeted regions: ∼95% of eYFP-positive neurons expressed DARPP-32 (D32 | eYFP; Figures 1 C-F). Viral expression was consistent with D1-and D2-SPNs representing two neuronal populations similar in size: in both transgenic strains and targeted regions ∼40% of DARPP-32-postive neurons expressed eYFP (eYFP | D32; Figures 1 C-F). DS infusion in D1-Cre rats resulted in eYFP-positive terminals (Figures 1 B-C) in the substantia nigra pars reticulata (SNr) and the globus pallidus externus (GPe). By contrast, the same infusion in A2a-Cre rats only produced eYFP-positive terminals in the GPe (Figures 1 B and D). NAc-S infusion in both D1-Cre and A2a-Cre rats resulted in eYFP-positive terminals in the VP (Figures 1 E-F). By contrast, eYFP-positive terminals in the lateral hypothalamus (Figures 1 E-F) were only observed in D1-Cre rats. Overall, these results are in line with current knowledge about the cellular composition of the striatum (Gerfen & Surmeier, 2011; Bertran-Gonzalez et al., 2010) and well-established projection patterns of striatal D1-and D2-SPNs (Gerfen & Surmeier, 2011; Lu et al., 1997; Kupchik et al., 2015; O’Connor et al., 2015). We were therefore confident in the capacity of the two rat transgenic lines to reveal the function of NAc-S D1-and D2-SPNs.

Anterograde tracing in D1-Cre and A2a-Cre rats.

A. D1-Cre or A2a-Cre rats were unilaterally infused in the dorsal striatum (DS; D1-Cre: n = 4 males; A2A-Cre: n = 3 males) or nucleus accumbens shell (NAc-S; D1-Cre: n = 1 female and 1 male; A2A-Cre: n = 1 females and 1 male) with DIO-eYFP. B. Sagittal micrographs obtained in D1-Cre (top) and A2A-Cre (bottom) rats following viral infusion in the DS. C. DIO-eYFP infusion in the DS of D1-Cre rats. Micrographs show eYFP expression in the DS, D32 staining and co-labelling (eYFP + D32) in the DS. They also show that DS D1-SPNs project to the substantia nigra pars reticulata (SNr) and the globus pallidus externus (GPe). Viral expression was restricted to putative SPNs (D32 | eYFP), with ∼40% of SPNs expressing eYFP (eYFP | D32). D. DIO-eYFP infusion in the DS of A2a-Cre rats. Micrographs show eYFP expression in the DS, D32 staining and co-labelling (eYFP + D32) in the DS. They also show that DS D2-SPNs project to the SNr but not the GPe. Viral expression was restricted to putative SPNs (D32 | eYFP), with ∼40% of SPNs expressing eYFP (eYFP | D32). E. DIO-eYFP infusion in the NAc-S of D1-Cre rats. Micrographs show eYFP expression in the NAc-S, D32 staining and co-labelling (eYFP + D32) in the DS. They also show that NAc-S D1-SPNs project to the ventral pallidum (VP) and the lateral hypothalamus (LH). Viral expression was restricted to putative SPNs (D32 | eYFP), with ∼41% of SPNs expressing eYFP (eYFP | D32). F. DIO-eYFP infusion in the NAc-S of A2a-Cre rats. Micrographs show eYFP expression in the NAc-S, D32 staining and co-labelling (eYFP + D32) in the DS. They also show that NAc-S D2-SPNs project to the VP but not the LH. Viral expression was restricted to putative SPNs (D32 | eYFP), with ∼40% of SPNs expressing eYFP (eYFP | D32).

Next, we used ex-vivo electrophysiological recordings to assess our capacity to silence activity of NAc-S D1-and D2-SPNs. D1-Cre and A2a-Cre rats were bilaterally infused in the NAc-S with either a null Cre-dependent eYFP virus (D1-Cre: 2 females; A2a-Cre: 2 females) or a Cre-dependent inhibitory halorhodospin (eNpHR3.0; D1-Cre: 2 females; A2a-Cre: 2 females) virus (Figures 2 A-B). LED light illumination (625 nm, continuous) of either NAc-S D1-SPNs (Figure 2 A) or NAc-S D2-SPNs (Figure 2 B) transfected with the eYFP virus had no ekect on action potential firing (Baseline vs. LED ON; D1-Cre: p = 1, 5 cells; A2A-Cre: p = 0.35, 8 cells). By contrast, the same illumination onto NAc-S D1-(Figure 2 A) or D2-SPNs (Figure 2 B) transfected with eNpHR3.0 suppressed action potential firing (D1-Cre: F(1,6) = 16.82; η2 = 0.74, p < 0.001, 7 cells; A2A-Cre: F(1,5) = 135.00; η2 = 0.96, p < 0.001, 6 cells). Importantly, there was no dikerence in firing prior to or following LED light illumination, indicating that the optical manipulation did not alter D1-or D2-SPNs physiology. Together, these results validate the ekicacy of eNpHR3.0 to specifically silence D1-and D2-SPNs activity at the time of LED light delivery.

Ex-vivo cell recordings in D1-Cre and A2a-Cre rats.

A. D1-Cre were bilaterally infused in the NAc-S with DIO-eYFP (black; 2 females) or DIO-eNpHR3.0 (blue; eNpHR3.0; 2 females). The representative raw traces of cell attached recordings are those of transfected neurons that were depolarized to elicit action potentials by injecting a brief positive current step (+150pA, 200ms duration, 0.5Hz). 625nm LED illumination (orange bar, continuous wave, 2mW) had no ekect in eYFP transfected neurons (black; 5 cells) but it inhibited action potential in eNpHR3.0 transfected neurons (blue; 7 cells). The grouped data for recordings include overlapping data points. B. A2a-Cre were bilaterally infused in the NAc-S with DIO-eYFP (black; 2 females) or DIO-eNpHR3.0 (red; 2 females). The representative raw traces of cell attached recordings are those of transfected neurons that were depolarized to elicit action potentials by injecting a brief positive current step (+150pA, 200ms duration, 0.5Hz). 625nm LED illumination (orange bar, continuous wave, 2mW) had no ekect in eYFP transfected neurons (black; 8 cells) but it inhibited action potential in eNpHR3.0 transfected neurons (red; 6 cells). The grouped data for recordings include overlapping data points.

NAc-S D1-SPNs mediate outcome-specific PIT

This experiment examined the ekects of silencing NAc-S D1-SPNs on outcome-specific PIT. D1-Cre rats were bilaterally infused in the NAc-S with either the null Cre-dependent eYFP virus (eYFP: 5 females and 5 males) or the Cre-dependent halorhodopsin (eNpHR3.0: 5 females and 5 males) virus and were implanted with fiber-optic cannulas above the NAc-S (Figure 3 A and Supplementary Figures S1 A-B). Using the outcome-specific PIT protocol depicted in Figure 3 B, the rats first received Pavlovian conditioning during which two stimuli (S1 and S2; white noise or clicker) were paired with two food outcomes (O1 and O2; grain pellets or sucrose solution). Next, the rats underwent instrumental conditioning and learned that pressing one lever earned one of the food outcomes whereas pressing a second lever delivered the other outcome. Finally, an outcome-specific PIT test was administered and assessed choice between the two lever press actions in the presence or absence of each stimulus. Critically, half of the trials for each stimulus took place under LED light stimulation (625 nm, continuous; ON trials) whereas the other half of the trials was conducted without the stimulation (OFF trials). Thus, NAc-S D1-SPNs were silenced on half of the trials in rats infused with eNpHR3.0.

NAc-S D1-SPNs mediate outcome-specific PIT.

A. D1-Cre rats were bilaterally infused in the NAc-S with DIO-eYFP (black; 5 females and 5 males) or DIO-eNpHR3.0 (blue; 5 females and 5 males). Fiber-optic cannulas were implanted above the NAc-S to provide 625nm LED illumination (continuous). B. Schematic representation of the behavioral design; S1 and S2: noise and clicker stimuli (counterbalanced); O1 and O2: grain pellets and sucrose solution (counterbalanced); A1 and A2: left and right lever press (counterbalanced). At test, S1 and S2 were presented 4 times each, in a pseudorandom order. Half of the trials for each stimulus was conducted under 625nm LED illumination (ON; continuous wave; ∼10mW) whereas the LED remained inactivated during the other half of the trials (OFF). ON/OFF trials were counterbalanced. C. Outcome-specific PIT test: net lever presses when the stimuli predicted the same outcome as the action (Same) or when the stimuli predicted the dikerent outcome (Dikerent). Lever presses are shown for each group in trials conducted under 625nm LED illumination (ON) and in trials without illumination (OFF). Data are shown as mean ± SEM. Panel C includes individual data points for female (filled circle) and male (open circle) rats. Asterisks denote significant ekect (* p < 0.05; ** p < 0.01; *** p < 0.001; n.s., nonsignificant).

Pavlovian and instrumental conditioning proceeded as expected (Supplementary Figures S1 C-D). The data of most interest are those from the outcome-specific PIT test and are presented in Figure 3 C (see also Supplementary Figure S1 E-F). They are plotted as the mean number of lever presses per minute when the stimulus predicted the same outcome as the action (Same) and when the stimulus predicted a dikerent outcome to the action (Dikerent). Thus, A1 was identified as “Same” and A2 as “Dikerent” in the presence of S1. Conversely, A2 was identified as “Same” and A1 as “Dikerent” in the presence of S2. Further, baseline responding (number of lever presses per minute on the two actions during the two minutes preceding each stimulus presentation) was subtracted from the rates of responding during the stimuli since it did not diker between groups (Group – eYFP vs. eNpHR3.0: p = 0.42). This approach allowed us to focus on the net ekect of the stimuli on choice performance.

Silencing NAc-S D1 SPNs eliminated outcome-specific PIT. Overall lever press rates were similar between groups (Group: p = 0.48), LED light condition (LED – ON vs OFF: p = 0.14), and the two factors did not interact (Group x Light: p = 0.07). However, lever press rates were higher on the action earning the same outcome as the stimuli compared to the action earning the dikerent outcome (Lever – Same vs. Dikerent: F(1,18) = 50.06; η2 = 0.74; p < 0.001), regardless of group (Group x Lever: p = 0.77). There was no Lever by LED light condition interaction (Lever x LED: p = 0.72) but critically, there was an interaction between Group, LED light condition, and Lever during the presentation of the predictive stimuli (Group x LED x Lever: F(1,18) = 7.19; η2 = 0.29; p = 0.02). Follow-up analyses revealed that rats in the eYFP group displayed intact outcome-specific PIT whether the LED light was OFF (F(1,9) = 6.14; η2 = 0.67; p < 0.05) or ON (F(1,9) = 39.46; η2 = 0.81; p < 0.001). By contrast, rats in the eNpHR3.0 group displayed outcome-specific PIT when the light was OFF (F(1,9) = 12.38; η2 = 0.58; p < 0.01) but not when it was ON (p = 0.67). Together, these results indicate that activity in NAc-S D1-SPNs is necessary for the outcome-specific influence exerted by predictive stimuli on choice between actions. This is consistent with previous pharmacological work showing that NAc-S D1Rs blockade removes specific PIT (Laurent et al., 2014). Importantly, the impairment produced by silencing NAc-S D1-SPNs was restricted to the influence of predictive stimuli on choice between actions. In the same rats, we found that this silencing had no ekect on valued-based choice (Supplemental Figure S1 G) using an outcome devaluation procedure: all rats performed the action that previously earned a valuable outcome more than the action that earned a devalued outcome. This agrees with studies reporting that NAc-S lesion or inactivation spares value-based choice (Corbit & Balleine, 2011; Corbit et al., 2001; Morse et al., 2020; Laurent et al., 2012; Laurent et al., 2014).

NAc-S D2-SPNs mediate outcome-specific PIT

The removal of outcome-specific PIT under silencing of NAc-S D1-SPNs reproduced the impairment previously observed with blockade of NAc-S D1Rs (Laurent et al., 2014). Interestingly, the latter pharmacological study did not find any ekect following NAc-S infusion of a D2Rs antagonist. To evaluate this finding, the present experiment examined the ekects of silencing NAc-S D2-SPNs on outcome-specific PIT. A2a-Cre rats were bilaterally infused in the NAc-S with either the null Cre-dependent eYFP virus (eYFP: 4 females and 4 males) or the Cre-dependent halorhodospin (eNpHR3.0: 4 females and 2 males) virus and were implanted with fiber-optic cannulas above the NAc-S (Figure 4 A and Supplementary Figures S2 A-B). The rats then received the outcome-specific PIT protocol previously described (Figure 3 B).

NAc-S D2-SPNs mediate for outcome-specific PIT.

A. A2a-Cre rats were bilaterally infused in the NAc-S with DIO-eYFP (black; 4 females and 4 males) or DIO-eNpHR3.0 (red; 4 females and 2 males). Fiber-optic cannulas were implanted above the NAc-S to provide 625nm LED illumination (continuous). B. Outcome-specific PIT test: net lever presses when the stimuli predicted the same outcome as the action (Same) or when the stimuli predicted the dikerent outcome (Dikerent). Lever presses are shown for each group in trials conducted under 625nm LED illumination (ON) and in trials without illumination (OFF). Data are shown as mean ± SEM. Panel B includes individual data points for female (filled circle) and male (open circle) rats. Asterisks denote significant ekect (* p < 0.05; ** p < 0.01; n.s., nonsignificant).

Pavlovian and instrumental conditioning went as expected (Supplementary Figures S2 C-D). The data from the outcome-specific PIT test are presented in Figure 4 B, again with baseline subtracted since it did not diker between groups (Group: p = 0.89; see also Supplemental Figure 2 E-F). Silencing NAc-S D2-SPNs activity eliminated outcome-specific PIT. Overall lever press rates during the test were similar between groups (Group: p = 0.86), LED light condition (LED: p = 0.21), and the two factors did not interact (Group x LED: p = 0.70). Lever press rates were higher on the action earning the same outcome as the stimuli compared to the action earning the dikerent outcome (Lever: F(1,12)= 22.77; η2 = 0.66; p < 0.001), regardless of group (Group x Lever: p = 0.46). There was no Lever by LED light condition interaction (Lever x LED: p = 0.15) but critically, there was an interaction between group, LED light condition, and Lever during the presentation of the predictive stimuli (Group x LED x Lever: F(1,12) = 8.73; η2 = 0.42; p < 0.01). Follow-up analyses revealed that control eYFP rats expressed outcome-specific PIT whether the LED light was OFF (F(1,7) = 16.25; η2 = 0.70; p < 0.01) or ON (F(1,7) = 20.58; η2 = 0.75; p < 0.01).

By contrast, rats in the eNpHR3.0 group displayed outcome-specific PIT when the light was OFF (F(1,5) = 6.94; η2 = 0.58; p < 0.05) but not ON (p = 0.34). Thus, these results show for the first time that activity in NAc-S D2-SPNs is necessary for the outcome-specific influence exerted by predictive stimuli on choice between actions.

The present finding contrasts with our previous observation that outcome-specific PIT remains unakected by NAc-S infusion of a D2Rs antagonist (Laurent et al., 2014). However, the capacity of such pharmacological manipulation to reveal the function of D2-SPNs is unclear. For instance, a D2Rs antagonist could be expected to enhance D2-SPN activity, since D2Rs are Gi coupled and so their activation should reduce adenylyl cyclase activity. Further, D2Rs are not exclusively expressed on this cell population (Gerfen & Surmeier, 2011), as they can be found on local CINs and presynaptic dopamine terminals. Finally, previous work found that pharmacological treatment targeting D2Rs can leave D2-SPNs activity unakected (Tozzi et al., 2007). These issues are overcome by the present optical manipulation in A2a-Cre rats, and we are therefore confident that NAc-S D2-SPNs activity is in fact critical for outcome-specific PIT. Importantly, we also found that the impairment produced by silencing NAc-S D2-SPNs was restricted to the influence of predictive stimuli on choice between actions. In the same rats, this silencing had no ekect on valued-based choice (Figure S2 G).

NAc-S D1-SPNs projections to the VP mediate outcome-specific PIT

Previous work indicates that communication between the NAc-S and VP is critical for outcome-specific PIT (Leung & Balleine, 2013). Consistent with the literature (Lu et al., 1997; Kupchik et al., 2015), we observed that both NAc-S D1-and D2-SPNs densely innervate the VP (Figure 1 E-F). Since we found that activity in both neuron populations mediates the outcome-specific PIT ekect, we investigated whether projections from each population contribute to the ekect. We first focused on those originating from NAc-S D1-SPNs. D1-Cre rats were bilaterally infused in the NAc-S with either the null Cre-dependent eYFP virus (eYFP: 3 females and 5 males) or the Cre-dependent halorhodopsin (eNpHR3.0: 6 females and 6 males) virus and were implanted with fiber-optic cannulas above the VP (Figure 5 A and Supplementary Figures S3 A-B). The rats then received the behavioral protocol previously described (Figure 3 B).

NAc-S D1-SPNs projections to the VP mediate outcome-specific PIT.

A. D1-Cre rats were bilaterally infused in the NAc-S with DIO-eYFP (black; 3 females and 5 males) or DIO-eNpHR3.0 (blue; 6 females and 6 males). Fiber-optic cannulas were implanted above the VP to provide 625nm LED illumination (continuous). B. Outcome-specific PIT test: net lever presses when the stimuli predicted the same outcome as the action (Same) or when the stimuli predicted the dikerent outcome (Dikerent). Lever presses are shown for each group in trials conducted under 625nm LED illumination (ON) and in trials without illumination (OFF). Data are shown as mean ± SEM. Panel B includes individual data points for female (filled circle) and male (open circle) rats. Asterisks denote significant ekect (** p < 0.01; *** p < 0.01; n.s., nonsignificant).

Pavlovian and instrumental conditioning went as expected (Supplementary Figures S3 C-D). The data from the outcome-specific PIT test are presented in Figure 5 B in the manner described previously since baseline responding did not diker between groups (Group: p = 0.34; see also Supplemental Figures 3 E-F). Silencing NAc-S D1-SPNs projections to the VP eliminated outcome-specific PIT. Overall lever press rates during the test were similar between groups (Group: p = 0.11). LED activation reduced these rates (LED: F(1,18)= 7.58; η2 = 0.30; p < 0.05) but this reduction depended on the group considered (Group x LED: F(1,18)= 9.78; η2 = 0.35; p < 0.01). Lever press rates were higher on the action earning the same outcome as the stimuli compared to the action earning the dikerent outcome (Lever: F(1,18)= 35.73; η2 = 0.67; p < 0.001), regardless of group (Group x Lever: p = 0.19). There was a Lever by LED light condition interaction (Lever x LED: F(1,18)= 7.56; η2 = 0.29; p < 0.05) and critically, there was an interaction between group, LED light condition, and Lever during the presentation of the predictive stimuli (Group x LED x Lever: F(1,18) = 5.01; η2 = 0.22; p < 0.05). Follow-up analyses revealed that control eYFP rats expressed outcome-specific PIT whether the LED light was OFF (F(1,7) = 26.63; η2 = 0.79; p < 0.001) or ON (F(1,7) = 12.31; η2 = 0.64; p < 0.01). By contrast, rats in the eNpHR3.0 group displayed outcome-specific PIT when the light was OFF (F(1,11) = 26.53; η2 = 0.71; p < 0.001) but not ON (p = 0.48). Thus, these results show for the first time that NAc-S D1-SPNs mediate outcome-specific PIT via their projections to the VP. Importantly, we also found that the impairment produced by silencing NAc-S D1-SPNs terminals in the VP was restricted to the influence of predictive stimuli on choice between actions. In the same rats, this silencing had no ekect on valued-based choice (Figure S3 G).

NAc-S D2-SPNs projections to the VP do not mediate outcome-specific PIT

Since activity in NAc-S D2-SPNs is required for outcome-specific PIT (Figure 3) and these neurons innervate the VP (Figure 1 F), we assessed whether D2-SPNs projections to the VP are involved in outcome-specific PIT. A2a-Cre rats were bilaterally infused in the NAc-S with either the null Cre-dependent eYFP virus (eYFP: 3 females and 5 males) or the Cre-dependent halorhodopsin (eNpHR3.0: 2 females and 5 males) virus and were implanted with fiber-optic cannulas above the VP (Figure 6 A and Supplementary Figures S4 A-B). The rats then received the behavioral protocol previously described (Figure 3 B).

NAc-S D2-SPNs projections to the VP do not mediate outcome-specific PIT.

A. A2a-Cre rats were bilaterally infused in the NAc-S with DIO-eYFP (black; 3 females and 5 males) or DIO-eNpHR3.0 (red; 2 females and 5 males). Fiber-optic cannulas were implanted above the VP to provide 625nm LED illumination (continuous). B. Outcome-specific PIT test: net lever presses when the stimuli predicted the same outcome as the action (Same) or when the stimuli predicted the dikerent outcome (Dikerent). Lever presses are shown for each group in trials conducted under 625nm LED illumination (ON) and in trials without illumination (OFF). Data are shown as mean ± SEM. Panel B includes individual data points for female (filled circle) and male (open circle) rats.

Pavlovian and instrumental conditioning went as expected (Supplementary Figures S4 C-D). The data from the outcome-specific PIT test are presented in Figure 6 B as previously since baseline responding did not diker between groups (Group: p = 0.90; see also Supplemental Figures 4 E-F). Silencing NAc-S D2-SPNs projections to the VP had no ekect on outcome-specific PIT. Overall lever press rates were similar between groups (Group: p = 0.56), LED light condition (LED: p = 0.75), and the two factors did not interact (Group x LED: p = 0.73). The rates were higher on the action earning the same outcome as the stimuli relative to the action earning the dikerent outcome (Lever: F(1,13) = 22.71; η2 = 0.64; p < 0.001), irrespective of group (Group x Lever: p = 0.55). There was no interaction between groups, LED light condition, and Lever (Group x Light x Lever: p = 0.59). Thus, NAc-S D2-SPNs do not appear to mediate outcome-specific PIT via their projections to the VP. Likewise, we found no evidence that these projections influence valued-based choice (Figure S4 G).

Discussion

The present experiments investigated the role of NAc-S D1-and D2-SPNs in choice between actions using an outcome-specific PIT task. First, they combined anatomical tract-tracing and ex-vivo electrophysiology to demonstrate that two recently developed knock-in rat lines (Pettibone et al., 2019) enable selective silencing of activity in either population of NAc-S SPNs. Consistent with previous findings (Laurent et al., 2014), they found that NAc-S D1-SPNs are necessary for PIT since their silencing eliminated outcome-specific choice. Additionally, they showed that this choice is also abolished by silencing NAc-S D2-SPNs, providing the first evidence that both SPNs populations contribute to PIT expression. Finally, the last two experiments revealed for the first time that outcome-specific choice is likely to involve downstream regulation of VP function by NAc-S D1-SPNs but not NAc-S D2-SPNs. Together, these findings oker novel insights into the cellular mechanisms that govern the outcome-specific influence of predictive stimuli on choice between actions.

Convincing evidence had been provided for a significant role of NAc-S D1-SPNs in PIT (Laurent et al., 2014). The evidence was based on the observations that PIT is associated with an increase in ERK1/2 phosphorylation within these SPNs and that pharmacological blockade of NAc-S D1Rs eliminates outcome-specific choice between actions. These findings align with the present study, showing that PIT is eliminated when NAc-S D1-SPNs are silenced during presentations of the predictive stimuli at the time of choice. Importantly, the ekect of NAc-S D1-SPNs silencing was specific to the assessment of choice between actions in the presence of predictive stimuli. For instance, the ability of stimuli to elicit approach behaviors toward the magazine remained intact despite NAc-S D1-SPNs silencing, indicating that the silencing ekect was not mediated by modulating potential competition between Pavlovian and instrumental responses (Lovibond, 1981; Holmes et al., 2010). Furthermore, NAc-S D1-SPNs silencing preserved the capacity to select between actions based on the value of their respective outcomes. This selectivity in the impairment produced by NAc-S D1-SPNs is consistent with previous findings demonstrating that the NAc-S does not contribute to learning and retrieving the S-O and A-O associations produced by Pavlovian and instrumental conditioning (Corbit & Balleine, 2011; Corbit et al., 2001; Morse et al., 2020; Laurent et al., 2012; Laurent et al., 2014).

Previous research failed to provide any evidence supporting the involvement of NAc-S D2-SPNs in outcome-specific choice (Laurent et al., 2014). Specifically, PIT was found to produce no change in ERK1/2 phosphorylation in these neurons and was left intact by NAc-S D2Rs blockade. However, these assessments have significant limitations, including the widespread distribution of striatal D2Rs (Gerfen & Surmeier, 2011) and the inability of ERK1/2 phosphorylation or D2Rs antagonism to capture or distort activity in NAc-S D2-SPNs (Bertran-Gonzalez et al., 2008; Tozzi et al., 2007). Further, pharmacological blockade of D2Rs would be expected to enhance D2-SPNs activity, since D2Rs are Gi coupled and so their activation reduces adenylyl cyclase activity. The present study addressed these limitations by implementing optogenetic silencing in a knock-in rat line giving access to D2-SPNs through the targeting of the A2a receptor that is predominantly expressed in these neurons in the NAc-S (Schikmann et al., 2007). We found that PIT is eliminated when NAc-S D2-SPNs are silenced during presentations of the predictive stimuli at the time of choice. Moreover, the same silencing left intact the ability of the predictive stimuli to elicit magazine approaches and preserved the capacity to choose between actions based on the value of their outcomes. Thus, we conclude that both NAc-S D1-and D2-SPNs are indispensable to outcome-specific PIT. Nevertheless, it will be important for future studies to confirm NAc-S D2-SPNs involvement in PIT using alternative approaches, such as chemogenetic or immunohistochemical assessments employing a marker capable of capturing the function of this neuronal population (Matamales et al., 2020).

Consistent with the literature (Lu et al., 1997; Kupchik et al., 2015), we found that both NAc-S D1-and D2-SPNs send dense projections to the VP, which is often described as the major ekerent of the nucleus accumbens (Heimer et al., 1991; Zahm & Heimer, 1990). Importantly, the VP has been shown to play a crucial role in outcome-specific choice (Leung & Balleine, 2013; Leung & Balleine, 2015; Leung et al., 2024b). VP neurons exhibit enhanced cFos activation following an outcome-specific PIT test, and their levels of activation correlate with PIT performance. Moreover, pharmacological inactivation of the VP eliminates outcome-specific choice, and the same elimination is observed when the NAc-S is disconnected from the VP. We therefore tested for a role of NAc-S D1-and D2-SPNs projections to the VP during choice between actions in our outcome-specific PIT task. We found that PIT was abolished by silencing NAc-S D1-SPNs terminals but not by silencing NAc-S D2-SPNs terminals. Neither silencing akected magazine approaches elicited by the predictive stimuli or choice between actions based on the value of their outcomes. It is important to note that our study does not provide any evidence about the ekicacy of NAc-S D2-SPNs terminals silencing in the VP, and future experiments should aim to provide such evidence or adopt other methods to study this pathway. This could involve using opsins with enhanced axonal silencing ekicacy (Mahn et al., 2021; Copits et al., 2021), or employing alternative methods known to disrupt neurotransmitter release such as chemogenetics (Rost et al., 2022). Yet, the same silencing for NAc-S D1-SPNs terminals resulted in PIT elimination. Therefore, it seems reasonable to conclude that NAc-S D1-SPNs, but not NAc-S D2 SPNs, projections to the VP are indispensable to observe outcome-specific choice in a PIT task. Since NAc-S D2-SPNs appear to exclusively project to the VP (Humphries & Prescott, 2010; Kupchik et al., 2015), our findings suggest that these SPNs mediate outcome-specific PIT by locally regulating NAc-S function. By contrast, it remains to be determined whether NAc-S D1-SPNs also coordinate outcome-specific PIT via their projections to the lateral hypothalamus (O’Connor et al., 2015) and/or the ventral tegmental area (VTA) (Humphries & Prescott, 2010), in addition to the VP.

The present findings are consistent with a recent model proposing that outcome-specific choice in PIT relies on an opioid-based memory residing in the NAc-S (for full description and illustration of the model see Leung et al., 2024a; Laurent & Balleine, 2021; Morse et al., 2020). In this model, as the basolateral amygdala encodes and stores outcome-specific S-O associations across Pavlovian conditioning, it also drives the formation of the NAc-S memory, which involves the durable accumulation of delta-opioid receptors (DOPRs) on the somatic membrane of local cholinergic interneurons (CINs) (Bertran-Gonzalez et al., 2013; Morse et al., 2020). Although this memory is not necessary for Pavlovian conditioning per se, its expression is later required to enable the predictive stimuli to guide outcome-specific choice during the PIT test (Morse et al., 2020). The model specifically proposes that memory expression is controlled by fluctuations in glutamatergic release from cortical inputs and dopamine release from projections originating from the VTA, as this brain region has been found to be important for outcome-specific PIT (Corbit et al., 2007; Sias et al., 2024; Seitz et al., 2022; Leung & Balleine, 2015). One consequence is to activate NAc-S D2-SPNs and enkephalin discharge by these neurons. Enkephalin then binds onto DOPRs that had accumulated on local CINs and dampens acetylcholine secretion from the interneurons. The sudden drop in NAc-S acetylcholine frees D1-SPNs from the inhibitory tone imposed by acetylcholine occupancy of muscarinic M4 receptors that are exclusively found on these SPNs (Tayebati et al., 2004; Lobo et al., 2006; Guo et al., 2010; Jeon et al., 2010). The ultimate consequence is to promote NAc-S D1-SPNs function, including the coordination of outcome-specific choice in PIT by regulating activity in downstream brain regions such as the VP. The model predicts all the findings presented here. NAc-S D2-SPNs silencing eliminates PIT at the time of choice by preventing enkephalin release and thereby, expression of the DOPR-based memory. The main consequence of this expression is precluded by NAc-S D1-SPNs silencing at the time of choice, while silencing the terminals of these SPNs in the VP squash their ability to regulate the activity of this brain region to coordinate outcome-specific choice in PIT. Thus, one fundamental implication of the present findings is to strengthen the proposal that an opioid-based memory system in the NAc-S enables outcome-specific choice between actions.

In summary, the present experiments found that the two main populations of SPNs in the NAc-S are indispensable for outcome-specific choice between actions in a PIT task. They also revealed that NAc-S D1-SPNs coordinate this choice by downstream regulation of VP activity. By contrast, NAc-S D2-SPNs function in PIT appears to be restricted to modulating local activity. These findings provide novel insights in the cellular mechanisms and circuitry underlying the outcome-specific influence of predictive stimuli on choice between actions, and are consistent with a recent model proposing that this influence is mediated by an opioid-based memory system in the NAc-S. Beyond these mechanistic considerations, the findings oker an opportunity to gain novel insights about various disorders in which the outcome-specific influence of predictive stimuli over our choices and decisions is dysfunctional. These disorders include depression (Geurts et al., 2013; Nord et al., 2018), anxiety disorders (Quail et al., 2017; Krypotos & Engelhard, 2020), substance-use disorders (Heinz et al., 2019; Hogarth et al., 2018; Hogarth et al., 2018; Steins-Loeber et al., 2020; Garbusow et al., 2016; Garbusow et al., 2019; Hogarth et al., 2018), gambling disorders (Genauck et al., 2019), anorexia nervosa (Vogel et al., 2019) and obesity (Watson et al., 2014; Lehner et al., 2017; Meemken & Horstmann, 2019).

Materials and methods

Subjects

135 rats from two genetically modified knock-in lines were used and obtained from the breeding facility at the University of New South Wales (Sydney, Australia). D1-Cre rats (Rat Research & Resource Center, Columbia, MO, USA; LE-Drd1em1(iCre)Berke, RRRC#: 00856) expressed Cre recombinase in neurons expressing the dopamine D1 receptors (D1R). A2a-Cre rats (Rat Research & Resource Center, Columbia, MO, USA; LE-Adora2aem1(iCre)Berke, RRRC#: 00857) expressed Cre recombinase in neurons expressing the adenosine A2A receptors (A2R). All rats were heterozygous and generated by crossing a heterozygous male with a Long-Evans wild-type female obtained from a colony maintained by the University of New South Wales (founding animals were sourced from Envigo/Inotiv, Blue Spruce outbred; HsdBlu:LE). Genotyping was completed by sending ear clippings to Transnetyx (Cordova, TN, USA). The rats were at least 8 weeks old at the beginning of each experiment. Ekorts were made to allocate an equal number of female and male rats in each group. However, these ekorts were hampered by exclusions following post-mortem assessments of viral spread and fiber-optic cannulas placement. The final number of female and male rats did not provide sukicient power to analyze an ekect of sex (our previous work found no influence of sex on PIT or value-based choice; Burton et al., 2024). Therefore, all test data present individual performance for female and male rats. Rats were housed in transparent plastic boxes with their littermates (up to four rats per box) throughout, and in a climate-controlled colony room maintained on a 12 h light-dark cycle (lights on at 7:00am). Behavioral procedures were conducted during the light phase (8:00am to 6:00pm). Water and standard lab chow were available ad libitum prior to the start of each experiment. Rats were food restricted to maintain them at ∼90% of their ad libitum body weight during the behavioral protocols. Food restriction was initiated 3-5 days prior to the start of the protocols and was maintained throughout all training and testing phases. Rats received a daily food amount, which was adjusted based on body weight measurements recorded every two days. The Animal Care and Ethics Committee at the University of New South Wales approved all procedures.

Experimental timeline

Rats underwent initial stereotaxic surgery for viral infusion, followed by 7-11 days recovery. After recovery, rats received 8 days of Pavlovian conditioning and 8 days of instrumental conditioning, except in the VP experiments (NAc-S D1-and D2-SPNs→VP projection experiment) where only 7 days of instrumental conditioning were given. Following a 3-4 day interval, rats received a second surgery to implant fiber-optic cannulas and 3-4 days additional recovery. Rats then completed reminder sessions: one Pavlovian and two instrumental conditioning sessions, except in the last experiment (NAc-S D2-SPNs→VP projection experiment) where three instrumental sessions were given. Testing began the following day with the PIT test, followed by two instrumental conditioning reminder sessions over the next two days. Finally, rats underwent outcome devaluation using sensory-specific satiety and choice tests over two consecutive days.

Viruses

The following viruses were used: AAV5-EF1a-DIO-eYFP (eYFP; AddGene; Watertown, MA, USA) and AAV5-EF1a-DIO-eNpHR3.0-eYFP (eNpHR3.0; UNC Vector Core; Chapel Hill, NC, USA).

Surgeries

Stereotaxic surgery was conducted under isoflurane gas anesthesia (0.8L/min; induction at 4-5%, maintenance 2-2.5%). Animals were placed in a stereotaxic frame (Kopf Instruments, Tujunga, CA, USA), the surgical area was shaved, and betadine iodine antiseptic solution was applied. At the incision site, bupivacaine hydrochloride (0.5%; 0.1-0.2mL) was injected subcutaneously. In the lower flank, Metacam (1mg/kg) was injected subcutaneously. An incision was made along the midline to expose the skull, the membrane on top of the skull was cleared, and the skull was adjusted to align bregma and lambda on the same horizontal plane.

For viral infusion, small holes were drilled in the skull above the DS or NAc-S. The following coordinates (indicated in mm from bregma) were used for the DS:-0.45 anteroposterior (A/P); ±2.6 (M/L);-4.85 dorsoventral (D/V). For the NAc-S, the coordinates were: +1.8 or +1.9 A/P; ±0.85 or ±0.95 mediolateral M/L;-7.7 or-7.85 D/V. An Infuse/Withdraw Pump (Standard Infuse/Withdraw Pump 11 Elite; Holliston, Harvard Apparatus, MA, USA) in combination with glass 1µL syringes (86200; Hamilton Company, Reno, NV, USA) were used to perform infusions. 0.5µL of the virus was infused in each hemisphere at a rate of 0.1µL/min. Following each infusion, the needle was left in position for 5-10 min to allow for dikusion. The incision site was closed and secured with surgical sutures (684G; Ethicon, North Ryde, NSW, Australia) or surgical staples (59027; Stoelting, Wood Dale, IL, USA), and betadine antiseptic ointment was applied to the area. Animals then received subcutaneous injections of the antibiotic Duplocillin (0.15ml/kg) and sodium chloride (0.9%, 2mL) and placed on a heat mat for recovery. Once alert and responsive, animals were returned to their home cages and monitored daily as they recovered for a minimum of one week following surgery with ad libitum access to food and water.

Rats received a second surgery to bilaterally implant fiber-optic cannulas (10mm length, 0.48mm diameter; Doric Lenses, Québec, QC, Canada) after completion of the initial 15-16 days of training (Pavlovian and Instrumental). The surgical area was prepared using the same method as described for viral surgeries. Three jewellers’ screws were inserted into the skull, distributed anterior and posterior to the implantation site and secured in place with dental cement. Fiber-optic cannulas were implanted into the targeted brain region. The following coordinates (indicated in mm from bregma) were used for the NAc-S: +1.8 or +1.9 A/P;-0.85 or-0.95 M/L for left hemisphere; +2.2 M/L with-10° angle for right hemisphere;-6.75 D/V for left hemisphere;-6.8 or-6.9 D/V for right hemisphere. The following coordinates (indicated in mm from bregma) were used for the VP: +0.24 A/P; ±1.8 M/L;-7.5 or-8.0 D/V. Post-implantation, rats received the same care as described for viral surgeries and were allowed to recover for 3-4 days before resuming the behavioral procedures.

Behavioral apparatus

Training and testing were conducted in 12 identical operant chambers (ENV-007-VP; L 29.53 x W 23.5 x H 27.31 cm; MED Associates, Fairfax, VT, USA) enclosed within light-and sound-resistant cabinets (ENV-018MD; L 59.69 x W 40.64 x H 55.88 cm; MED Associates, Fairfax, VT, USA). The chambers consisted of two stainless-steel side walls, three clear polycarbonate walls and ceiling, and a stainless-steel grid floor. Each chamber was equipped with a recessed magazine connected to a 45mg grain pellet (F0165; Bio-Serv Biotechnologies, Flemington, NJ, USA; one pellet per delivery) dispenser and a pump fitted with a 60mL syringe that delivered a 0.2mL sucrose solution (20% w/v; delivered across 2 seconds; white sugar; Winc, Erskine Park, NSW, Australia) into the left side of the magazine. Head entries into the magazine were detected via an infrared beam that crossed the magazine opening. One retractable lever was located on either side of the magazine. A house light (3W, 24V) situated at the top of the wall opposite to the food magazine was used to illuminate the chamber during the training and test sessions. Each chamber contained a Sonalert white noise generator (80dB), and a 28V DC mechanical relay that delivered a 2Hz clicker stimulus. Training and testing sessions were programmed and controlled by computers external to the testing rooms using MED-PC V software (MED Associates, Fairfax, VT, USA) which also recorded experimental data from each session. All cabinets were fitted with cameras (IPC-K35AP; Dahua, Artarmon, NSW, Australia) to view real-time chamber activity using D-ViewCam DCS-100 software (D-Link, North Ryde, NSW, Australia). The ceiling of these chambers contained a hole which enabled the connection of the patch cable between the LED and each freely moving rat.

Outcome devaluation was conducted in a separate room equipped with 16 individual, open-top plastic boxes with stainless steel wire mesh lids. During devaluation, the lights were dimmed, and each individual chamber was fitted with either a ceramic dish for grain pellet devaluation, or a plastic sipper bottle for sucrose solution devaluation.

Optogenetic equipment

Fiber-optic cannulas (10mm; Doric Lenses, Québec, QC, Canada) consisted of a zircona ferrule contained within a stainless-steel receptacle from which a silica/polymer flat-tip fiber extended from. Fiber-optic patch cords (custom; Doric Lenses, Québec, QC, Canada) were enclosed in a flexible cladding that consisted of one transparent fiber that split into two smaller fibers for attachment to the implanted fiber-optic cannulas. Delivery of the LED light was controlled through a TTL adapter which converted 28V DC output to a TTL transition (MED Associates, Fairfax, VT, USA). The adaptor was connected to an LED driver (Doric Lenses, Québec, QC, Canada) and Connectorised LED light source (Doric Lenses, Québec, QC, Canada) with a fiber-optic rotary joint for attachment to the fiber-optic patch cords. LED light was delivered as orange light (625nm; continuous) and measured at least 10mW at the cannula tip using a photometer (PM200; ThorLabs, Newton, NJ, USA) when the LED was switched on.

Behavioral procedures

Pavlovian conditioning

Pavlovian conditioning involved 8 daily 1hr training sessions. Two auditory conditioned stimuli (S1 and S2; white noise or clicker) were paired with two food outcomes (O1 and O2; grain pellets or sucrose solution). Each stimulus was presented for 2min in duration, 4 times each in a pseudorandom order within each training session. Stimulus presentations were separated by pseudorandom intertrial intervals (ITIs) ranging between 3 to 5min in length (4min average). The stimulus order and ITIs were changed daily. Stimulus-outcome (S-O) pairings were counterbalanced within-and between-groups. Food outcomes were delivered on a random time 30s schedule throughout the stimulus presentation. Magazine entries were recorded throughout the session, and conditioned responding was analyzed by comparing magazine entries in the 2min prior to stimulus presentation and during the 2min stimulus presentation period.

Instrumental conditioning

During instrumental conditioning, two actions (A1 and A2; left or right lever presses) were trained with the two dikerent outcomes (O1 and O2) in 2 separate daily sessions for 8 days. The order of the sessions was counterbalanced, as were the A-O relationships, which were further counterbalanced with the S-O relationships. Each session ended when 20 outcomes were earned or when 30min had elapsed. For the first 2 days, actions were continuously reinforced (i.e. one action earned one outcome delivery). Across the next three sessions, actions were rewarded on a random ratio 5 (RR5) schedule (i.e. each action delivered an outcome with a probability of 0.2). For the final 3 days, animals were rewarded on an RR10 schedule (i.e. each response delivered an outcome with a probability of 0.1). In the VP experiments (NAc-S D1-and D2-SPNs to VP projections), rats only received 2 days on RR10. Lever press actions were measured throughout the duration of the session, and performance during instrumental training was assessed using the mean number of lever presses per min.

Pavlovian Instrumental transfer test

Across the three days preceding this test, rats were given one session of Pavlovian retraining and two separate sessions of instrumental retraining. In the VP experiments (NAc-S D1-and D2-SPNs→VP projections), rats first received one day of Pavlovian conditioning followed by 3 days of instrumental conditioning (RR10). Immediately before the test, patch cables were attached to the fiber-optic cannulas and the LED. During the test, both levers were inserted into the chamber simultaneously and remained in the chamber for the duration of the test. The test was conducted in extinction (i.e. no outcomes were delivered). To reduce baseline (pre S) lever presses, responding on both levers was extinguished for 8min prior to any stimuli being presented. Each stimulus was presented for 2min, four times each during the test in the following order: noise-clicker-clicker-noise-clicker-noise-noise-clicker. Each stimulus presentation was separated by a fixed 3min interval. Critically, the LED was activated during half of the trials (ON) but left inactive during the other half (OFF). The order for ON and OFF trials was counterbalanced. Responses were recorded prior to and during the S presentation. These were then collated into ‘pre’ (lever press rates in the two min prior to S presentation), ‘same’ (when S shares the same outcome with the action) and ‘dikerent’ (when S has a di9erent outcome with the action) following data collection.

Choice tests following outcome devaluation

Across the two consecutive days preceding these tests, rats were given 2 daily instrumental conditioning sessions (RR10). Following the second session, rats were habituated to the devaluation chambers for 30min. During these habituation sessions, animals had access to one outcome (counterbalanced). On the first test day, rats were placed in the devaluation chambers for 1hr where they had ad libitum access to one of the previously trained outcomes (O1 or O2). Once removed from the devaluation boxes, patch cables were attached to the fiber-optic cannulas and the LED, and the first choice test was conducted. Both levers were made available for 2min each, four times during the session with an intertrial interval of 2min where the levers were retracted. The LED was activated during 2 of the 2min trials where the levers were available (ON) and was left inactive during the other 2 trials (OFF). The order of the ON and OFF trials was counterbalanced. A similar procedure was implemented the following day except that the other outcome was devalued. Lever presses and magazine entries were recorded while the levers were available.

Tissue preparation and histology

Transcardial fixation and brain sectioning

Rats were euthanised with an intraperitoneal injection of sodium pentobarbital (1mL) and transcardially perfused with 4% paraformaldehyde (PFA) in 0.1M sodium phosphate buker (PBS, pH 7.5; 400mL for rats). Brains were immediately removed and placed in individual specimen jars containing PFA and stored at 4°C overnight for post-fixation. Brains were sliced into 40μm coronal or sagittal sections in 0.1M Tris-bukered saline (TBS, 0.25M Tris, 0.5M NaCl, pH 7.5) using a vibratome (VT1200; Leica Microsystems, North Ryde, NSW, Australia) and stored at-20°C in cryoprotective solution containing 30% ethylene glycol, 30% glycerol and 0.1M PBS.

Immunofluorescence

Free-floating sections were selected in TBS to contain the region of interest and washed three times for 10min each in TBS solution. Sections were then incubated in 0.5% Triton-X, 10% Normal Horse Serum (NHS) in TBS blocking solution for two hours. Sections were washed again in TBS solution three times for 10 min each. Sections were then incubated between 12-48h (dependent on antibody) in a solution containing the primary antibodies in 0.2% Triton-X, 2% NHS in TBS at 4°C. The primary antibodies were the following: purified mouse anti DARPP-32 (1:300; BD Biosciences, Macquarie Park, Australia; #AB398980) and polyclonal rabbit anti-GFP (1:1000; Invitrogen, Mulgrave, Australia; #AB221569). After the specified incubation period, the sections were washed for three times in TBS for 10min each and then transferred into a solution containing the secondary antibodies in TBS solution at room temperature for 1h. The secondary antibodies were the following: donkey anti-mouse Cy3 (1:400; Jackson ImmunoResearch Laboratories, West Grove, USA; #AB2340813) and donkey anti-rabbit Alexa-488 (1:500; Jackson ImmunoResearch Laboratories, West Grove, USA; #AB2313584). Next, sections were washed three times in TBS for 10min and then mounted onto glass slides using Vectashield hardset antifade mounting medium (H-1400; Vector Laboratories, Newark, CA, USA). Sections that did not require immunofluorescence staining were washed three times for 10min each in TBS and mounted using Vectashield hardset antifade mounting medium.

Microscopy and imaging

Imaging of the mounted brain sections was performed on a spinning disk confocal microscope (Zyla 4.2 sCMOS; Andor Technology, Belfast, UK) using the Diskovery multi-modal imaging platform (Oxford Instruments, Belfast, UK). The Diskovery platform was mounted on an inverted microscope body with a motorized stage and Perfect Focus System (Eclipse Ti; Nikon, Melville, NY, USA). Image capture was controlled using the NIS-Elements Software (Nikon, Melville, NY, USA). Single high-magnification field-of-view images were generated by automatic stitching of multiple adjacent frames from within a defined area using the motorized stage (optical magnification: 20X; pixel depth: 16-bit). All images were processed using Fiji ImageJ open-source software (Rueden et al., 2017). Additional images were taken with a confocal microscope (BX61W1; Olympus, Shinjuku, Tokyo, Japan) using Fluoview software (FV1000; Olympus, Shinjuku, Tokyo, Japan) for further placement analysis. The images were used to assess neuronal pathways, as well as the spread of viral infusions, placement of fiber-optic cannulas and potential damages due to the surgical procedures. Based on the latter, the following numbers of rats were excluded from the final statistical analyses due to misplaced viral or fiber-optics misplacements: NAc-S D1-SPNs silencing (n = 9 females and 9 males), NAc-S D2-SPNs silencing (n = 4 females and 5 males), NAc-S D1-SPNs→VP silencing (n = 5 females and 10 males), NAc-S D2-SPNs→VP silencing (n = 4 females and 1 male).

Immunofluorescence quantifications

The specificity of viral transduction in D1-Cre and A2-Cre rats (Figures S1 C-F) was achieved through manual cell counting and involved calculating a percentage of the total number of fluorophore-labeled (eYFP) cells relative to the total number of DARPP-32-positive cells, and a percentage of the total number of DARPP-32-positive cells relative to the total number of fluorophore-labeled (eYFP) cells. Two slices per rat were selected around the following anteroposterior coordinates (in mm relative to bregma):-0.45 for DS and +1.8 for NAc-S. Before quantification, all images were randomly renumbered, and cell counts were performed in Fiji Image J in one region of interest (ROI) per image (ROI size: 0.25 mm2 for both DS and NAc-S). The ROI was defined in the eYFP channel around the viral injection site. The average number (+/- sd) of DARPP-32-positive cells per ROI was: 194 ± 59 for DS and 132± 24 for NAc-S.

Electrophysiological recordings

Brain slice preparation

Rats were euthanized under isoflurane gas anesthesia (4% in air). Brains were immediately removed and sliced using a vibratome (VT1200; Leica Microsystems, North Ryde, NSW, Australia) in ice-cold oxygenated sucrose buker solution (consisting of [in mM]: 241 sucrose, 28 NaHCO3, 11 glucose, 1.4 NaH2PO4, 3.3 KCl, 0.2 CaCl2, 7 MgCl2). Coronal brain slices (300μm thickness) containing the NAc were sampled and maintained at 33°C in a submerged chamber containing physiological saline (consisting of [in mM]: 124 NaCl, 2.5 KCl, 1.25 NaH2PO4, 1 MgCl2, 1 CaCl2, 10 glucose and 26 NaHCO3) and equilibrated with 95% O2 and 5% CO2.

Recordings

Following 1h equilibration, slices were transferred to a recording chamber. An upright microscope (BX50WI; Olympus, Shinjuku, Tokyo, Japan) was used to visualise neurons using dikerential interference contrast (DIC) Dodt tube optics and eYFP fluorescence, and superfused continuously (1.5mL/min) with oxygenated physiological saline at 33°C. Whole-cell patch-clamp recordings were performed using glass pipettes (tip size 2–5 MΩ) containing internal solution (consisting of [in mM]: 115 K gluconate, 20 NaCl, 1 MgCl2, 10 HEPES, 11 EGTA, 5 Mg-ATP, 0.33 Na-GTP, and 5 phosphocreatine di(tris) salt, pH 7.3, osmolarity 285-290 mOsm/L). Neurons sampled during recording were marked with Biocytin (0.1%) added to the internal solution. A Multiclamp 700B amplifier (Molecular Devices, San Jose, CA, USA) connected to a Macintosh computer and interface ITC-18 (InstruTECH, Longmont, CO, USA) was used for data acquisition. Liquid junction potentials of-10mV were not corrected. In current-clamp mode, membrane potentials were sampled at 5kHz (low pass filter 2kHz; Axograph X, Axograph, CA, USA). Stock solution of drug was diluted to working concentration in the extracellular solution immediately before use and applied by continuous superfusion. Data from whole-cell recordings were included in analyses if the neurons appeared healthy under DIC on the monitor screen.

LED stimulation

Transfected neurons with eNpHR3.0 or eYFP in the NAc-S were illuminated by a continuous wave of LED light (625nm, 2 minutes, 2mW; ThorLabs, Newton, NJ, USA) onto the brain slice under the 40X water-immersion objective. Since SPNs in ex-vivo brain slices are not spontaneously active, brief injection of positive current step pulses were applied to make them fire action potentials.

Post-hoc histological analysis

Immediately after physiological recording, brain slices were fixed overnight using 4% PFA in 0.16M phosphate buker (PB) solution, washed and then transferred into 0.5% Triton-X in PB solution for 3 days to permeabilise cells. Slices were washed, mounted onto glass slides, and then coverslipped with Fluoromount-G mounting medium (Southern Biotech, Birmingham, AL, USA). A 2D projection image was then obtained from a collated image stack using the confocal microscope (X61W1; Olympus, Shinjuku, Tokyo, Japan).

Statistical analyses

The data presented met the assumptions of the statistical test used. The main measures were the number of magazine entries and lever presses per minute. The dikerence between groups were analyzed using a planned orthogonal contrast procedure controlling the per contrast error rate (Hays, 1963). A first contrast (Group) compared rats that had been infused with the control eYFP virus and rats that had been infused with the inhibitory eNpHR3.0 virus. Changes across days (Day) during Pavlovian and instrumental conditioning were assessed by planned linear trend analyses. Pavlovian conditioning and the PIT tests included a within-subject factor (Period) comparing magazine entry rates during the 2min stimuli against rates during a 2min baseline period immediately preceding the stimuli. During the PIT tests, a within-subject factor (Lever) examined the net (i.e., minus baseline) rates of lever presses on the response (Same) earning the same outcome as predicted by the stimulus against those on the response (Dikerent) earning the dikerent outcome. A second within-subject factor (LED) examined these rats when the LED was activated (ON) or not activated (OFF). The choice tests following outcome devaluation used a within-subject factor of response identity (Lever) comparing rates of lever presses on the response (Valued) earning the valued outcome against those on the response (Devalued) earning the devalued outcome. It also used the LED within-subject factor described above. All analyses were carried out using the PSY statistical software (School of Psychology, the University of New South Wales, Australia). The Type I error rate was controlled at alpha = 0.05 for each contrast tested. If interactions were detected, follow-up simple ekects analyses were calculated to determine the source of the interactions. For each significant statistical comparison, we report measures of ekect size partial eta-squared (η2; η2 = 0.01 is a small ekect, η2 = 0.06 a medium ekect and η2 = 0.14 a large ekect).